### 2016-04-07 Paper/paper.tex
added folder for the paper

\documentclass[times,10pt,twocolumn]{article}
\usepackage{latex8}
\usepackage{times}
\usepackage{graphicx}

\usepackage{amsmath}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{colortbl}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage{url}
\usepackage{balance}

\newcommand{\ub}{user defined-block}
\newcommand{\nPrograms}{250.166}
\newcommand{\nemptyPrograms}{14.285}
\newcommand{\nScriptPrograms}{233.514}

\newcommand{\numkids}{61~}
\newcommand{\lm}{\emph{Long Method}~}
\newcommand{\dup}{\emph{Duplication}~}

\newcommand{\todo}[1]{\textbf{#1}}

\begin{document}

\title{Clones in Block-Based Languages: A Large Scale Analysis of Scratch Programs}

\author{Felienne Hermans\\
       {Delft University of Technology}\\
      {Mekelweg 4}\\
       {Delft, the Netherlands}\\
       {f.f.j.hermans@tudelft.nl}
}

\maketitle

\begin{abstract}
Recently, block-based programming languages like Alice, Scratch and Blockly have become popular tools for programming education. There is substantial research showing that block-based languages are suitable for early programming education. But do block-based programs suffer from code clones \emph{smelly} too? In a recent controlled experiment, we found that code clones in Scratch indeed hamper novice Scratch programmers. In this paper we explore how commonly code clones occur in Scratch. To that end we have scraped Scratch public repository and retrieved 250.000 programs. We have analyzed three types of clones within those programs': clones of entire scripts within sprites, clones of entire between sprites, and cloned conditions. We find that those clones occur in ... of programs.
\end{abstract}

\section{Introduction}

Scratch is a programming language developed to teach children programming by enabling them to create games and interactive animations. The public repository of Scratch programs contains over 12 million projects. Scratch is a \emph{block-based} language: users manipulate blocks to program. Block-based languages are visual languages, but also use some successful aspects of text-based languages such as limited text-entry and indentation, and as such are closer to `real', textual programming than other forms of visual programming, like dataflow languages are.

Block-based languages have existed since the eighties, but have recently found adoption as tools for programming education. In addition to Scratch, also Alice~\cite{conway_alice:_1994}, Blockly\footnote{\url{https://developers.google.com/blockly/}} and App Inventor~\cite{wolber_app_2011} are block-languages aimed at novice programmers.

In this paper we explore code clones smells in the context of block-based languages. \todo{hier nog wat background over clones, zie ICSE `13 uiteraard}

Clones so far have mostly been studied in the contact of object-oriented, textual source code, but some other directions have been taken, including spreadsheets and SimuLink programs.

% Recently, it has been demonstrated that code smells occur in various alternative programming environments including spreadsheets~\cite{hermans_detecting_2012, hermans_detecting_2014}, Yahoo! Pipes~\cite{stolee_refactoring_2011} and LabView~\cite{chambers_smell-driven_2013}.

We ourselves have recently performed an experiment where we compared the performance of novice Scratch programmers between three

Knowing that code clones can be harmful, we now turn to the question of whether they are common. As such, the goal of this paper is to \textbf{investigate whether code clones are common in public Scratch programs}.  To address this goal we have obtained \nPrograms~public Scratch programs by scraping the list of recent programs \todo{link}. We investigate three types of clones: clones of entire scripts within sprites, clones of entire between sprites, and cloned conditions.

The results show that \todo{todo}

The contributions of this paper are as follows:

\begin{itemize}
\item{A public data set of \nScriptPrograms~Scratch programs (Section \ref{dataset})}
\item{A clone Scratch detection algorithm with an open source implementation (Section \ref{algo})}
\item{An empirical evaluation of the occurrence of three types of clones in Scratch programs (Section \ref{results})}
\end{itemize}

\section{Background and Motivation}
\label{sec:background}
Block-based languages go back to 1986, when Glinert introduced the BLOX language~\cite{e._glinert_towards_1986}. BLOX consists of puzzle-like programming statements that can be combined into programs by combining them both vertically and horizontally. After a decade of little activity into block-based languages, they became a research topic again, starting with Alice~\cite{conway_alice:_1994}. More recently, new block-based languages have gained widespread popularity, especially powered by Scratch~\cite{resnick_scratch:_2009} and Blockly\footnote{\url{https://developers.google.com/blockly/}}. Over 100 million students have tried Blockly via Code.org, and the Scratch repository currently hosts over 12 million projects. Unlike in BLOX, in these new block-based languages the programming blocks can only be combined vertically, resembling textual code more.

Since their introduction, studies have demonstrated the applicability of block-based languages as a tool for education. Scratch, for example, was evaluated with a two-hour introductory programming curriculum for 46 subjects aged 14~\cite{meerbaum-salant_learning_2010}. This study indicated that Scratch could be used to teach computer science concepts: analysis of the pre- and post-tests showed a significant improvement after the Scratch course, although some concepts like variables and concurrency remained hard for students.

Moskal \emph{et al.}~\cite{b._moskal_evaluating_2005} compared computer science students who studied Alice before or during their first programming course to students that only took the introductory computer science course. Their results show that exposure to Alice significantly improved students' grades in the course, and their retention in computer science in general over a two year period. A follow-up study by Cooper \emph{et al.}~\cite{cooper_teaching_2003} obtained similar results, showing that a curriculum in Alice resulted in improved grades and higher retention in computer science.

Most convincingly, Price and Barnes performed a controlled experiment in which students were randomly assigned to either a text-based or a block-based interface in  which they had to perform small programming tasks~\cite{price_comparing_2015}. Their experiment showed that students in the block-based interface were more focused and completed more of the activity's goals in less time.

Summarizing the above, we conclude that block-based languages have a clear potential to be a great tool for introductory programming education, in some cases even outperforming text-based languages.

\section{Relevant Scratch Concepts}
\label{sec:scratch}
This paper is by no means an introduction into Scratch programming, we refer the reader to \cite{brennan_creative_2014} for an extensive overview. To make this paper self-contained, however, we explain a number of relevant concepts in this section.

Scratch is a block-based programming language aimed at children, developed by MIT. Scratch can be used to create games and interactive animations, and is available both as a stand-alone application and as a web application. Figure \ref{fig:ui} shows the Scratch user interface in the Chrome browser.

\subsection{Sprites}
Scratch code is organized by `sprites': two-dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui}\footnote{\url{https://scratch.mit.edu/projects/97086781/}} consists of one sprite, the cat, which is Scratch's default sprite and logo. The code in the sprite will cause the cat to jump up, say ``hello'', and come back down, when the green flag is clicked, and to make the `meow' sound when the space bar is pressed.

\begin{figure}
  \begin{center}
  \includegraphics[width=\columnwidth]{fig/ui.png}
  \caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `motion' in the middle and the code associated with the sprite on the right. The upper right corner shows the actual location of the sprite.}
  \label{fig:ui}
  \end{center}
\end{figure}

\subsection{Events}
Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events. The canonical event is the `when Green Flag clicked', activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. In the Scratch code in Figure \ref{fig:ui} there are two events: `when Green Flag clicked' and `when space key pressed'

\subsection{Scripts}
Source code within sprites is organized in scripts: a script always starts with an event, followed by a number of blocks. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously. For example, the code on the left of Figure \ref{fig:comparison} has five scripts associated with the `when Green Flag clicked' event.

\subsection{Remixing}
Scratch programs can be shared by their creators in the global Scratch repository\footnote{\url{https://scratch.mit.edu/explore/projects/all/}}. Shared Scratch programs can be `remixed' by other Scratch users, which means that a copy of this program is placed in the user's own project collection, and can be then further changed. The `remix tree' of projects is public, so users can track which users remix their programs, a bit similar forking in GitHub. Contrary to forking though, changes upstream cannot be integrated back into the original project.

\begin{figure}
  \begin{center}
  \includegraphics[width=6cm]{fig/makeBlockSmall.png}
  \caption{}
  \label{fig:makeBlock}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width=6cm]{fig/useBlockSmall.png}
  \caption{}
  \label{fig:useBlock}
  \end{center}
\end{figure}

\section{Dataset}
\label{dataset}
We scraped all Scratch programs uploaded \todo{add date range?} from the Scratch website with a scraping program. This resulted in he JSON code for \nPrograms~Scratch programs. In addition to the programs themselves, we also gathered metadata including the numbers of views, loves, favorites and remixes. Out of the \nPrograms~, we failed to analyze 2.367 programs due to various technical difficulties \todo{add details?}

\subsection{Empty programs}

Also, interestingly enough, \nemptyPrograms~were empty (5.7\%), at least in terms of scripts. So Scratch users were sharing program already that did not contain code yet. In some cases \todo{analyze too?} they contained scripts and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default.

\subsection{Block Usage}

\subsection{Block Size}

\subsection{User-defined Blocks}

\subsection{Remixing}

\section{Clones Detection for Scratch programs}
\label{algo}
\label{clone_Detection}

\section{Results}
\label{results}
The overarching research question of this work is how common Scratch users clone in their programs. As explained in Section \ref{clone_Detection} we distinguish three different forms of clones: Exact clones within one sprite, exact clones between sprites of one program, clones of `wait' blocks.

\subsection{Exact clones within one sprite}

\subsection{Exact clones between sprites of one program}

\subsection{Clones of `wait' blocks}

This section presents the results of our clone detection algorithm on the \nScriptPrograms~ with scripts.

\section{Related Work}

\bibliographystyle{IEEEtran}
\bibliography{zotero}

\end{document}

### 2016-04-07 Paper/scratch-batch.tex
added folder for the paper

--- 2016-04-07 Paper/paper.tex
@@
-\documentclass[times,10pt,twocolumn]{article}
+\documentclass{sig-alternate}
+
 \usepackage{latex8}
@@
 \newcommand{\nScriptPrograms}{233.514}
+\newcommand{\nProgramsWithClones}{67.177}
 
@@
 \begin{document}
+%
+% --- Author Metadata here ---
+\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
+%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
+%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
+% --- End of Author Metadata ---
 
 \title{Clones in Block-Based Languages: A Large Scale Analysis of Scratch Programs}
+\numberofauthors{2} 
+\author{
+% 1st. author
+\alignauthor
+Author1\\
+       \affaddr{Institute 1}\\
+% 2nd. author
+\alignauthor
+Author2\\
+       \affaddr{Institute 2}\\
+}
+% There's nothing stopping you putting the seventh, eighth, etc.
+% author on the opening page (as the 'third row') but we ask,
+% for aesthetic reasons that you place these 'additional authors'
+% in the \additional authors block, viz.
 
-\author{Felienne Hermans\\
-       {Delft University of Technology}\\
-      {Mekelweg 4}\\
-       {Delft, the Netherlands}\\
-       {f.f.j.hermans@tudelft.nl}
-}
+% Just remember to make sure that the TOTAL number of authors
+% is the number that will appear on the first page PLUS the
+% number that will appear in the \additionalauthors section.
 
 \maketitle
-
 \begin{abstract}
@@
 \end{abstract}
+
+\todo{terms}
+% A category with the (minimum) three required fields
+\category{H.4}{Information Systems Applications}{Miscellaneous}
+%A category including the fourth, optional field follows...
+\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+
+\terms{Theory}
+
+\keywords{ACM proceedings, \LaTeX, text tagging}
 
@@
 
+\subsection{Signals}
+\todo{uitleggen wat een signal is en evt ook al hoe het een kloon vervangt?}
+
 \subsection{Events}
@@
 In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam.
-In the Scratch code in Figure \ref{fig:ui} there are two events: `when Green Flag clicked' and `when space key pressed'
+In the Scratch code in Figure \ref{fig:ui} there are two events: `when Green Flag clicked' and `when space key pressed'.
 
@@
 
+\subsection{Dead code}
+
 \subsection{User-defined Blocks}
@@
 As explained in Section \ref{clone_Detection} we distinguish three different forms of clones: Exact clones within one sprite, exact clones between sprites of one program, clones of `wait' blocks.
+This section presents the results of our clone detection algorithm on the \nScriptPrograms~ with scripts.
+
+In total \nProgramsWithClones~(26.9\% of programs with scripts) exhibit one of the three types of cloning.The most common type of cloning is copying entire scripts between sprites.
+This occurs in 62.939 programs (25.2\%).
+In 8.220 programs (3.3\%) wait conditions are cloned and in 2.463 files there is a script cloned within a sprite.
 
 \subsection{Exact clones within one sprite}
+In total, we found \todo{x} cases of exact script clones within one sprite.
+Upon further inspection, we found that some of them consisted of unconnected blocks, as depicted in Figure \ref{fig:Unconnected_clones}.
+While this is technically a clone, as this is \emph{dead code} and will never be executed, this could be considered unharmful sketching.
+Looking only at clones starting with an event block, \todo{...} clones remain over \todo{?} files.
+
+\begin{figure}
+  \begin{center}
+  \includegraphics[width=6cm]{fig/Unconnected_clones.png}
+  \caption{Cloned blocks that are not connected to an event. Program id: 12237615}
+  \label{fig:Unconnected_clones}
+  \end{center}
+\end{figure}
 
@@
 \subsection{Clones of `wait' blocks}
+Previous work has established that the cloned `wait' blocks can be harmful when modifying Scratch code, which is the reason why we investigated this particular type of cloning.
+As outlined in section \todo{?} signals can replaced repeated wait blocks, by catching the condition in one place and broadcasting a signal to all other sprites.
+This is why we found it interesting to see in how many of the cases where a wait block was cloned, the creator of the program also used signals in their program.
+This indicates that the concept of signals is known, and hence the duplicate condition is not used because the user is not aware of the alternative.
+In \todo{...} of the cases the program with the duplicated wait block also contained one or more signals.
 
-This section presents the results of our clone detection algorithm on the \nScriptPrograms~ with scripts.
+\subsection{Conclusion}
+
+\subsection{Implications}
+Our results have some interesting implications for designers of educational programming languages.
+
+\subsubsection{Unconnected Blocks}
+In the whole dataset, not just in the programs with clones, we observe unconnected scripts, scripts without an event block and events without associated code.
+This is a form of \emph{dead code}, code that will never be executed.
+Because this dead code caused visual clutter, it would be better to have a separate workspace (much like the `backpack in Scratch' where users can store blocks temporarily and use them later.
+A programming interface could ask actively encourage users to move unconnected block when they exit the environment to encourage a `clean' working space.
+
+\subsubsection{Code Clones Between Sprites}
+With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common.
+In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of \todo{right word?
+procedures} between scripts, only within them.
+So in many cases there is no other way to share the functionality than making a copy.
+We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users needs this functionality.
+
+\subsubsection{Refactoring support for duplicate wait blocks}
+In todo{...} of the cases, the programs with duplicate conditions, also signals were present. This means that it is feas
+
+\section{Discussion}
 
 \section{Related Work}
+\label{sec:related}
+
+\subsection{Code Smells}
+Efforts related to our research include works on code smells, initiated by the work by Fowler~\cite{fowler_refactoring:_1999}.
+His book gives an overview of code smells and corresponding refactorings.
+Fowler's work was followed by efforts focused on the automatic identification of code smells by means of metrics.
+Marinescu~\cite{marinescu_detecting_2001} for instance, uses metrics to identify \emph{suspect} classes: classes which could have design flaws.
+Lanza and Marinescu~\cite{lanza_object-oriented_2006} explain this methodology in more detail.
+Alves \emph{et al.}~\cite{alves_deriving_2010} focus on a strategy to obtain thresholds for metrics from a benchmark.
+Olbrich \emph{et al.} furthermore investigates the changes in smells over time, and discusses their impact~\cite{olbrich_evolution_2009}.
+Moha \emph{et al.}~\cite{moha_decor:_2010} designed the `DECOR' method which automatically generates a smell detection algorithms from specifications.
+The CCFinder tool~\cite{kamiya_ccfinder:_2002} finally, aims at detecting clones in source code, which are similar to our \dup smell.
+
+\subsection{Smells beyond the OO paradigm}
+In recent times, code smells have been applied to programs outside of the regular programming domain.
+In our past work, we have, for example, studied code smells within spreadsheets, both at the formula level~\cite{hermans_detecting_2014} and between worksheets~\cite{hermans_detecting_2012}.
+
+More recently, we compared two datasets: one containing spreadsheets which users found unmaintainable, and a version of the same spreadsheets rebuilt by professional spreadsheet developers.
+The results show that the improved versions suffered from smells to a lesser degree, increasing our confidence that presence of smells indeed coincides with users finding spreadsheets hard to maintain~\cite{jansen_code_2015}.
+
+In addition to spreadsheets, code smells have also been studied in the context of Yahoo!
+Pipes a web mashup tool.
+An experiment demonstrated that users preferred the non-smelly versions of Yahoo Pipes programs~\cite{stolee_refactoring_2011}.
+
+%\subsection{Agile practices in education}
+%A third related line of work are research directions where other methods from agile development, like testing and refactoring, have been applied in education.
+For example, the use of test-driven development has been studied within the context of programming education at university level, using text-based languages in several studies.
+Most controlled experiments resulted in increased code quality for the group of students using TDD, for example [13] reported on a 35\% reduction in defects, while [14] found 45\%.
+For an extensive overview of TDD approaches in university programming courses, see [15].
+\todo{referenties fixen}
+
+\subsection{Quality of Scratch programs}
+Finally, there are other works on the quality of Scratch programs.
+There is for example the Hairball Scratch extension~\cite{boe_hairball:_2013}, which is a lint-like static analysis tool for Scratch that can detect, for example, unmatched broadcast and receive blocks, infinite loops and duplication.
+An evaluation of 100 Scratch programs showed that Scratch programs indeed suffer from duplication and bad naming~\cite{moreno_automatic_2014}.
+
+Most related to our study is the work by Moreno \emph{et al.}~\cite{moreno-leon_dr._2015} who gave automated feedback on Scratch programs to 100 children aged 10 to 14.
+Their results demonstrated that feedback on code quality helped improve students' programming skills.
 

### 2016-04-15 Paper/paper.tex
Updated paper style, title and bibliography

@@
-\documentclass[times,10pt,twocolumn]{article}
-\usepackage{latex8}
-\usepackage{times}
+\documentclass{sig-alternate}
+
+\usepackage{url}
 \usepackage{graphicx}
 
-\usepackage{amsmath}
-\usepackage{tabularx}
-\usepackage{booktabs}
-\usepackage{colortbl}
-\usepackage{tikz}
-\usepackage{xcolor}
-\usepackage{url}
-\usepackage{balance}
-
-\newcommand{\ub}{user defined-block}
 \newcommand{\nPrograms}{250.166}
@@
 \newcommand{\nScriptPrograms}{233.514}
-
-\newcommand{\numkids}{61~}
-\newcommand{\lm}{\emph{Long Method}~}
+\newcommand{\nProgramsWithClones}{67.177}
 \newcommand{\dup}{\emph{Duplication}~}
@@
 \begin{document}
+%
+% --- Author Metadata here ---
+%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
+% --- End of Author Metadata ---
 
-\title{Clones in Block-Based Languages: A Large Scale Analysis of Scratch Programs}
+\title{How kids code and how we know it:\\An exploratory study in the Scratch project repository }
 
-\author{Felienne Hermans\\
-       {Delft University of Technology}\\
-      {Mekelweg 4}\\
-       {Delft, the Netherlands}\\
-       {f.f.j.hermans@tudelft.nl}
+\numberofauthors{1}
+\author{
+% 1st. author
+\alignauthor
+(anonymized submission)
+%Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
+%       \affaddr{Institute for Clarity in Documentation}\\
+%       \affaddr{1932 Wallamaloo Lane}\\
+%       \affaddr{Wallamaloo, New Zealand}\\
+%       \email{trovato@corporation.com}
+% 2nd. author
+%\alignauthor
+%G.K.M. Tobin\titlenote{The secretary disavows
+%any knowledge of this author's actions.}\\
+%       \affaddr{Institute for Clarity in Documentation}\\
+%       \affaddr{P.O. Box 1212}\\
+%       \affaddr{Dublin, Ohio 43017-6221}\\
+%       \email{webmaster@marysville-ohio.com}
 }
 
+%\date{30 July 1999}
+
 \maketitle
-
 \begin{abstract}
@@
 \end{abstract}
+
+% A category with the (minimum) three required fields
+%\category{H.4}{Information Systems Applications}{Miscellaneous}
+%A category including the fourth, optional field follows...
+%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+
+\terms{terms}
+
+\keywords{keywords}
 
@@
 \end{itemize}
+
+Research questions?
 
@@
 \label{clone_Detection}
+\todo{How it works, Fenia, I can write this if you want}
 
@@
 As explained in Section \ref{clone_Detection} we distinguish three different forms of clones: Exact clones within one sprite, exact clones between sprites of one program, clones of `wait' blocks.
+This section presents the results of our clone detection algorithm on the \nScriptPrograms~ with scripts.
+
+In total \nProgramsWithClones~(26.9\% of programs with scripts) exhibit one of the three types of cloning.The most common type of cloning is copying entire scripts between sprites.
+This occurs in 62.939 programs (25.2\%).
+In 8.220 programs (3.3\%) wait conditions are cloned and in 2.463 files there is a script cloned within a sprite.
 
 \subsection{Exact clones within one sprite}
+In total, we found \todo{x} cases of exact script clones within one sprite.
+Upon further inspection, we found that some of them consisted of unconnected blocks, as depicted in Figure \ref{fig:Unconnected_clones}.
+While this is technically a clone, as this is \emph{dead code} and will never be executed, this could be considered unharmful sketching.
+Looking only at clones starting with an event block, \todo{...} clones remain over \todo{?} files.
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=6cm]{fig/Unconnected_clones.png}
+		\caption{Cloned blocks that are not connected to an event. Program id: 12237615}
+		\label{fig:Unconnected_clones}
+	\end{center}
+\end{figure}
 
@@
 \subsection{Clones of `wait' blocks}
+Previous work has established that the cloned `wait' blocks can be harmful when modifying Scratch code, which is the reason why we investigated this particular type of cloning.
+As outlined in section \todo{?} signals can replaced repeated wait blocks, by catching the condition in one place and broadcasting a signal to all other sprites.
+This is why we found it interesting to see in how many of the cases where a wait block was cloned, the creator of the program also used signals in their program.
+This indicates that the concept of signals is known, and hence the duplicate condition is not used because the user is not aware of the alternative.
+In \todo{...} of the cases the program with the duplicated wait block also contained one or more signals.
 
-This section presents the results of our clone detection algorithm on the \nScriptPrograms~ with scripts.
+\subsection{Conclusion}
+
+\subsection{Implications}
+Our results have some interesting implications for designers of educational programming languages.
+
+\subsubsection{Unconnected Blocks}
+In the whole dataset, not just in the programs with clones, we observe unconnected scripts, scripts without an event block and events without associated code.
+This is a form of \emph{dead code}, code that will never be executed.
+Because this dead code caused visual clutter, it would be better to have a separate workspace (much like the `backpack in Scratch' where users can store blocks temporarily and use them later.
+A programming interface could ask actively encourage users to move unconnected block when they exit the environment to encourage a `clean' working space.
+
+\subsubsection{Code Clones Between Sprites}
+With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common.
+In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of \todo{right word?
+procedures} between scripts, only within them.
+So in many cases there is no other way to share the functionality than making a copy.
+We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users needs this functionality.
+
+\subsubsection{Refactoring support for duplicate wait blocks}
+In todo{...} of the cases, the programs with duplicate conditions, also signals were present. This means that it is feas
+
+\section{Discussion}
 
 \section{Related Work}
+\label{sec:related}
 
-\bibliographystyle{IEEEtran}
+\subsection{Code Smells}
+Efforts related to our research include works on code smells, initiated by the work by Fowler~\cite{fowler_refactoring:_1999}.
+His book gives an overview of code smells and corresponding refactorings.
+Fowler's work was followed by efforts focused on the automatic identification of code smells by means of metrics.
+Marinescu~\cite{marinescu_detecting_2001} for instance, uses metrics to identify \emph{suspect} classes: classes which could have design flaws.
+Lanza and Marinescu~\cite{lanza_object-oriented_2006} explain this methodology in more detail.
+Alves \emph{et al.}~\cite{alves_deriving_2010} focus on a strategy to obtain thresholds for metrics from a benchmark.
+Olbrich \emph{et al.} furthermore investigates the changes in smells over time, and discusses their impact~\cite{olbrich_evolution_2009}.
+Moha \emph{et al.}~\cite{moha_decor:_2010} designed the `DECOR' method which automatically generates a smell detection algorithms from specifications.
+The CCFinder tool~\cite{kamiya_ccfinder:_2002} finally, aims at detecting clones in source code, which are similar to our \dup smell.
+
+\subsection{Smells beyond the OO paradigm}
+In recent times, code smells have been applied to programs outside of the regular programming domain.
+In our past work, we have, for example, studied code smells within spreadsheets, both at the formula level~\cite{hermans_detecting_2014} and between worksheets~\cite{hermans_detecting_2012}.
+
+More recently, we compared two datasets: one containing spreadsheets which users found unmaintainable, and a version of the same spreadsheets rebuilt by professional spreadsheet developers.
+The results show that the improved versions suffered from smells to a lesser degree, increasing our confidence that presence of smells indeed coincides with users finding spreadsheets hard to maintain~\cite{jansen_code_2015}.
+
+In addition to spreadsheets, code smells have also been studied in the context of Yahoo!
+Pipes a web mashup tool.
+An experiment demonstrated that users preferred the non-smelly versions of Yahoo Pipes programs~\cite{stolee_refactoring_2011}.
+
+%\subsection{Agile practices in education}
+%A third related line of work are research directions where other methods from agile development, like testing and refactoring, have been applied in education.
+For example, the use of test-driven development has been studied within the context of programming education at university level, using text-based languages in several studies.
+Most controlled experiments resulted in increased code quality for the group of students using TDD, for example [13] reported on a 35\% reduction in defects, while [14] found 45\%.
+For an extensive overview of TDD approaches in university programming courses, see [15].
+\todo{referenties fixen}
+
+\subsection{Quality of Scratch programs}
+Finally, there are other works on the quality of Scratch programs.
+There is for example the Hairball Scratch extension~\cite{boe_hairball:_2013}, which is a lint-like static analysis tool for Scratch that can detect, for example, unmatched broadcast and receive blocks, infinite loops and duplication.
+An evaluation of 100 Scratch programs showed that Scratch programs indeed suffer from duplication and bad naming~\cite{moreno_automatic_2014}.
+
+Most related to our study is the work by Moreno \emph{et al.}~\cite{moreno-leon_dr._2015} who gave automated feedback on Scratch programs to 100 children aged 10 to 14.
+Their results demonstrated that feedback on code quality helped improve students' programming skills.
+
+\bibliographystyle{abbrv}
 \bibliography{zotero}
-
 \end{document}

### 2016-04-15 Paper/paper.tex
Updated abstract and keywords

@@
 \begin{abstract}
-Recently, block-based programming languages like Alice, Scratch and Blockly have become popular tools for programming education.
-There is substantial research showing that block-based languages are suitable for early programming education.
-But do block-based programs suffer from code clones \emph{smelly} too?
-In a recent controlled experiment, we found that code clones in Scratch indeed hamper novice Scratch programmers.
-In this paper we explore how commonly code clones occur in Scratch.
-To that end we have scraped Scratch public repository and retrieved 250.000 programs.
-We have analyzed three types of clones within those programs': clones of entire scripts within sprites, clones of entire between sprites, and cloned conditions.
-We find that those clones occur in ...
-of programs.
+Block-based programming languages like Scratch, Alice and Blockly are becoming increasingly common as introductory languages in programming education.
+There is substantial research showing that those visual programming environments are suitable for teaching programming concepts.
+In this paper we explore the characteristics of Scratch programs.
+To that end we have scraped the Scratch public repository and retrieved 250.000 projects.
+We present an analysis of those projects in terms of complexity, used abstractions and programming concepts, and coding style.
+We found that programming abstraction concepts like functions are not commonly used.
+We further investigate the presence of code smells and bad programming practices, including code duplication, dead code, long method and large class smells.
+Our findings indicate that Scratch programs suffer from code smells and especially code duplication.
 \end{abstract}
@@
 
-\keywords{keywords}
+\keywords{Scratch, programming practices, code smells, static analysis}
 

### 2016-04-17 csvCleanup/Program.cs
Program for preparing analysis.csv for import

@@ -1,0 +1,98 @@
+﻿using System;
+using System.Collections.Generic;
+using System.IO;
+using System.Linq;
+using System.Text;
+using System.Threading.Tasks;
+
+namespace csvCleanup
+{
+    class Program
+    {
+        static void Main(string[] args)
+        {
+            //using (var input = File.OpenText(@"C:\Users\Efthimia\Documents\GitHub\Kragle\KragleData\procedures.csv"))
+            //using (var output = new StreamWriter(@"C:\Users\Efthimia\Documents\GitHub\Kragle\KragleData\proceduresReplaced.txt", true, Encoding.UTF8))
+            //{
+            //    string line;
+            //    while (null != (line = input.ReadLine()))
+            //    {
+            //        var procdelf = line.IndexOf(",procDef,");
+            //        var start = line.Substring(0, procdelf).Split(',');
+            //        var end = line.Substring(procdelf + 1, line.Length - procdelf - 1).Split(',');
+
+            //        var outputLine = start[0] + ",\"" + start[1].Replace("\"", "") + "\",\"" + end[1].Replace("\"", "") + "\"," + end[2];
+
+            //        output.WriteLine(outputLine);
+
+            //    }
+            //}
+
+            var write = true;
+
+            using (var input = File.OpenText(@"C:\Users\Efthimia\Documents\GitHub\Kragle\KragleData\analysisutf8.txt"))
+            using (var output = new StreamWriter(@"C:\Users\Efthimia\Documents\GitHub\Kragle\KragleData\analysisutf8Replaced.txt", true, Encoding.UTF8))
+            {
+                string line;
+                while (null != (line = input.ReadLine()))
+                {
+                    //if (line.StartsWith("99850394,5-118.5,23,sprite"))
+                    //{
+                    //    write = true;
+                    //    line = line;//.Replace("The \"Exit\"", "The Exit");
+                    //}
+                    line = line.Replace(",\"I have no talent.\", said Sprite 44.,", ",\"I have no talent., said Sprite 44.\",");
+                    line = line.Replace(",\"I lagged so much, so I decided, \"Whatever\"\",", ",\"I lagged so much, so I decided, Whatever\",");
+                    if (line.Contains(",sprite,\",jefsdfaorjsdnf\""))
+                    {
+                        line = line.Replace(",sprite,\",jefsdfaorjsdnf\",",",sprite,\"jefsdfaorjsdnf\",");
+                    }
[…]

### 2016-04-17 csvCleanup/Properties/AssemblyInfo.cs

@@ -1,0 +1,36 @@
+﻿using System.Reflection;
+using System.Runtime.CompilerServices;
+using System.Runtime.InteropServices;
+
+// General Information about an assembly is controlled through the following 
+// set of attributes. Change these attribute values to modify the information
+// associated with an assembly.
+[assembly: AssemblyTitle("ConsoleApplication1")]
+[assembly: AssemblyDescription("")]
+[assembly: AssemblyConfiguration("")]
+[assembly: AssemblyCompany("")]
+[assembly: AssemblyProduct("ConsoleApplication1")]
+[assembly: AssemblyCopyright("Copyright ©  2016")]
+[assembly: AssemblyTrademark("")]
+[assembly: AssemblyCulture("")]
+
+// Setting ComVisible to false makes the types in this assembly not visible 
+// to COM components.  If you need to access a type in this assembly from 
+// COM, set the ComVisible attribute to true on that type.
+[assembly: ComVisible(false)]
+
+// The following GUID is for the ID of the typelib if this project is exposed to COM
+[assembly: Guid("d4cddfd1-c5d5-40de-91e0-9ae37983aa44")]
+
+// Version information for an assembly consists of the following four values:
+//
+//      Major Version
+//      Minor Version 
+//      Build Number
+//      Revision
+//
+// You can specify all the values or you can default the Build and Revision Numbers 
+// by using the '*' as shown below:
+// [assembly: AssemblyVersion("1.0.*")]
+[assembly: AssemblyVersion("1.0.0.0")]
+[assembly: AssemblyFileVersion("1.0.0.0")]

### 2016-04-17 Paper/paper.tex
Updated paper structure and research questions

@@
 \newcommand{\nemptyPrograms}{14.285}
-\newcommand{\nScriptPrograms}{233.514}
+\newcommand{\nScriptPrograms}{233.513}
 \newcommand{\nProgramsWithClones}{67.177}
@@
 
-In this paper we explore code clones smells in the context of block-based languages. \todo{hier nog wat background over clones, zie ICSE `13 uiteraard}
+refer to:
+Research on the effectiveness of those languages for programming education
+Research on programming habits
+Our research on smells
 
-Clones so far have mostly been studied in the contact of object-oriented, textual source code, but some other directions have been taken, including spreadsheets and SimuLink programs.
+The goal of this paper is to obtain a deep understanding of how people program in Scratch and to analyze the characteristics of Scratch programs.
+Moreover, knowing that bad programming habits and code smells can be harmful, we also want to explore whether they are common.
+Specifically, the research questions that we are trying to answer are:
 
-% Recently, it has been demonstrated that code smells occur in various alternative programming environments including spreadsheets~\cite{hermans_detecting_2012, hermans_detecting_2014}, Yahoo!
-Pipes~\cite{stolee_refactoring_2011} and LabView~\cite{chambers_smell-driven_2013}.
+\begin{itemize}
+\item[RQ1] What are the size and complexity characteristics of Scratch programs?
+\item[RQ2] Which coding abstractions and programming concepts and features are commonly used when programming in the Scratch environment?
+\item[RQ3] How are coding style conventions applied in Scratch?
+\item[RQ4] Which code smells and bad programming practices are common in Scratch programs?
+\end{itemize}
 
-We ourselves have recently performed an experiment where we compared the performance of novice Scratch programmers between three
-
-Knowing that code clones can be harmful, we now turn to the question of whether they are common.
-As such, the goal of this paper is to \textbf{investigate whether code clones are common in public Scratch programs}.
-To address this goal we have obtained \nPrograms~public Scratch programs by scraping the list of recent programs \todo{link}.
-We investigate three types of clones: clones of entire scripts within sprites, clones of entire between sprites, and cloned conditions.
-
-The results show that \todo{todo}
+Our study is based on data from the Scratch program repository.
+By scraping the list of recent programs \todo{link}, we have obtained \nPrograms~public Scratch programs and performed static analysis on them.
 
@@
 	\item{A public data set of \nScriptPrograms~Scratch programs (Section \ref{dataset})}
-	\item{A clone Scratch detection algorithm with an open source implementation (Section \ref{algo})}
-	\item{An empirical evaluation of the occurrence of three types of clones in Scratch programs (Section \ref{results})}
+	\item{An empirical evaluation of the data set for answering the aforementioned research questions}
 \end{itemize}
-
-Research questions?
 
@@
 
-\section{Relevant Scratch Concepts}
+\subsection{Relevant Scratch Concepts}
 \label{sec:scratch}
@@
 
-\subsection{Sprites}
+\subsubsection{Sprites}
 Scratch code is organized by `sprites': two-dimensional pictures each having their own associated code.
@@
 
-\subsection{Signals}
+\subsubsection{Signals}
 \todo{uitleggen wat een signal is en evt ook al hoe het een kloon vervangt?}
 
-\subsection{Events}
+\subsubsection{Events}
 Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events.
@@
 
-\subsection{Scripts}
+\subsubsection{Scripts}
 Source code within sprites is organized in scripts: a script always starts with an event, followed by a number of blocks.
@@
 
-\subsection{Remixing}
+\subsubsection{Remixing}
 Scratch programs can be shared by their creators in the global Scratch repository\footnote{\url{https://scratch.mit.edu/explore/projects/all/}}.
@@
 
-\section{Dataset}
+\section{Research design and dataset}
+\label{researchDesign}
+
+The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch programs.
+To answer our research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch repository.
+The dataset is described in Section \ref{dataset}.
+Below, we present how we approached each research question.
+
+\textbf{RQ1} What are the size and complexity characteristics of Scratch programs?
+quantitative analysis, provide and analyze descriptive statistics on the size, complexity metrics, statistical analysis .
+mentioned taht it was imported to db and done and statistical data from db queries?
+The outcomes are presented in Section...
+
+\textbf{RQ2} Which coding abstractions and programming concepts and features are commonly used when programming in the Scratch environment?
+Using the code in the dataset, we perform a detailed statistical analysis of code characteristics to answer RQ2 (Section ??)
+
+\textbf{RQ3} How are coding style conventions applied in Scratch? to examine if coding style conventions are used, we qualitatively analyze..
+
+\textbf{RQ4} Which code smells and bad programming practices are common in Scratch programs?
+
+\subsection{Dataset}
 \label{dataset}
+
 We scraped all Scratch programs uploaded \todo{add date range?} from the Scratch website with a scraping program.
@@
 
-\subsection{Empty programs}
+describe:
+in what form is the dataset we analyzed and we are making available (json code + analysis.csv? what fields does it contain?)
+how we obtained the programs
+how we analyzed the programs to obtain analysis.csv
+
+\section{Results}
+
+\subsection{Program size and complexity}
+Size -per Project:sprites, blocks and locs, events/triggers
+nestendness	
+cyclomatic complexity
+
+\subsubsection{Empty programs}
 
@@
 
-\subsection{Block Usage}
+\subsection{Programming abstractions and concepts}
+loops	
+logic operations	
+variables	
+lists	
+user interativity	
+functions - arguments
+parallelism	- wait blocks, briadcast, receive
 
-\subsection{Block Size}
+\subsection{Coding style}
+naming -functions, sprites, variables
+positioning - overlapping /visibility
 
-\subsection{Dead code}
+\subsection{Code smells}
+dead code
+duplicates/clones -graph, number of overlapping positioning, number of dead code duplicates
+long method
+large class
+conditional complexity
+Dublicated wait block-same trigger caught by more than 1
 
-\subsection{User-defined Blocks}
+\subsubsection{Clones Detection for Scratch programs}
+\label{clone_Detection}
 
-\subsection{Remixing}
-
-\section{Clones Detection for Scratch programs}
-\label{algo}
-\label{clone_Detection}
-\todo{How it works, Fenia, I can write this if you want}
-
-\section{Results}
-\label{results}
-The overarching research question of this work is how common Scratch users clone in their programs.
-As explained in Section \ref{clone_Detection} we distinguish three different forms of clones: Exact clones within one sprite, exact clones between sprites of one program, clones of `wait' blocks.
+We distinguish three different forms of clones: Exact clones within one sprite, exact clones between sprites of one program, clones of `wait' blocks.
 This section presents the results of our clone detection algorithm on the \nScriptPrograms~ with scripts.
@@
 
-\subsection{Exact clones within one sprite}
+\paragraph{Exact clones within one sprite}
 In total, we found \todo{x} cases of exact script clones within one sprite.
@@
 
-\subsection{Exact clones between sprites of one program}
+\paragraph{Exact clones between sprites of one program}
 
-\subsection{Clones of `wait' blocks}
+\paragraph{Clones of `wait' blocks}
 Previous work has established that the cloned `wait' blocks can be harmful when modifying Scratch code, which is the reason why we investigated this particular type of cloning.
@@
 
-\subsection{Conclusion}
+\section{Discussion}
 
@@
 In todo{...} of the cases, the programs with duplicate conditions, also signals were present. This means that it is feas
-
-\section{Discussion}
 

### 2016-04-18 Paper/paper.tex
From 4 to 3 research questions

@@
 \newcommand{\nemptyPrograms}{14.285}
-\newcommand{\nScriptPrograms}{233.513}
+\newcommand{\nScriptPrograms}{233.491}
 \newcommand{\nProgramsWithClones}{67.177}
@@
 \item[RQ2] Which coding abstractions and programming concepts and features are commonly used when programming in the Scratch environment?
-\item[RQ3] How are coding style conventions applied in Scratch?
-\item[RQ4] Which code smells and bad programming practices are common in Scratch programs?
+\item[RQ3] How common are code smells and bad programming practices in Scratch programs?
 \end{itemize}
@@
 \textbf{RQ2} Which coding abstractions and programming concepts and features are commonly used when programming in the Scratch environment?
-Using the code in the dataset, we perform a detailed statistical analysis of code characteristics to answer RQ2 (Section ??)
+Using the code in the dataset, we perform a detailed statistical analysis of code characteristics to answer RQ2 (Section ??).
+We manually assigned types to commands/blocks.
 
-\textbf{RQ3} How are coding style conventions applied in Scratch? to examine if coding style conventions are used, we qualitatively analyze..
-
-\textbf{RQ4} Which code smells and bad programming practices are common in Scratch programs?
+\textbf{RQ3} Which code smells and bad programming practices are common in Scratch programs?  to examine if coding style conventions are used, we qualitatively analyze..
 
@@
 Size -per Project:sprites, blocks and locs, events/triggers
+lines in stage/sprites/procedures
 nestendness	
 cyclomatic complexity
+
+lines do not count the ending line of enclosures
+only sprites with code are counted
+indentation: 1 for top-level, for nested (rest-1)/2 (goes 1,3,5, etc), for others normal
 
@@
 
-\subsection{Coding style}
-naming -functions, sprites, variables
-positioning - overlapping /visibility
-
 \subsection{Code smells}
@@
 Dublicated wait block-same trigger caught by more than 1
+
+coding style -naming -functions, sprites, variables
+code block positioning - overlapping /visibility
 

### 2016-04-18 Paper/paper.tex
added info on the scraper

@@
 \label{dataset}
+\subsubsection{Obtaining Scratch programs}
+We obtained the set of Scratch programs by using a scraping program.
+Our scraping program called Kragle starts by reading \url{https://scratch.mit.edu/explore/projects/all/} and thus obtaining program ids of programs that were most recently shared.
+Subsequently, Kragle obtains the JSON code for each of the listed programs.
+In addition to the programs themselves, Kragle also gathers metadata including the numbers of views, loves, favorites and remixes.
 
-We scraped all Scratch programs uploaded \todo{add date range?} from the Scratch website with a scraping program.
-This resulted in he JSON code for \nPrograms~Scratch programs.
-In addition to the programs themselves, we also gathered metadata including the numbers of views, loves, favorites and remixes.
-Out of the \nPrograms~, we failed to analyze 2.367 programs due to various technical difficulties \todo{add details?}
+ % For a given id $x$ a Scratch programs JSON representation can be obtained via \url{https://cdn.projects.scratch.mit.edu/internalapi/project/x/get}.
+Does not fit on the page nicely and might also not be needed?
 
-describe:
-in what form is the dataset we analyzed and we are making available (json code + analysis.csv? what fields does it contain?)
-how we obtained the programs
-how we analyzed the programs to obtain analysis.csv
+We ran Kragle on March 2nd 2016 and let it running for 24 hours, when it had obtained a little over 250.000 programs.
+Out of the \nPrograms~, we failed to analyze 2.367 programs due to various technical difficulties \todo{add details, I will check a few to see what went wrong - Felienne}
+
+Kragle, as well as all scraped programs and our analysis files are available via \todo{url} \todo{I do not think we need to describe the fields in the paper, but we need a good readme with the data (I can do that)}
+
+\subsubsection{Analyzing the Scratch programs}
+Once we obtained the Scratch programs, we parsed the JSON files according to the specification of the format\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}}.
+This resulted in a list if used blocks per program, with the sprites and the stage of the program.
+We also assign types to all the blocks, both their shapes and the tab they originate from.
+For example, When Green Flag Clicked is a \emph{Hat block} from the \emph{Events category}.
 

### 2016-04-18 Kragle/Program.cs

@@ -14,9 +14,9 @@ namespace Kragle
    {
       static void Main(string[] args)
       {
-         string param1 = args[0];
+         string param1 = "-p";//args[0];
 
-         string path = args[1];
+         string path = @"C:\Users\Felienne\Dropbox\Code\KragleData\testData\";//args[1];
 
          if (param1=="-p")
          {

### 2016-04-18 Paper/paper.tex
Section Results: Size and Complexity

--- 2016-04-18 Paper/paper.tex
@@
 
-\newcommand{\nPrograms}{250.166}
-\newcommand{\nemptyPrograms}{14.285}
-\newcommand{\nScriptPrograms}{233.491}
-\newcommand{\nProgramsWithClones}{67.177}
+\newcommand{\nPrograms}{250,166}
+\newcommand{\nAnalyzedPrograms}{247,798}
+\newcommand{\nemptyPrograms}{14,307}
+\newcommand{\nScriptPrograms}{233,491}
+\newcommand{\nProgramsWithClones}{67,177}
+\newcommand{\nLOC}{36,085,654}
+\newcommand{\nscripts}{4,049,356}
 \newcommand{\dup}{\emph{Duplication}~}
@@
 
+In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms Scratch projects in our dataset.
+
 \subsection{Program size and complexity}
-Size -per Project:sprites, blocks and locs, events/triggers
-lines in stage/sprites/procedures
-nestendness	
-cyclomatic complexity
 
-lines do not count the ending line of enclosures
-only sprites with code are counted
-indentation: 1 for top-level, for nested (rest-1)/2 (goes 1,3,5, etc), for others normal
+The dataset is found to contain a relatively small number of projects without any code: \nemptyPrograms (5.77\%).
+Through random sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default.
+Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.
 
-\subsubsection{Empty programs}
+\begin{table*}[ht]
+	\centering
+	\begin{tabular}{lrrrrrr}
+		\hline
+		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
+		\hline
+		Sprites with code per project&5.68&1&1&2&5&525\\
+		Scripts per project&17.35&1&2&5&12&3038\\
+		Maximum indentation per project&4.56&1&3&3&5&488\\
+		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
+		LOC in Stage&4.80&0&0&0&3&2613\\
+		LOC in Sprites&115.57&0&10&26&68&34613\\
+		LOC in Procedures&34.17&0&0&0&0&20552\\
+		McCabe Cyclomatic Complexity per script&1.58&0&0&0&0&246\\
+		\hline
+	\end{tabular}
+	\caption{Size-related statistics from the dataset of \nScriptPrograms  non-empty Scratch projects}
+	\label{tbl-size}
+\end{table*}
 
-Also, interestingly enough, \nemptyPrograms~were empty (5.7\%), at least in terms of scripts.
-So Scratch users were sharing program already that did not contain code yet.
-In some cases \todo{analyze too?} they contained scripts and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default.
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
+	\caption{Histogram of the number of sprites in the analyzed projects}
+	\label{fig:size-sprites}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
+	\caption{Histogram of the number of scripts in the analyzed projects}
+	\label{fig:size-scripts}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
+	\caption{Histogram of the lines of code in the analyzed projects}
+	\label{fig:size-locs}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/1indent}
+	\caption{Histogram of the maximum code indentation level in the analyzed projects}
+	\label{fig:size-indent}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
+	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts analyzed scripts}
+	\label{fig:cyclomatic}
+\end{figure}
+
+In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity.
+We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project, where each script is a code-block of interconnected Scratch blocks.
+For example, in Figure \ref{fig:ui} the cat sprite contains 2 scripts.
+We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines.
+The maximum code indentation level is used as a code complexity metric.
+It was calculated using the JSON array depth?
+\todo{felienne, how did we measure this?}.
+This corresponds to code indentation level returning 1 for all first-level blocks and increasing by 2 for block enclosures in loops or conditional statements.
+Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.
+
+We find that the majority of Scratch projects are small; 75\% of the projects have up to 5 sprites, 12 scripts and 76 lines of code, while one fourth of the projects have up to 12 lines of code.
+On the other end, 5\% of the projects (11,712) are found to have more than 18 sprites and 4.8\% (11,214) consist of more than 500 lines of code.
+The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.
+
+The lines of code metric was further analyzed to understand code placement.
+As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC lines of code---is written within sprites.
+An additional 3.1\% of the total lines are found in the stage class.
+More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects.
+
+As a complexity metric we also used the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code.
+This is calculated per script by counting the number of decision points in the script plus one.
+In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks.
+The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}.
+The majority (78.33\%) of \nscripts scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point.
+The complexity is higher, over 4, for 3.67\% of the scripts.
+The analysis also highlighted 209 scripts with cyclomatic complexity over 100 and up to 246.
 

### 2016-04-18 Paper/paper.tex
Added section summary box

@@
 
+\noindent
+\fbox{
+	\begin{minipage}{0.46\textwidth}
+		\emph{RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points.
+Most code is written in sprites.
+Few projects (7.7\%) use procedures, but they use them a lot.
+There exist surprisingly large and complex projects.}
+	\end{minipage}}
+
 \subsection{Programming abstractions and concepts}

### 2016-04-19 Paper/paper.tex
Added more info on procedures complexity

@@
 		LOC in Procedures&34.17&0&0&0&0&20552\\
-		McCabe Cyclomatic Complexity per script&1.58&0&0&0&0&246\\
+		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
+		McCabe CC per Procedure script&3.75&1&1&2&4&183\\
 		\hline
@@
 More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects.
+Those projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.
 
@@
 The complexity is higher, over 4, for 3.67\% of the scripts.
-The analysis also highlighted 209 scripts with cyclomatic complexity over 100 and up to 246.
+The analysis also highlighted 209 scripts with cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.
+Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.
 
@@
 Most code is written in sprites.
-Few projects (7.7\%) use procedures, but they use them a lot.
+Few projects (7.7\%) use procedures, but they use them a lot and for more complex code.
 There exist surprisingly large and complex projects.}
@@
 \subsection{Programming abstractions and concepts}
+
+functions - arguments
+variables
+
 loops	
 logic operations		
-variables	
 lists	
 user interativity	
-functions - arguments
 parallelism	- wait blocks, briadcast, receive
+
+Control abstraction is the abstraction of actions while data abstraction is that of data structures.
+
+Control abstraction involves the use of subprograms and related concepts control flows
+Data abstraction allows handling data bits in meaningful ways. For example, it is the basic motivation behind datatype.
 

### 2016-04-19 Paper/paper.tex
Input on functions

@@
 		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
-		LOC in Stage&4.80&0&0&0&3&2613\\
-		LOC in Sprites&115.57&0&10&26&68&34613\\
-		LOC in Procedures&34.17&0&0&0&0&20552\\
+		LOC in Stage per project&4.80&0&0&0&3&2613\\
+		LOC in Sprites per project&115.57&0&10&26&68&34613\\
+		LOC in Procedures per project&34.17&0&0&0&0&20552\\
 		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
-		McCabe CC per Procedure script&3.75&1&1&2&4&183\\
+		McCabe CC per procedure script&3.75&1&1&2&4&183\\
+		Procedures per project with procedures&11.50&1&1&2&6&847\\
+		Arguments per Procedure&0.95&0&0&0&1&53\\
+		Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
+		Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
+		Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
+		Calls per procedure&2.14&0&1&1&2&526\\
+		Scripts with calls per procedure&1.13&0&1&1&1&59\\
 		\hline
 	\end{tabular}
-	\caption{Size-related statistics from the dataset of \nScriptPrograms  non-empty Scratch projects}
+	\caption{Summary statistics from the dataset of \nScriptPrograms  non-empty Scratch projects}
 	\label{tbl-size}
@@
 
-functions - arguments
+As a first programming abstraction we investigated the use of procedures.
+In the dataset we found 206,799 procedures in 17,979 projects.
+As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of those projects having up to 2.
+Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects.
+Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\%\% have only one (shown in Figure \ref{fig:procedurearguments}).
+The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/6proceduresperproject}
+	\caption{Histogram of the number of procedures for the 17,979 projects that include at least one}
+	\label{fig:proceduresperproject}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/5arguments}
+	\caption{Histogram of the number of arguments for the procedures in the dataset}
+	\label{fig:procedurearguments}
+\end{figure}
+
+The use of procedures in projects was further investigated through the procedure calls, summarized in Figure \ref{fig:procedurecalls}.
+A significant number of the defined procedures (13,036 or 6.3\%) are not called in the projects.
+Most procedures are called exactly once (62.32\% of them) or twice (15.65\% ) and from exactly one script (85.92\% of them).
+Examining the origin of procedure calls, we found remarkable that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, thus making it recursive.
+Those recursive procedures are found in 1052 projects.\footnote{list of recursive procedures published \todo{link}}.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/7procedurecalls}
+	\caption{Histogram of the number of calls of each procedure in the dataset}
+	\label{fig:procedurecalls}
+\end{figure}
+
 variables
+lists
 
@@
 logic operations
-lists	
+
 user interativity	

### 2016-04-19 Paper/paper.tex
Results for second research question + additional statistics table

@@
 		Scripts with calls per procedure&1.13&0&1&1&1&59\\
+		Variables per project&2.06&0&0&0&1&340\\
+		Scripts utilizing variable&4.97&1&1&3&5&1127\\
+		Lists per project&0.55&0&0&0&0&319\\
+		Conditional statements per project&10.02&0&0&0&3&5950\\
+		Recursive loop statements per project&7.65&0&1&2&5&2503\\
+		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\
 		\hline
@@
 	\label{tbl-size}
+\end{table*}
+
+\begin{table*}[ht]
+	\centering
+	\begin{tabular}{lrr}
+		\hline
+		&\textbf{number}&\textbf{percentage}\\
+		\hline
+		Retrieved projects & 250,166 & \\
+		Analyzed projects & 247,798 & \\
+		Non-empty projects (used for statistics) & \textbf{233,491} & \\
+
+		Projects utilizing variables & 73,577 & 31.51\% \\
+		Projects utilizing lists & 9,358 & 4.01\% \\
+		Projects with conditional statements & 92,959 & 39.81\% \\
+		Projects with recursive loop statements & 180,210 & 77.18\% \\
+		Projects with \texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
+		Projects with \texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
+		\hline
+	\end{tabular}
+	\caption{Characteristics in the projects in the dataset}
+	\label{tbl-characteristics}
 \end{table*}
@@
 
-variables
-lists
+As shown in table \ref{tbl-characteristics}, almost one-third of the projects utilize variables and a small number (4.01\%) utilize lists.
+The number of variables that is being used is also limited, with only 7.48\% of the projects having 5 or more variables.
+The distribution of variable and list utilization is shown in Figure \ref{fig:variables}.
+Exceptional cases exist: the analysis highlighted 842 projects with more than 100 variables and up to 340.
+Examining the initialization of variables through the \texttt{set <variable> to <value>} blocks, we found that for 4.83\% of all variables this was missing.
+The scope of variables was also analyzed: As shown in Table \ref{tbl-size}, each variable is used by an average of 4.97 different scripts, with a median value of 3.
+However, we must note that this measurement might not be entirely accurate, because variables might be defined in the scope of specific sprites.
+With our dataset we cannot differentiate global from local variables and we would consider local variables defined with the same name in different sprites as global variables.
 
-loops	
-logic operations
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/8variableslists}
+	\caption{Histogram of the number of variables and lists utilized in the projects}
+	\label{fig:variables}
+\end{figure}
 
-user interativity	
-parallelism	- wait blocks, briadcast, receive
+Regarding program control features, conditional statements (blocks \texttt{if <condition> then} and \texttt{if <condition> then else}) are used by 39.81\% of the projects.
+Recursive loops (blocks \texttt{repeat <times>}, \texttt{forever} and \texttt{repeat until <condition>}) are more common, used by 77.18\% of the projects.
+The most common of the three is the \texttt{forever} block, accounting for 51.86\% of all recursive loops and the least common one is the \texttt{repeat until <condition>} block, accounting for 11.57\% and used in 13.59\% of the total projects.
 
-Control abstraction is the abstraction of actions while data abstraction is that of data structures.
+For event management Scratch offers blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}. 
+Those blocks are utilized by 29.57\% of the projects.
+\texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects.
+The events are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received.
+This lack of syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).
 
-Control abstraction involves the use of subprograms and related concepts control flows
-Data abstraction allows handling data bits in meaningful ways. For example, it is the basic motivation behind datatype.
+%user interativity		
+\noindent
+\fbox{
+	\begin{minipage}{0.46\textwidth}
+		\emph{RQ2: One third of the projects utilize variables, sometimes without initializing them.
+39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used.
+29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized.
+Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
+Recursive procedure calls exist.}
+	\end{minipage}}
 

### 2016-04-20 Paper/paper.tex
slightly changed title and fixed use of latex commands (add tilde)

@@
 
-\title{How kids code and how we know it:\\An exploratory study in the Scratch project repository }
+\title{How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository }
 
@@
 
-In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms Scratch projects in our dataset.
+In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms~Scratch projects in our dataset.
 
@@
 The dataset is found to contain a relatively small number of projects without any code: \nemptyPrograms (5.77\%).
-Through random sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default.
+Through random manual sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default.
 Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.
@@
 		Projects with \texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
+		Projects with cloned scripts (>=5 blocks) across sprites & 59,634 & 25.54\% \\
+		Projects with cloned scripts (>=5 blocks) within sprites & 23,671 & 10.14\%\\
+		Projects with cloned procedures (>=5 blocks) & 4,945 & 2.12\%\\
+		Projects with cloned functionality blocks (>=4 blocks) across sprites & 60,554 & 25.93\%\\
+		Projects with exact clones (>=5 blocks) across sprites & 27,574 & 11.81\%\\
+		Projects with exact clones (>=5 blocks) within sprites & 2,043 & 0.87\%\\
 		\hline
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
-	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts analyzed scripts}
+	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts~analyzed scripts}
 	\label{fig:cyclomatic}
@@
 We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project, where each script is a code-block of interconnected Scratch blocks.
-For example, in Figure \ref{fig:ui} the cat sprite contains 2 scripts.
+For example, in Figure \ref{fig:ui} the cat sprite contains two scripts.
 We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines.
@@
 
-The lines of code metric was further analyzed to understand code placement.
+The lines of code metric was further analyzed to understand code organization.
 As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC lines of code---is written within sprites.
@@
 The complexity is higher, over 4, for 3.67\% of the scripts.
-The analysis also highlighted 209 scripts with cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.
+The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.
 Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.
@@
 Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
-Recursive procedure calls exist.}
+Recursive procedure calls exist in 1052 (0.45\% of the total) projects.}
 	\end{minipage}}
 
-\subsection{Code smells}
-dead code
-duplicates/clones -graph, number of overlapping positioning, number of dead code duplicates
+\subsection{Code smells and bad programming practices}
+
+The duplicated code smell is the first smell that we examine.
+The first step for our analysis is to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same program.
+For the identification of clones we do not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal.
+We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}.
+The next step in the analysis is to determine the minimum size of the scripts that are considered clones instead of incidentally similar.
+For this, we examine the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}.
+Based on this distribution, we opt to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/11cloneslines}
+	\caption{Number of cloned scripts of different block sizes across and within sprites}
+	\label{fig:cloneslines}
+\end{figure}
+
+In total, in the dataset we found 170,532 scripts cloned across sprites in 59,634 (25.54\% of the total non-empty) projects.
+726,316 copies of those scripts were found, making each clone being copied an average of 4.26 times.
+Figure \ref{fig:clonesprojects} plots the distribution of clones across projects.
+The majority of projects contain up to two cloned scripts; 7.24\% of the projects contain three or more.
+Figure \ref{fig:clonescopies} plots the number of copies of the identified clones.
+It is of interest that 79,378 (46.55\%) of the identified clones are copied three or more times, and even in 585 cases from 411 projects they are copied more than 50 times and up to 974 \footnote{\todo{link}}.
+Manual inspection of the mostly copied clones revealed that they were scripts used multiple times to cater for different parameter values, sometimes in multiple sprites.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/11clonesprojects}
+	\caption{Histogram of the number of cloned scripts found in the dataset projects}
+	\label{fig:clonesprojects}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/11clonescopies}
+	\caption{Histogram of the number of copies of the identified clones}
+	\label{fig:clonescopies}
+\end{figure}
+
+We further inspected which of the identified clones were duplicated only within the same sprite: 63,682 (37.34\% of the total) clones, in 10.14\% of the projects.
+These are duplication smells that could be avoided by using procedures, which in Scratch can only be defined for specific sprites.
+The lack of support for project-wide procedures could be the cause for procedure clones.
+These were measured to 12,878 (7.55\% of the total) clones, in 2.12\% of the projects.
+
+Exact clones were found in 11.81\% of the total projects.
+Their total number was 66,750 (39.14\% of the total) clones.
+Exact clones in the same sprite are cases of clearly redundant code.
+These were rare, found in only 0.87\% of the projects.
+
+Apart from whole scripts we also examined cases with only the first block being different.
+In first block is commonly the event block and, by separating this one from the subsequent functionality blocks, we could examine if programmers assign the same functionality to handle different events.
+Cloned functionality blocks
+
+dead code -unconnected blocks,
+previous section: A significant number of the defined procedures (13,036 or 6.3\%) are not called in the projects.
+This lack of wait-receive syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).
+
 long method
 large class
-conditional complexity
-Dublicated wait block-same trigger caught by more than 1
 
@@
 code block positioning - overlapping /visibility
-
-\subsubsection{Clones Detection for Scratch programs}
-\label{clone_Detection}
-
-We distinguish three different forms of clones: Exact clones within one sprite, exact clones between sprites of one program, clones of `wait' blocks.
-This section presents the results of our clone detection algorithm on the \nScriptPrograms~ with scripts.
-
-In total \nProgramsWithClones~(26.9\% of programs with scripts) exhibit one of the three types of cloning.The most common type of cloning is copying entire scripts between sprites.
-This occurs in 62.939 programs (25.2\%).
-In 8.220 programs (3.3\%) wait conditions are cloned and in 2.463 files there is a script cloned within a sprite.
-
-\paragraph{Exact clones within one sprite}
-In total, we found \todo{x} cases of exact script clones within one sprite.
-Upon further inspection, we found that some of them consisted of unconnected blocks, as depicted in Figure \ref{fig:Unconnected_clones}.
-While this is technically a clone, as this is \emph{dead code} and will never be executed, this could be considered unharmful sketching.
-Looking only at clones starting with an event block, \todo{...} clones remain over \todo{?} files.
 
@@
 
-\paragraph{Exact clones between sprites of one program}
-
-\paragraph{Clones of `wait' blocks}
-Previous work has established that the cloned `wait' blocks can be harmful when modifying Scratch code, which is the reason why we investigated this particular type of cloning.
-As outlined in section \todo{?} signals can replaced repeated wait blocks, by catching the condition in one place and broadcasting a signal to all other sprites.
-This is why we found it interesting to see in how many of the cases where a wait block was cloned, the creator of the program also used signals in their program.
-This indicates that the concept of signals is known, and hence the duplicate condition is not used because the user is not aware of the alternative.
-In \todo{...} of the cases the program with the duplicated wait block also contained one or more signals.
+Scratch programs are found to be particularly susceptible to the duplicate code smell: code clones were found in one fourth of the projects in the dataset.
 
@@
 With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common.
-In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of \todo{right word?
-procedures} between scripts, only within them.
+In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of procedures between scripts, only within them.
 So in many cases there is no other way to share the functionality than making a copy.

### 2016-04-20 Paper/paper.tex
removed link, it is on my account and reveals me (my account name is felienne
changed example and removed link (reveals my name)

@@
 Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes.
-The Scratch program in Figure \ref{fig:ui}\footnote{\url{https://scratch.mit.edu/projects/97086781/}} consists of one sprite, the cat, which is Scratch's default sprite and logo.
-The code in the sprite will cause the cat to jump up, say ``hello'', and come back down, when the green flag is clicked, and to make the `meow' sound when the space bar is pressed.
+The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a Piano.
+The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
 
@@
 		\includegraphics[width=\columnwidth]{fig/ui.png}
-		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `motion' in the middle and the code associated with the sprite on the right.
+		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `Events' in the middle and the code associated with the sprite on the right.
 The upper right corner shows the actual location of the sprite.}
@@
 
-\subsubsection{Signals}
-\todo{uitleggen wat een signal is en evt ook al hoe het een kloon vervangt?}
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=2cm]{fig/bump.png}
+		\caption{The script for the Piano sprite in Figure \ref{fig:ui}, causing a sound on receiving the `bump' signal.}
+		\label{fig:bump}
+	\end{center}
+\end{figure}
 
@@
 In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam.
-In the Scratch code in Figure \ref{fig:ui} there are two events: `when Green Flag clicked' and `when space key pressed'.
+The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: `when Green Flag clicked' and `when right arrow key pressed', each with associated blocks.
+
+\subsubsection{Signals}
+Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
+All other sprites can then react by using the `when I receive' Hat block.
+In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano.
+The Piano reacts on that by making a sound, see Figure \ref{fig:bump}.
 
 \subsubsection{Scripts}
-Source code within sprites is organized in scripts: a script always starts with an event, followed by a number of blocks.
+Source code within sprites is organized in scripts: a script can start with an event (a Hat block\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}) which indicates when something should happen.
+A hat block can be followed by an arbitrary number of other blocks.
+
 The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar.
@@
 In that case, all scripts will be executed simultaneously.
-For example, the code on the left of Figure \ref{fig:comparison} has five scripts associated with the `when Green Flag clicked' event.
 

### 2016-04-20 Paper/paper.tex
some assorted changes

--- 2016-04-20 Paper/paper.tex
@@
 
+\subsubsection{Procedures}
+Scratch users can define their own blocks, with a pop-up screen displayed in on the left hand side of Figure \ref{fig:twoBlocks}.
+As shown in this Figure, Scratch users can set the name of the sprite and define input parameters.
+The block in Figure \ref{fig:twoBlocks} uses two parameter, one textual (string1) and one numerical (number1).
+A user-created block is programmed using ordinary Scratch blocks as shown on the right hand side of Figure \ref{fig:twoBlocks}.
+
 \subsubsection{Remixing}
+\todo{maybe only useful with we will talk about remixing in the remainder?}
 Scratch programs can be shared by their creators in the global Scratch repository\footnote{\url{https://scratch.mit.edu/explore/projects/all/}}.
@@
 	\begin{center}
-		\includegraphics[width=6cm]{fig/makeBlockSmall.png}
-		\caption{}
-		\label{fig:makeBlock}
-	\end{center}
-\end{figure}
-
-\begin{figure}
-	\begin{center}
-		\includegraphics[width=6cm]{fig/useBlockSmall.png}
-		\caption{}
-		\label{fig:useBlock}
+		\includegraphics[width=8cm]{fig/twoBlocks.png}
+		\caption{The user interface for creating and defining user-defined blocks \todo{could be nicer}}
+		\label{fig:twoBlocks}
 	\end{center}
@@
 The dataset is described in Section \ref{dataset}.
-Below, we present how we approached each research question.
+\todo{Below, we present how we approached each research question.}
 
@@
 	\end{tabular}
-	\caption{Summary statistics from the dataset of \nScriptPrograms  non-empty Scratch projects}
+	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
 	\label{tbl-size}
@@
 The lines of code metric was further analyzed to understand code organization.
-As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC lines of code---is written within sprites.
+As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC~lines of code---is written within sprites.
 An additional 3.1\% of the total lines are found in the stage class.
@@
 	\begin{minipage}{0.46\textwidth}
-		\emph{RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points.
+		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points.
 Most code is written in sprites.
-Few projects (7.7\%) use procedures, but they use them a lot and for more complex code.
-There exist surprisingly large and complex projects.}
+A small number of projects (7.7\%) use procedures, but they use them a lot and for more complex code.
+There exist surprisingly large and complex projects.
 	\end{minipage}}
@@
 
-As a first programming abstraction we investigated the use of procedures.
+The first method for abstraction that we investigate is the use of procedures.
 In the dataset we found 206,799 procedures in 17,979 projects.
@@
 	\begin{minipage}{0.46\textwidth}
-		\emph{RQ2: One third of the projects utilize variables, sometimes without initializing them.
+		RQ2: One third of the projects utilize variables, sometimes without initializing them.
 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used.
@@
 Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
-Recursive procedure calls exist.}
+Recursive procedure calls exist.
 	\end{minipage}}
@@
 With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common.
-In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of procedures between scripts, only within them.
-So in many cases there is no other way to share the functionality than making a copy.
-We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users needs this functionality.
+In a sense, the Scratch users are not to blame here, as Scratch does not support the sharing of procedures between scripts, only within them.
+So in many cases there is no other way to share the functionality than by making a copy.
+We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.
 
@@
 In todo{...} of the cases, the programs with duplicate conditions, also signals were present. This means that it is feas
+
+\subsection{Threads to validity}
+A thread to validity of this study is the fact that we did not scrap a random sample, but the most recent 250,000 programs.
+It could be the case that programming habits of Scratch users are changing over time.
+However, we counterbalanced that by using a very large dataset.
+Furthermore, for some of the analyses we manually inspected programs \todo{Fenia, which ones?}
 

### 2016-04-20 Paper/paper.tex
some assorted changes

--- 2016-04-20 Paper/paper.tex
@@
 	\begin{minipage}{0.46\textwidth}
-		\emph{RQ2: One third of the projects utilize variables, sometimes without initializing them.
+RQ2: One third of the projects utilize variables, sometimes without initializing them.
 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used.

### 2016-04-20 Paper/paper.tex
added repo link

--- 2016-04-20 Paper/paper.tex
@@
 
-Kragle, as well as all scraped programs and our analysis files are available via \todo{url} \todo{I do not think we need to describe the fields in the paper, but we need a good readme with the data (I can do that)}
+Kragle, as well as all scraped programs and our analysis files are available\footnote{\url{https://github.com/ScratchLover42/ICER-Data-Code}}.
 

### 2016-04-20 Paper/paper.tex
added repo link

--- 2016-04-20 Paper/paper.tex
@@
 Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
-Recursive procedure calls exist in 1052 (0.45\% of the total) projects.}
-	\end{minipage}}
+Recursive procedure calls exist in 1052 (0.45\% of the total) projects.
+	\end{minipage}
+	}
 

### 2016-04-20 Paper/paper.tex
added reference to example for indentation

--- 2016-04-20 Paper/paper.tex
@@
 We ran Kragle on March 2nd 2016 and let it running for 24 hours, when it had obtained a little over 250.000 programs.
-Out of the \nPrograms~, we failed to analyze 2.367 programs due to various technical difficulties \todo{add details, I will check a few to see what went wrong - Felienne}
-
+Out of the \nPrograms~, we failed to parse and further analyze 2.367 programs due technical difficulties with the provided JSON files.
 Kragle, as well as all scraped programs and our analysis files are available\footnote{\url{https://github.com/ScratchLover42/ICER-Data-Code}}.
@@
 We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines.
-The maximum code indentation level is used as a code complexity metric.
-It was calculated using the JSON array depth?
-\todo{felienne, how did we measure this?}.
-This corresponds to code indentation level returning 1 for all first-level blocks and increasing by 2 for block enclosures in loops or conditional statements.
+Furthermore, we measure the nestedness of program, which is the number of times blocks are placed within each other.
+For example, the program in Figure \ref{fig:ui} has a maximum indentation of 3 in the `Wgen Green flag clicked' script, as that has a broadcast block within an if, within a forever\todo{for Felienne: double check}.
 Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.

### 2016-04-20 Paper/paper.tex
added remark on init of vars
First version of research question 3 answers

--- 2016-04-20 Paper/paper.tex
@@
 
-\title{How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository }
+\title{How kids code and how we know it:\\An exploratory study in the Scratch project repository }
 
@@
 Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes.
-The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a Piano.
-The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
+The Scratch program in Figure \ref{fig:ui}\footnote{\url{https://scratch.mit.edu/projects/97086781/}} consists of one sprite, the cat, which is Scratch's default sprite and logo.
+The code in the sprite will cause the cat to jump up, say ``hello'', and come back down, when the green flag is clicked, and to make the `meow' sound when the space bar is pressed.
 
@@
 		\includegraphics[width=\columnwidth]{fig/ui.png}
-		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `Events' in the middle and the code associated with the sprite on the right.
+		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `motion' in the middle and the code associated with the sprite on the right.
 The upper right corner shows the actual location of the sprite.}
@@
 
-\begin{figure}
-	\begin{center}
-		\includegraphics[width=2cm]{fig/bump.png}
-		\caption{The script for the Piano sprite in Figure \ref{fig:ui}, causing a sound on receiving the `bump' signal.}
-		\label{fig:bump}
-	\end{center}
-\end{figure}
+\subsubsection{Signals}
+\todo{uitleggen wat een signal is en evt ook al hoe het een kloon vervangt?}
 
@@
 In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam.
-The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: `when Green Flag clicked' and `when right arrow key pressed', each with associated blocks.
-
-\subsubsection{Signals}
-Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
-All other sprites can then react by using the `when I receive' Hat block.
-In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano.
-The Piano reacts on that by making a sound, see Figure \ref{fig:bump}.
+In the Scratch code in Figure \ref{fig:ui} there are two events: `when Green Flag clicked' and `when space key pressed'.
 
 \subsubsection{Scripts}
-Source code within sprites is organized in scripts: a script can start with an event (a Hat block\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}) which indicates when something should happen.
-A hat block can be followed by an arbitrary number of other blocks.
-
+Source code within sprites is organized in scripts: a script always starts with an event, followed by a number of blocks.
 The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar.
@@
 In that case, all scripts will be executed simultaneously.
-
-\subsubsection{Procedures}
-Scratch users can define their own blocks, with a pop-up screen displayed in on the left hand side of Figure \ref{fig:twoBlocks}.
-As shown in this Figure, Scratch users can set the name of the sprite and define input parameters.
-The block in Figure \ref{fig:twoBlocks} uses two parameter, one textual (string1) and one numerical (number1).
-A user-created block is programmed using ordinary Scratch blocks as shown on the right hand side of Figure \ref{fig:twoBlocks}.
+For example, the code on the left of Figure \ref{fig:comparison} has five scripts associated with the `when Green Flag clicked' event.
 
 \subsubsection{Remixing}
-\todo{maybe only useful with we will talk about remixing in the remainder?}
 Scratch programs can be shared by their creators in the global Scratch repository\footnote{\url{https://scratch.mit.edu/explore/projects/all/}}.
@@
 	\begin{center}
-		\includegraphics[width=8cm]{fig/twoBlocks.png}
-		\caption{The user interface for creating and defining user-defined blocks \todo{could be nicer}}
-		\label{fig:twoBlocks}
+		\includegraphics[width=6cm]{fig/makeBlockSmall.png}
+		\caption{}
+		\label{fig:makeBlock}
+	\end{center}
+\end{figure}
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=6cm]{fig/useBlockSmall.png}
+		\caption{}
+		\label{fig:useBlock}
 	\end{center}
@@
 The dataset is described in Section \ref{dataset}.
-\todo{Below, we present how we approached each research question.}
+Below, we present how we approached each research question.
 
@@
 
-Kragle, as well as all scraped programs and our analysis files are available\footnote{\url{https://github.com/ScratchLover42/ICER-Data-Code}}.
+Kragle, as well as all scraped programs and our analysis files are available via \todo{url} \todo{I do not think we need to describe the fields in the paper, but we need a good readme with the data (I can do that)}
 
@@
 
-In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms~Scratch projects in our dataset.
+In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms Scratch projects in our dataset.
 
@@
 The dataset is found to contain a relatively small number of projects without any code: \nemptyPrograms (5.77\%).
-Through random manual sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default.
+Through random sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default.
 Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.
@@
 	\end{tabular}
-	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
+	\caption{Summary statistics from the dataset of \nScriptPrograms  non-empty Scratch projects}
 	\label{tbl-size}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
-	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts~analyzed scripts}
+	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts analyzed scripts}
 	\label{fig:cyclomatic}
@@
 We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project, where each script is a code-block of interconnected Scratch blocks.
-For example, in Figure \ref{fig:ui} the cat sprite contains two scripts.
+For example, in Figure \ref{fig:ui} the cat sprite contains 2 scripts.
 We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines.
@@
 
-The lines of code metric was further analyzed to understand code organization.
-As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC~lines of code---is written within sprites.
+The lines of code metric was further analyzed to understand code placement.
+As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC lines of code---is written within sprites.
 An additional 3.1\% of the total lines are found in the stage class.
@@
 The complexity is higher, over 4, for 3.67\% of the scripts.
-The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.
+The analysis also highlighted 209 scripts with cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.
 Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.
@@
 	\begin{minipage}{0.46\textwidth}
-		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points.
+		\emph{RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points.
 Most code is written in sprites.
-A small number of projects (7.7\%) use procedures, but they use them a lot and for more complex code.
-There exist surprisingly large and complex projects.
+Few projects (7.7\%) use procedures, but they use them a lot and for more complex code.
+There exist surprisingly large and complex projects.}
 	\end{minipage}}
@@
 
-The first method for abstraction that we investigate is the use of procedures.
+As a first programming abstraction we investigated the use of procedures.
 In the dataset we found 206,799 procedures in 17,979 projects.
@@
 	\begin{minipage}{0.46\textwidth}
-RQ2: One third of the projects utilize variables, sometimes without initializing them.
+		\emph{RQ2: One third of the projects utilize variables, sometimes without initializing them.
 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used.
@@
 Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
-Recursive procedure calls exist in 1052 (0.45\% of the total) projects.
-	\end{minipage}
-	}
+Recursive procedure calls exist in 1052 (0.45\% of the total) projects.}
+	\end{minipage}}
 
@@
 With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common.
-In a sense, the Scratch users are not to blame here, as Scratch does not support the sharing of procedures between scripts, only within them.
-So in many cases there is no other way to share the functionality than by making a copy.
-We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.
+In a sense, the Scratch users are not to blame here, Scratch does not support the sharing of \todo{right word?
+procedures} between scripts, only within them.
+So in many cases there is no other way to share the functionality than making a copy.
+We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users needs this functionality.
 
@@
 In todo{...} of the cases, the programs with duplicate conditions, also signals were present. This means that it is feas
-
-\subsection{Threads to validity}
-A thread to validity of this study is the fact that we did not scrap a random sample, but the most recent 250,000 programs.
-It could be the case that programming habits of Scratch users are changing over time.
-However, we counterbalanced that by using a very large dataset.
-Furthermore, for some of the analyses we manually inspected programs \todo{Fenia, which ones?}
 

### 2016-04-20 Paper/paper.tex
finished RQ3

@@
 		Projects with exact clones (>=5 blocks) within sprites & 2,043 & 0.87\%\\
+		Projects containing dead code & 56,890	& 24.36\%\\
+		Projects containing large scripts & 69,521 & 29.77\%\\
+		Projects containing large sprites & 31,954 & 13.68\%\\		
 		\hline
@@
 The use of procedures in projects was further investigated through the procedure calls, summarized in Figure \ref{fig:procedurecalls}.
-A significant number of the defined procedures (13,036 or 6.3\%) are not called in the projects.
-Most procedures are called exactly once (62.32\% of them) or twice (15.65\% ) and from exactly one script (85.92\% of them).
+A significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects.
+Most procedures are called exactly once (62.32\% of them) or twice (14.30\%) and from exactly one script (85.92\% of them).
 Examining the origin of procedure calls, we found remarkable that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, thus making it recursive.
-Those recursive procedures are found in 1052 projects.\footnote{list of recursive procedures published \todo{link}}.
+Those recursive procedures are found in 1,052 projects.\footnote{list of recursive procedures published \todo{link}}.
 
@@
 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used.
-29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized.
+29.57\% of the projects utilize broadcast and receive blocks, but in 7.99\% of the projects they are not synchronized.
 Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
@@
 Apart from whole scripts we also examined cases with only the first block being different.
-In first block is commonly the event block and, by separating this one from the subsequent functionality blocks, we could examine if programmers assign the same functionality to handle different events.
-Cloned functionality blocks
+In first block is commonly the event block and, by separating this one from the subsequent functionality blocks, we could examine if programmers assign the same functionality to handle different types of events.
+Cloned functionality blocks are found to be rare: without considering the first block, only 2,243 additional clones were found in 920 projects.
 
-dead code -unconnected blocks,
-previous section: A significant number of the defined procedures (13,036 or 6.3\%) are not called in the projects.
-This lack of wait-receive syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).
+The second smell that we examine is the dead code smell.
+Already in the previous section, Scratch projects were found to have two types of dead code: defined procedures that are not called (5.06\% of the defined procedures in 2,079 projects) and unmatched broadcast-receive messages (in 7.99\% of the projects).
+In this section we examine an additional case: scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality.
+A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects.
+The majority of those scripts (86.6\%) are scripts missing the starting block.
+Examining the size of those dead scripts, 72.34\% are composed of a single block.
+As shown in diagram \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of those scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610\footnote{\todo{link}}.
 
-long method
-large class
-
-coding style -naming -functions, sprites, variables
-code block positioning - overlapping /visibility
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/12deadcode}
+	\caption{Size of scripts that are identified as dead code}
+	\label{fig:deadcode}
+\end{figure}
 
@@
 
-Scratch programs are found to be particularly susceptible to the duplicate code smell: code clones were found in one fourth of the projects in the dataset.
+Finally, we examine the long method and the large class smell, considering them in the context of Scratch as large script and large sprite smells respectively.
+For those two smells we use the number of blocks as the size metric \todo{add this to threats to validity?
+many blocks do not increase the length of scripts in Scratch}.
+Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset.
+We used those numbers to split the dataset and retrieve the top 10\% \todo{felienne, can we justify why 10 percent?} largest scripts and sprites, and in this way we found the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
+Using these thresholds, the number of projects exhibiting the large script smell, i.e., containing at least one script with 18 or more blocks, is 69,521 (29.77\% of the total projects) and the number of projects with the large sprite smell is 31,954 (13.68\%).
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/13longmethod}
+	\caption{Size of sprites and scripts in number of blocks}
+	\label{fig:longmethod}
+\end{figure}
+
+\noindent
+\fbox{
+	\begin{minipage}{0.46\textwidth}
+		\emph{RQ3: Code clones are found in 25.54\% of the projects, with 46.55\% of the clones copied three or more times, in the same or across sprites.
+24.36\% of the projects contain scripts that are never invoked, and thus exhibit the dead code smell.
+In some cases those scripts are large.
+The large script smell is found in 29.77\% of the projects and the large sprite in 13.68\%.}
+	\end{minipage}}
 

### 2016-04-21 Paper/paper.tex

--- 2016-04-20 Paper/paper.tex
@@
 Examining the initialization of variables through the \texttt{set <variable> to <value>} blocks, we found that for 4.83\% of all variables this was missing.
+While failing to initialize a variable in Scratch will not result in a runtime error as in some other programming languages, it is considered bad style \todo{citation needed}.
+
 The scope of variables was also analyzed: As shown in Table \ref{tbl-size}, each variable is used by an average of 4.97 different scripts, with a median value of 3.

### 2016-04-21 Paper/paper.tex
first stab at a new version of the discussion and some changes in the intro

@@
 Our study is based on data from the Scratch program repository.
-By scraping the list of recent programs \todo{link}, we have obtained \nPrograms~public Scratch programs and performed static analysis on them.
+By scraping the list of recent programs \footnote{\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch programs and performed analysis on them.
 
@@
 	\item{A public data set of \nScriptPrograms~Scratch programs (Section \ref{dataset})}
-	\item{An empirical evaluation of the data set for answering the aforementioned research questions}
+	\item{An evaluation of the data set in terms of (Section \ref{sec:results})}
+	\item{A discussion of implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
 \end{itemize}
@@
 \section{Results}
-
+\label{sec:results}
 In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms~Scratch projects in our dataset.
@@
 Finally, we examine the long method and the large class smell, considering them in the context of Scratch as large script and large sprite smells respectively.
-For those two smells we use the number of blocks as the size metric \todo{add this to threats to validity?
-many blocks do not increase the length of scripts in Scratch}.
+For those two smells we use the number of blocks as the size metric.
 Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset.
@@
 	\begin{minipage}{0.46\textwidth}
-		\emph{RQ3: Code clones are found in 25.54\% of the projects, with 46.55\% of the clones copied three or more times, in the same or across sprites.
+RQ3: Code clones are found in 25.54\% of the projects, with 46.55\% of the clones copied three or more times, in the same or across sprites.
 24.36\% of the projects contain scripts that are never invoked, and thus exhibit the dead code smell.
 In some cases those scripts are large.
-The large script smell is found in 29.77\% of the projects and the large sprite in 13.68\%.}
+The large script smell is found in 29.77\% of the projects and the large sprite in 13.68\%.
 	\end{minipage}}
@@
 \section{Discussion}
-
+\label{sec:discussion}
 \subsection{Implications}
@@
 
-\subsubsection{Unconnected Blocks}
-In the whole dataset, not just in the programs with clones, we observe unconnected scripts, scripts without an event block and events without associated code.
-This is a form of \emph{dead code}, code that will never be executed.
-Because this dead code caused visual clutter, it would be better to have a separate workspace (much like the `backpack in Scratch' where users can store blocks temporarily and use them later.
-A programming interface could ask actively encourage users to move unconnected block when they exit the environment to encourage a `clean' working space.
+\subsubsection{Dead Code}
+In our analysis, we find that almost one quarter of the Scratch programs contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality.
+\todo{Fenia, this is correct, right, the 23\% is just these types and not also uncalled procs and unmatched broadcasts?}.
 
-\subsubsection{Code Clones Between Sprites}
-With occurrences in one quarter of the 250 thousand programs, the use of exactly identical clones between sprites is extremely common.
-In a sense, the Scratch users are not to blame here, as Scratch does not support the sharing of procedures between scripts, only within them.
+In a sense, the dead scripts are harmless, as they are not executed, however, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead.
+In contrast with other visual educational languages, Scratch does on indicate scripts are dead.
+LEGO Mindstorms, for example, does give the user feedback by making unconnected blocks gray.
+
+Looking at the number of unconnected blocks, we hypothesize that Scratch programmers have a need for a separate workspace to store unconnected blocks temporarily.
+We envision that would be like the `backpack' meant to move scripts across sprites.
+In order to help novice programmers to keep their code clean, the programming interface could  actively encourage users to move unconnected blocks to that workspace when they exit the environment.
+
+\subsubsection{Exact Clones between Sprites}
+With occurrences in 11\% of the Scratch programs in our dataset, the use of exactly identical clones between sprites is relatively common.
+In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them.
 So in many cases there is no other way to share the functionality than by making a copy.
@@
 
-\subsubsection{Refactoring support for duplicate wait blocks}
-In todo{...} of the cases, the programs with duplicate conditions, also signals were present. This means that it is feas
+\subsubsection{Sharing of Scripts and Procedures}
+Investigating the use of clones between programs, we observe that there are 1700 scripts that are used in multiple programs, sometimes as often as in 1600 different programs.
+This seems to indicate that there are common patterns in Scratch programs.
+This again leads us to the idea that it might be very beneficial to Scratch programmers if they could not only share their programs, but also share some of their functionality, for others to use, like a library.
+An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'.
+This might empower new Scratch users to get started quicker.
 
@@
 Furthermore, for some of the analyses we manually inspected programs \todo{Fenia, which ones?}
+
+Furthermore we use the number of blocks in the Scratch programs as a measure for the length of a program, while this does not exactly correspond to the `length' of a program in lines, and there can be multiple Scratch blocks on one line.
+For example, in Figure \ref{fig:ui} the `if' and `touching' blocks are present on the same `line'.
+We however believe that the number of blocks is a good proxy for size, and we plan a future experiment in which we will compare `lines of Scratch code' to `number of blocks'.
 

### 2016-04-21 Paper/paper.tex
corrected number
input on User interactivity in RQ2

@@
 		Recursive loop statements per project&7.65&0&1&2&5&2503\\
+		User input blocks per project&4.77&0&0&1&4&1889\\
 		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\	
@@
 		Projects with conditional statements & 92,959 & 39.81\% \\
+		Projects accepting user input & 131,314 & 56.24\% \\
 		Projects with recursive loop statements & 180,210 & 77.18\% \\
@@
 
+Investigating user interactivity functionality, we found that 56.24\% of the projects in the dataset contain user input blocks---an average of 8.48 blocks per such project.
+Table \ref{tbl-userinput} lists the frequency of use of user input controls.
+The most commonly used one is the \texttt{when key pressed}, found in 71,096 (30.45\% of the total) projects.
+The most frequently used parameter for the \texttt{key} attribute is the space key, followed by the arrows and then the letters and numbers.
+
+\begin{table}
+	\centering
+	\begin{tabular}{lrr}
+		\hline
+		\textbf{Block}&\textbf{Projects}&\textbf{Occurrences}\\
+		\hline
+	\texttt{when <> key pressed}&71,096&294,771\\
+	\texttt{when this sprite clicked}&39,179&198,342\\
+	(Sensing) \texttt{key <> pressed?}&37,919&291,657\\
+	(Sensing) \texttt{ask <> and wait}&19,039&66,850\\
+	(Sensing) \texttt{mouse down?}&9,115&54,079\\
+	(Sensing) \texttt{<attrib> of <>}&9,068&155,468\\
+	(Sensing) \texttt{mouse X}&5,977&27,321\\
+	(Sensing) \texttt{mouse Y}&3,940&22,035\\
+	\texttt{when <sensor> > <value>}&705&1,570\\
+	(Sensing) \texttt{video <> on <>}&434&1,397\\
+		\hline
+	\end{tabular}
+	\caption{Frequency of use of user input blocks in the \nScriptPrograms projects of the dataset}
+	\label{tbl-userinput}
+\end{table}
+
 For event management Scratch offers blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}. 
@@
 39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used.
-29.57\% of the projects utilize broadcast and receive blocks, but in 29.57\% of the projects they are not synchronized.
+More than half of the projects are interactive.
+29.57\% of the projects utilize broadcast and receive blocks, but in 7.99\% of the projects they are not synchronized.
 Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
@@
 
-\subsection{Code Smells}
+The evaluation of block-based languages in general, and Scratch in particular, as tools for programming education has received significant research attention during the past years.
+A number of studies have been carried out on the programming practices of novice programmers in block-based environments, on the programming skills they develop, and on the quality of Scratch programs.
+
+Finally, there are other works on the quality of Scratch programs.
+There is for example the Hairball Scratch extension~\cite{boe_hairball:_2013}, which is a lint-like static analysis tool for Scratch that can detect, for example, unmatched broadcast and receive blocks, infinite loops and duplication.
+An evaluation of 100 Scratch programs showed that Scratch programs indeed suffer from duplication and bad naming~\cite{moreno_automatic_2014}.
+
+Most related to our study is the work by Moreno \emph{et al.}~\cite{moreno-leon_dr._2015} who gave automated feedback on Scratch programs to 100 children aged 10 to 14.
+Their results demonstrated that feedback on code quality helped improve students' programming skills.
+
 Efforts related to our research include works on code smells, initiated by the work by Fowler~\cite{fowler_refactoring:_1999}.
@@
 
-\subsection{Smells beyond the OO paradigm}
-In recent times, code smells have been applied to programs outside of the regular programming domain.
-In our past work, we have, for example, studied code smells within spreadsheets, both at the formula level~\cite{hermans_detecting_2014} and between worksheets~\cite{hermans_detecting_2012}.
-
-More recently, we compared two datasets: one containing spreadsheets which users found unmaintainable, and a version of the same spreadsheets rebuilt by professional spreadsheet developers.
-The results show that the improved versions suffered from smells to a lesser degree, increasing our confidence that presence of smells indeed coincides with users finding spreadsheets hard to maintain~\cite{jansen_code_2015}.
-
-In addition to spreadsheets, code smells have also been studied in the context of Yahoo!
-Pipes a web mashup tool.
-An experiment demonstrated that users preferred the non-smelly versions of Yahoo Pipes programs~\cite{stolee_refactoring_2011}.
-
-%\subsection{Agile practices in education}
-%A third related line of work are research directions where other methods from agile development, like testing and refactoring, have been applied in education.
-For example, the use of test-driven development has been studied within the context of programming education at university level, using text-based languages in several studies.
-Most controlled experiments resulted in increased code quality for the group of students using TDD, for example [13] reported on a 35\% reduction in defects, while [14] found 45\%.
-For an extensive overview of TDD approaches in university programming courses, see [15].
-\todo{referenties fixen}
-
-\subsection{Quality of Scratch programs}
-Finally, there are other works on the quality of Scratch programs.
-There is for example the Hairball Scratch extension~\cite{boe_hairball:_2013}, which is a lint-like static analysis tool for Scratch that can detect, for example, unmatched broadcast and receive blocks, infinite loops and duplication.
-An evaluation of 100 Scratch programs showed that Scratch programs indeed suffer from duplication and bad naming~\cite{moreno_automatic_2014}.
-
-Most related to our study is the work by Moreno \emph{et al.}~\cite{moreno-leon_dr._2015} who gave automated feedback on Scratch programs to 100 children aged 10 to 14.
-Their results demonstrated that feedback on code quality helped improve students' programming skills.
-
 \bibliographystyle{abbrv}

### 2016-04-21 Paper/paper.tex
Related work

@@
 The evaluation of block-based languages in general, and Scratch in particular, as tools for programming education has received significant research attention during the past years.
-A number of studies have been carried out on the programming practices of novice programmers in block-based environments, on the programming skills they develop, and on the quality of Scratch programs.
+A number of studies have been carried out on the understanding of programming concepts and the programming practices of novice programmers in block-based environments, on the programming skills they develop, and on the quality of Scratch programs.
 
-Finally, there are other works on the quality of Scratch programs.
-There is for example the Hairball Scratch extension~\cite{boe_hairball:_2013}, which is a lint-like static analysis tool for Scratch that can detect, for example, unmatched broadcast and receive blocks, infinite loops and duplication.
-An evaluation of 100 Scratch programs showed that Scratch programs indeed suffer from duplication and bad naming~\cite{moreno_automatic_2014}.
+A study on the internalization of programming concepts with Scratch with 46 students was presented in \cite{meerbaum-salant_learning_2010}.
+Concepts like loops, conditional loops, message passing, initialization, variables and concurency were examined, and it was found that students had problems with the last three.
+In a later study with an equal set of subjects \cite{Meerbaum_habits_2011} the same authors identified two bad programming habits in Scratch, namely bottom-up development and extremely fine-grained programming.
+They connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops, a finding that is verified by our study.
+In \cite{wilson_evaluation_2012} 29 Scratch projects created from 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
 
-Most related to our study is the work by Moreno \emph{et al.}~\cite{moreno-leon_dr._2015} who gave automated feedback on Scratch programs to 100 children aged 10 to 14.
-Their results demonstrated that feedback on code quality helped improve students' programming skills.
+Most related to our study for the second research question of programming abstractions and concepts is the work by Maloney \emph{et al.} \cite{Maloney_2008}, who analyzed 536 Scratch projects for blocks that relate to programming concepts including loops, conditional statements, variables, user interaction, synchronization, and random numbers.
+Compared to their findings, our investigation reveals increased use of the first three concepts, and especially variables.
 
-Efforts related to our research include works on code smells, initiated by the work by Fowler~\cite{fowler_refactoring:_1999}.
+The Scratch automated quality analysis tools Hairball \cite{boe_hairball:_2013} and Dr.
+Scratch \cite{moreno-leon_dr._2015} are also related to our work on smell detection.
+The Hairball Scratch extension is a lint-like static analysis tool for Scratch that can detect initialization problems and unmatched broadcast and receive blocks.
+In their work \cite{moreno_automatic_2014}, Moreno and Robles extended Hairball to detect two bad programming habits in Scratch: not changing the default object names and duplicating scripts, and apply them for evaluating 100 projects from the Scratch repository.
+The results on script duplication are substantially different than ours---we find projects with script clones to appear half as frequently.
+The Dr.
+Scratch tool \cite{moreno-leon_dr._2015} includes bad naming, code duplication and dead code identification functionality, and also evaluates Scratch projects in terms of abstraction, parallelism, logical thinking, synchronization, flow control, user interactivity and data representation.
+
+Also related to our research are works on code smells, initiated by the work by Fowler~\cite{fowler_refactoring:_1999}.
 His book gives an overview of code smells and corresponding refactorings.
@@
 Marinescu~\cite{marinescu_detecting_2001} for instance, uses metrics to identify \emph{suspect} classes: classes which could have design flaws.
-Lanza and Marinescu~\cite{lanza_object-oriented_2006} explain this methodology in more detail.
-Alves \emph{et al.}~\cite{alves_deriving_2010} focus on a strategy to obtain thresholds for metrics from a benchmark.
-Olbrich \emph{et al.} furthermore investigates the changes in smells over time, and discusses their impact~\cite{olbrich_evolution_2009}.
-Moha \emph{et al.}~\cite{moha_decor:_2010} designed the `DECOR' method which automatically generates a smell detection algorithms from specifications.
-The CCFinder tool~\cite{kamiya_ccfinder:_2002} finally, aims at detecting clones in source code, which are similar to our \dup smell.
+Alves \emph{et al.}~\cite{alves_deriving_2010} focus on a strategy to obtain thresholds for metrics from a benchmark Moha \emph{et al.}~\cite{moha_decor:_2010} designed the `DECOR' method which automatically generates a smell detection algorithms from specifications.
+The CCFinder tool~\cite{kamiya_ccfinder:_2002} finally, aims at detecting clones in source code, which are similar to our code duplication smell.
 

### 2016-04-21 Paper/paper.tex
Updated table 2

@@
 		\hline
-		&\textbf{number}&\textbf{percentage}\\
+		&\textbf{Number of}&\\
+		&\textbf{projects}&\textbf{Percentage}\\
 		\hline
-		Retrieved projects & 250,166 & \\
-		Analyzed projects & 247,798 & \\
-		Non-empty projects (used for statistics) & \textbf{233,491} & \\
-
-		Projects utilizing variables & 73,577 & 31.51\% \\
-		Projects utilizing lists & 9,358 & 4.01\% \\
-		Projects with conditional statements & 92,959 & 39.81\% \\
-		Projects accepting user input & 131,314 & 56.24\% \\
-		Projects with recursive loop statements & 180,210 & 77.18\% \\
-		Projects with \texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
-		Projects with \texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
-		Projects with cloned scripts (>=5 blocks) across sprites & 59,634 & 25.54\% \\
-		Projects with cloned scripts (>=5 blocks) within sprites & 23,671 & 10.14\%\\
-		Projects with cloned procedures (>=5 blocks) & 4,945 & 2.12\%\\
-		Projects with cloned functionality blocks (>=4 blocks) across sprites & 60,554 & 25.93\%\\
-		Projects with exact clones (>=5 blocks) across sprites & 27,574 & 11.81\%\\
-		Projects with exact clones (>=5 blocks) within sprites & 2,043 & 0.87\%\\
-		Projects containing dead code & 56,890	& 24.36\%\\
-		Projects containing large scripts & 69,521 & 29.77\%\\
-		Projects containing large sprites & 31,954 & 13.68\%\\		
+		Retrieved & 250,166 & \\
+		Analyzed & 247,798 & \\
+		Non-empty (used for statistics) & \textbf{233,491} & \\
+		\hline
+		Procedures & 17,979 & 7.70\%\\
+		Variables & 73,577 & 31.51\% \\
+		Lists & 9,358 & 4.01\% \\
+		Conditional statements & 92,959 & 39.81\% \\
+		User input blocks& 131,314 & 56.24\% \\
+		Loop statements & 180,210 & 77.18\% \\
+		\texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
+		\texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
+		Cloned scripts across sprites & 59,634 & 25.54\% \\
+		Cloned scripts within sprites & 23,671 & 10.14\%\\
+		Cloned procedures & 4,945 & 2.12\%\\
+		Cloned functionality blocks across sprites & 60,554 & 25.93\%\\
+		Exact clones across sprites & 27,574 & 11.81\%\\
+		Exact clones within sprites & 2,043 & 0.87\%\\
+		Dead code & 56,890	& 24.36\%\\
+		Large scripts & 69,521 & 29.77\%\\
+		Large sprites & 31,954 & 13.68\%\\		
 		\hline

### 2016-04-22 Paper/paper.tex
intro

--- 2016-04-21 Paper/paper.tex
@@
 Scratch is a \emph{block-based} language: users manipulate blocks to program.
-Block-based languages are visual languages, but also use some successful aspects of text-based languages such as limited text-entry and indentation, and as such are closer to `real', textual programming than other forms of visual programming, like dataflow languages are.
 
@@
 
-refer to:
-Research on the effectiveness of those languages for programming education
-Research on programming habits
-Our research on smells
+Several studies have shown that block-based languages are powerful as a tool for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003}.
+\todo{Fenia, do you have some work on Research on programming habits}
+
+Recent work has focused on \emph{code smells} within Scratch programs, parts of a program that are not designed in the best possible way.
+\todo{add dr.
+scratch papers here} Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \todo{cite}.
 

### 2016-04-22 Paper/paper.tex
changes upto 4.2

@@
 There is substantial research showing that those visual programming environments are suitable for teaching programming concepts.
+But, what do people do when they use Scratch?
 In this paper we explore the characteristics of Scratch programs.
 To that end we have scraped the Scratch public repository and retrieved 250.000 projects.
-We present an analysis of those projects in terms of complexity, used abstractions and programming concepts, and coding style.
-We found that programming abstraction concepts like functions are not commonly used.
+We present an analysis of those projects in three different dimensions.
+Firstly, we look at the types of blocks used and the size of the programs.
+We then investigate complexity, used abstractions and programming concepts.
+Finally we look at coding style and detect \emph{code smells and clones}.
+We find that programming abstraction concepts like functions are not commonly used.
 We further investigate the presence of code smells and bad programming practices, including code duplication, dead code, long method and large class smells.
@@
 \label{sec:background}
-Block-based languages go back to 1986, when Glinert introduced the BLOX language~\cite{e._glinert_towards_1986}.
-BLOX consists of puzzle-like programming statements that can be combined into programs by combining them both vertically and horizontally.
-After a decade of little activity into block-based languages, they became a research topic again, starting with Alice~\cite{conway_alice:_1994}.
-More recently, new block-based languages have gained widespread popularity, especially powered by Scratch~\cite{resnick_scratch:_2009} and Blockly\footnote{\url{https://developers.google.com/blockly/}}.
+Recently new block-based languages have gained widespread popularity, especially powered by Scratch~\cite{resnick_scratch:_2009} and Blockly\footnote{\url{https://developers.google.com/blockly/}}.
 Over 100 million students have tried Blockly via Code.org, and the Scratch repository currently hosts over 12 million projects.
-Unlike in BLOX, in these new block-based languages the programming blocks can only be combined vertically, resembling textual code more.
 
@@
 Their experiment showed that students in the block-based interface were more focused and completed more of the activity's goals in less time.
-
-Summarizing the above, we conclude that block-based languages have a clear potential to be a great tool for introductory programming education, in some cases even outperforming text-based languages.
 
@@
 
-\subsubsection{Events}
-Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events.
-The canonical event is the `when Green Flag clicked', activated by clicking the green flag at the top of the user interface.
-In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam.
-The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: `when Green Flag clicked' and `when right arrow key pressed', each with associated blocks.
-
-\subsubsection{Signals}
-Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
-All other sprites can then react by using the `when I receive' Hat block.
-In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano.
-The Piano reacts on that by making a sound, see Figure \ref{fig:bump}.
-
 \subsubsection{Scripts}
-Source code within sprites is organized in scripts: a script can start with an event (a Hat block\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}) which indicates when something should happen.
-A hat block can be followed by an arbitrary number of other blocks.
-
+Sprites can have multiple code blocks, called `scripts'.
 The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar.
@@
 
+\subsubsection{Events}
+Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}).
+The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface.
+In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam.
+The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when right arrow key pressed}, each with associated blocks.
+
+\subsubsection{Signals}
+Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
+All other sprites can then react by using the \texttt{when I receive} Hat block.
+In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano.
+The Piano reacts on that by making a sound, see Figure \ref{fig:bump}.
+
 \subsubsection{Procedures}
 Scratch users can define their own blocks, with a pop-up screen displayed in on the left hand side of Figure \ref{fig:twoBlocks}.
-As shown in this Figure, Scratch users can set the name of the sprite and define input parameters.
+As shown in this figure, Scratch users can enter the name of the sprite and define input parameters.
 The block in Figure \ref{fig:twoBlocks} uses two parameter, one textual (string1) and one numerical (number1).
 A user-created block is programmed using ordinary Scratch blocks as shown on the right hand side of Figure \ref{fig:twoBlocks}.
-
-\subsubsection{Remixing}
-\todo{maybe only useful with we will talk about remixing in the remainder?}
-Scratch programs can be shared by their creators in the global Scratch repository\footnote{\url{https://scratch.mit.edu/explore/projects/all/}}.
-Shared Scratch programs can be `remixed' by other Scratch users, which means that a copy of this program is placed in the user's own project collection, and can be then further changed.
-The `remix tree' of projects is public, so users can track which users remix their programs, a bit similar forking in GitHub.
-Contrary to forking though, changes upstream cannot be integrated back into the original project.
 
@@
 		\includegraphics[width=8cm]{fig/twoBlocks.png}
-		\caption{The user interface for creating and defining user-defined blocks \todo{could be nicer}}
+		\caption{The user interface for creating (left) and defining (right) user-defined blocks.}
 		\label{fig:twoBlocks}
@@
 The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch programs.
-To answer our research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch repository.
+To answer our three research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch repository.
 The dataset is described in Section \ref{dataset}.
@@
 Subsequently, Kragle obtains the JSON code for each of the listed programs.
-In addition to the programs themselves, Kragle also gathers metadata including the numbers of views, loves, favorites and remixes.
 
@@
 
-We ran Kragle on March 2nd 2016 and let it running for 24 hours, when it had obtained a little over 250.000 programs.
+We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 programs.
 Out of the \nPrograms~, we failed to parse and further analyze 2.367 programs due technical difficulties with the provided JSON files.
@@
 Once we obtained the Scratch programs, we parsed the JSON files according to the specification of the format\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}}.
-This resulted in a list if used blocks per program, with the sprites and the stage of the program.
-We also assign types to all the blocks, both their shapes and the tab they originate from.
+This resulted in a list of used blocks per program, with the sprites and the stage of the program.
+We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks.
 For example, When Green Flag Clicked is a \emph{Hat block} from the \emph{Events category}.
@@
 
-\subsection{Program size and complexity}
+\subsection{Program Size and Complexity}
 
-The dataset is found to contain a relatively small number of projects without any code: \nemptyPrograms (5.77\%).
-Through random manual sampling we found that in some cases those projects contained sprites and costumes but no code and in others they were entirety empty apart from the Scratch cat added by default.
+The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%).
+Through random manual sampling we found that in some cases those projects contains only sprites and costumes, but no code, while in other programs they were entirety empty apart from the Scratch cat added by default.
 Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.
@@
 In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity.
-We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project, where each script is a code-block of interconnected Scratch blocks.
-For example, in Figure \ref{fig:ui} the cat sprite contains two scripts.
-We further measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines.
-Furthermore, we measure the nestedness of program, which is the number of times blocks are placed within each other.
-For example, the program in Figure \ref{fig:ui} has a maximum indentation of 3 in the `Wgen Green flag clicked' script, as that has a broadcast block within an if, within a forever\todo{for Felienne: double check}.
+We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project.
+We furthermore measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines.
 Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.
@@
 We find that the majority of Scratch projects are small; 75\% of the projects have up to 5 sprites, 12 scripts and 76 lines of code, while one fourth of the projects have up to 12 lines of code.
-On the other end, 5\% of the projects (11,712) are found to have more than 18 sprites and 4.8\% (11,214) consist of more than 500 lines of code.
+On the other end, 5\% of the projects (11,712) have more than 18 sprites and 4.8\% (11,214) consist of more than 500 lines of code.
 The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.
@@
 More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects.
-Those projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.
+The projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.
 
-As a complexity metric we also used the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code.
+To understand the complexity of the Scratch programs in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code.
 This is calculated per script by counting the number of decision points in the script plus one.
@@
 The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}.
-The majority (78.33\%) of \nscripts scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point.
+The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point.
 The complexity is higher, over 4, for 3.67\% of the scripts.
@@
 
-\subsection{Programming abstractions and concepts}
+\subsection{Programming Abstractions and Concepts}
 
 The first method for abstraction that we investigate is the use of procedures.
-In the dataset we found 206,799 procedures in 17,979 projects.
+In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects.
 As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of those projects having up to 2.
 Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects.
-Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\%\% have only one (shown in Figure \ref{fig:procedurearguments}).
+Regarding procedure arguments, we found that 55.57\% have no arguments and 19.48\% have only one (shown in Figure \ref{fig:procedurearguments}).
 The majority of procedure arguments (80.59\%) are numeric, and the least used argument type is the boolean one---6.23\% of the total procedure arguments, found in 5.32\% of the procedures.
@@
 Most procedures are called exactly once (62.32\% of them) or twice (14.30\%) and from exactly one script (85.92\% of them).
-Examining the origin of procedure calls, we found remarkable that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, thus making it recursive.
-Those recursive procedures are found in 1,052 projects.\footnote{list of recursive procedures published \todo{link}}.
+Examining the origin of procedure calls, we observed that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, making them recursive calls.
+Those recursive procedures are found in 1,052 projects\footnote{list of recursive procedures published \todo{link}}.
 
@@
 
-As shown in table \ref{tbl-characteristics}, almost one-third of the projects utilize variables and a small number (4.01\%) utilize lists.
+As shown in Table \ref{tbl-characteristics}, almost one-third of the projects use variables and a small number (4.01\%) use lists.
 The number of variables that is being used is also limited, with only 7.48\% of the projects having 5 or more variables.
@@
 
-The scope of variables was also analyzed: As shown in Table \ref{tbl-size}, each variable is used by an average of 4.97 different scripts, with a median value of 3.
+%commented by Felienne: 1) this is not really scope and 2) we cannot distinguish, so why mention it?
+%The scope of variables was also analyzed: As shown in Table \ref{tbl-size}, each variable is used by an average of 4.97 different scripts, with a median value of 3.
 However, we must note that this measurement might not be entirely accurate, because variables might be defined in the scope of specific sprites.
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/8variableslists}
-	\caption{Histogram of the number of variables and lists utilized in the projects}
+	\caption{Histogram of the number of variables and lists used in the projects}
 	\label{fig:variables}
@@
 Table \ref{tbl-userinput} lists the frequency of use of user input controls.
-The most commonly used one is the \texttt{when key pressed}, found in 71,096 (30.45\% of the total) projects.
+We do not include the \texttt{when Green flag pressed} block here, as this is just used to start a Scratch program and hence cannot really be considered input into the program.
+The most commonly used user input block is the \texttt{when key pressed}, found in 71,096 (30.45\% of the total) projects.
 The most frequently used parameter for the \texttt{key} attribute is the space key, followed by the arrows and then the letters and numbers.
@@
 
-For event management Scratch offers blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}. 
-Those blocks are utilized by 29.57\% of the projects.
+Users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}. 
+Those blocks are used by 29.57\% of the projects.
 \texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects.
@@
 	\begin{minipage}{0.46\textwidth}
-RQ2: One third of the projects utilize variables, sometimes without initializing them.
-39.81\% of the projects utilize conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used.
+RQ2: One third of the projects use variables, sometimes without initializing them.
+39.81\% of the projects contain conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used.
 More than half of the projects are interactive.
-29.57\% of the projects utilize broadcast and receive blocks, but in 7.99\% of the projects they are not synchronized.
+29.57\% of the projects use broadcast and receive blocks, but in 7.99\% of the projects they are not synchronized.
 Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/11clonesprojects}
-	\caption{Histogram of the number of cloned scripts found in the dataset projects}
+	\caption{Histogram of the number of cloned scripts}
 	\label{fig:clonesprojects}

### 2016-04-22 Paper/paper.tex
Added blocks per category and type in RQ1

--- 2016-04-22 Paper/paper.tex
@@
 Scratch is a \emph{block-based} language: users manipulate blocks to program.
+Block-based languages are visual languages, but also use some successful aspects of text-based languages such as limited text-entry and indentation, and as such are closer to `real', textual programming than other forms of visual programming, like dataflow languages are.
 
@@
 
-Several studies have shown that block-based languages are powerful as a tool for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003}.
-\todo{Fenia, do you have some work on Research on programming habits}
-
-Recent work has focused on \emph{code smells} within Scratch programs, parts of a program that are not designed in the best possible way.
-\todo{add dr.
-scratch papers here} Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \todo{cite}.
+refer to:
+Research on the effectiveness of those languages for programming education
+Research on programming habits
+Our research on smells
 
@@
 Those projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.
+
+We further analyzed the utilization frequency of the different block shapes and categories, as defined in the Scratch documentation.
+Figures \ref{fig:categories} and \ref{fig:shapes} present the results in terms of number of blocks from the total \nLOC blocks in the dataset projects.
+The most commonly used blocks are from the Control and Data categories.
+The Others category includes the blocks related to procedure calls and arguments.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/16Categories}
+	\caption{Number of blocks from each category in the analyzed projects}
+	\label{fig:categories}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/16Shapes}
+	\caption{Number of blocks of each shape in the analyzed projects}
+	\label{fig:shapes}
+\end{figure}
 

### 2016-04-22 Paper/paper.tex

--- 2016-04-22 Paper/paper.tex
@@
 
+We further analyzed the utilization frequency of the different block shapes and categories, as defined in the Scratch documentation.
+Figures \ref{fig:categories} and \ref{fig:shapes} present the results in terms of number of blocks from the total \nLOC blocks in the dataset projects.
+The most commonly used blocks are from the Control and Data categories.
+The Others category includes the blocks related to procedure calls and arguments.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/16Categories}
+	\caption{Number of blocks from each category in the analyzed projects}
+	\label{fig:categories}
+\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/16Shapes}
+	\caption{Number of blocks of each shape in the analyzed projects}
+	\label{fig:shapes}
+\end{figure}
+
 To understand the complexity of the Scratch programs in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code.

### 2016-04-22 Paper/paper.tex
further changes

--- 2016-04-22 Paper/paper.tex
@@
 
-\begin{figure}
-	\begin{center}
-		\includegraphics[width=2cm]{fig/bump.png}
-		\caption{The script for the Piano sprite in Figure \ref{fig:ui}, causing a sound on receiving the `bump' signal.}
-		\label{fig:bump}
-	\end{center}
-\end{figure}
+%\begin{figure}
+%	\begin{center}
+%		\includegraphics[width=2cm]{fig/bump.png}
+%		\caption{The script for the Piano sprite in Figure \ref{fig:ui}, causing a sound on receiving the `bump' signal.}
+%		\label{fig:bump}
+%	\end{center}
+%\end{figure}
 
@@
 In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano.
-The Piano reacts on that by making a sound, see Figure \ref{fig:bump}.
 
-\subsubsection{Procedures}
-Scratch users can define their own blocks, with a pop-up screen displayed in on the left hand side of Figure \ref{fig:twoBlocks}.
-As shown in this figure, Scratch users can enter the name of the sprite and define input parameters.
-The block in Figure \ref{fig:twoBlocks} uses two parameter, one textual (string1) and one numerical (number1).
-A user-created block is programmed using ordinary Scratch blocks as shown on the right hand side of Figure \ref{fig:twoBlocks}.
+\subsubsection{Custom Blocks}
+Scratch users can define their own blocks, which users can name themselves, called custom blocks.
+To align with other papers on this topic \todo{Fenia, a reference?} we will refer to them as `procedures' in the following.
+Procedures can have input parameters of type string, number, and boolean.
+When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
 
-\begin{figure}
-	\begin{center}
-		\includegraphics[width=8cm]{fig/twoBlocks.png}
-		\caption{The user interface for creating (left) and defining (right) user-defined blocks.}
-		\label{fig:twoBlocks}
-	\end{center}
-\end{figure}
+%\begin{figure}
+%	\begin{center}
+%		\includegraphics[width=8cm]{fig/twoBlocks.png}
+%		\caption{The user interface for creating (left) and defining (right) user-defined blocks.}
+%		\label{fig:twoBlocks}
+%	\end{center}
+%\end{figure}
 
@@
 
-Apart from whole scripts we also examined cases with only the first block being different.
-In first block is commonly the event block and, by separating this one from the subsequent functionality blocks, we could examine if programmers assign the same functionality to handle different types of events.
+Apart from whole scripts we also examined cases where scripts differed only in the first (Hat) block.This way we examine if Scratch programmers assign the same functionality to handle different types of events.
 Cloned functionality blocks are found to be rare: without considering the first block, only 2,243 additional clones were found in 920 projects.
@@
 Examining the size of those dead scripts, 72.34\% are composed of a single block.
-As shown in diagram \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of those scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610\footnote{\todo{link}}.
+As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of those scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610\footnote{\todo{link}}.
 
@@
 Investigating the use of clones between programs, we observe that there are 1700 scripts that are used in multiple programs, sometimes as often as in 1600 different programs.
-This seems to indicate that there are common patterns in Scratch programs.
-This again leads us to the idea that it might be very beneficial to Scratch programmers if they could not only share their programs, but also share some of their functionality, for others to use, like a library.
+This seems to indicate that there are common patterns in Scratch programs, which means it might be very beneficial to Scratch programmers if they could not only share their programs, but also share some of their functionality, for others to use, like a library.
 An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'.

### 2016-04-22 Paper/paper.tex
added conclusions

@@
 
+\section{Conclusions}
+\label{sec:conclusion}
+In this paper we presented a large scale study on \nAnalyzedPrograms projects we scraped from the Scratch repository.
+We analyzed those projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality.
+We found that functions and conditional loops are not commonly used.
+We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells.
+Our findings indicate that Scratch programs suffer from code smells and especially dead code and code duplication.
+
+In addition to the findings presented in this paper, we provide as contributions the dataset that we used for our study \todo{do we?}, as well as information on the edge cases that we found in the dataset in terms of size and number of procedures, variables, cyclomatic complexity, clones and dead code.
+
 \bibliographystyle{abbrv}

### 2016-04-22 Paper/paper.tex
Replaced those with these

@@
 Block-based programming languages like Scratch, Alice and Blockly are becoming increasingly common as introductory languages in programming education.
-There is substantial research showing that those visual programming environments are suitable for teaching programming concepts.
+There is substantial research showing that these visual programming environments are suitable for teaching programming concepts.
 But, what do people do when they use Scratch?
@@
 To that end we have scraped the Scratch public repository and retrieved 250.000 projects.
-We present an analysis of those projects in three different dimensions.
+We present an analysis of these projects in three different dimensions.
 Firstly, we look at the types of blocks used and the size of the programs.
@@
 The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%).
-Through random manual sampling we found that in some cases those projects contains only sprites and costumes, but no code, while in other programs they were entirety empty apart from the Scratch cat added by default.
-Since those projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.
+Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other programs they were entirety empty apart from the Scratch cat added by default.
+Since these projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.
 
@@
 We furthermore measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines.
-Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of those size metrics.
+Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of these size metrics.
 
@@
 On the other end, 5\% of the projects (11,712) have more than 18 sprites and 4.8\% (11,214) consist of more than 500 lines of code.
-The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.
+The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code\footnote{The Scratch identifiers of these projects can be found in \todo{link}}.
 
@@
 The complexity is higher, over 4, for 3.67\% of the scripts.
-The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of those projects can be found in \todo{link}}.
+The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of these projects can be found in \todo{link}}.
 Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.
@@
 In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects.
-As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of those projects having up to 2.
+As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of these projects having up to 2.
 Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects.
@@
 Examining the origin of procedure calls, we observed that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, making them recursive calls.
-Those recursive procedures are found in 1,052 projects\footnote{list of recursive procedures published \todo{link}}.
+These recursive procedures are found in 1,052 projects\footnote{list of recursive procedures published \todo{link}}.
 
@@
 Users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}. 
-Those blocks are used by 29.57\% of the projects.
+These blocks are used by 29.57\% of the projects.
 \texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects.
@@
 In total, in the dataset we found 170,532 scripts cloned across sprites in 59,634 (25.54\% of the total non-empty) projects.
-726,316 copies of those scripts were found, making each clone being copied an average of 4.26 times.
+726,316 copies of these scripts were found, making each clone being copied an average of 4.26 times.
 Figure \ref{fig:clonesprojects} plots the distribution of clones across projects.
@@
 A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects.
-The majority of those scripts (86.6\%) are scripts missing the starting block.
-Examining the size of those dead scripts, 72.34\% are composed of a single block.
-As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of those scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610\footnote{\todo{link}}.
+The majority of these scripts (86.6\%) are scripts missing the starting block.
+Examining the size of these dead scripts, 72.34\% are composed of a single block.
+As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610\footnote{\todo{link}}.
 
@@
 Finally, we examine the long method and the large class smell, considering them in the context of Scratch as large script and large sprite smells respectively.
-For those two smells we use the number of blocks as the size metric.
+For these two smells we use the number of blocks as the size metric.
 Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset.
-We used those numbers to split the dataset and retrieve the top 10\% \todo{felienne, can we justify why 10 percent?} largest scripts and sprites, and in this way we found the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
+We used these numbers to split the dataset and retrieve the top 10\% \todo{felienne, can we justify why 10 percent?} largest scripts and sprites, and in this way we found the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
 Using these thresholds, the number of projects exhibiting the large script smell, i.e., containing at least one script with 18 or more blocks, is 69,521 (29.77\% of the total projects) and the number of projects with the large sprite smell is 31,954 (13.68\%).
@@
 24.36\% of the projects contain scripts that are never invoked, and thus exhibit the dead code smell.
-In some cases those scripts are large.
+In some cases these scripts are large.
 The large script smell is found in 29.77\% of the projects and the large sprite in 13.68\%.
@@
 In this paper we presented a large scale study on \nAnalyzedPrograms projects we scraped from the Scratch repository.
-We analyzed those projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality.
+We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality.
 We found that functions and conditional loops are not commonly used.

### 2016-04-22 Paper/paper.tex
put table in column

--- 2016-04-22 Paper/paper.tex
@@
 
-\begin{table*}[ht]
+\begin{table}[ht]
 	\centering
@@
 		\hline
-		&\textbf{Number of}&\\
-		&\textbf{projects}&\textbf{Percentage}\\
+		&\textbf{\# projects}&\textbf{\%}\\
 		\hline
@@
 		Loop statements & 180,210 & 77.18\% \\
-		\texttt{repeat until <condition>} statements & 31,739 & 13.59\% \\
-		\texttt{broadcast} - \texttt{receive} statements & 69,039 & 29.57\%\\
+		\texttt{repeat until <condition>} & 31,739 & 13.59\% \\
+		\texttt{broadcast} - \texttt{receive} & 69,039 & 29.57\%\\
 		Cloned scripts across sprites & 59,634 & 25.54\% \\
@@
 		Cloned procedures & 4,945 & 2.12\%\\
-		Cloned functionality blocks across sprites & 60,554 & 25.93\%\\
+		Cloned blocks across sprites & 60,554 & 25.93\%\\
 		Exact clones across sprites & 27,574 & 11.81\%\\
@@
 	\label{tbl-characteristics}
-\end{table*}
+\end{table}
 

### 2016-04-22 Paper/paper.tex
added recursive procedures to stats table

--- 2016-04-22 Paper/paper.tex
@@
 		Procedures & 17,979 & 7.70\%\\
+		Recursive procedures &  1,052 & 0.45\%\\
 		Variables & 73,577 & 31.51\% \\

### 2016-04-22 Paper/paper.tex
rephrased discussion

@@
 
+\subsubsection{Popularity of Different Block Types}
+Our analysis shows that some categories of blocks are rarely used, like the `Pen' blocks, of which only \todo{x} occur within \todo{y} programs.
+When further evolving the Scratch programming language, this category might be less important to users, and hence our current analysis supports the language designers in making future investments.
+
 \subsubsection{Dead Code}
@@
 We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.
-
-\subsubsection{Sharing of Scripts and Procedures}
-Investigating the use of clones between programs, we observe that there are 1700 scripts that are used in multiple programs, sometimes as often as in 1600 different programs.
-This seems to indicate that there are common patterns in Scratch programs, which means it might be very beneficial to Scratch programmers if they could not only share their programs, but also share some of their functionality, for others to use, like a library.
-An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'.
-This might empower new Scratch users to get started quicker.
 

### 2016-04-22 Paper/paper.tex
Moved descriptions from Results to Section 3

--- 2016-04-22 Paper/paper.tex
@@
 The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch programs.
-To answer our three research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch repository.
-The dataset is described in Section \ref{dataset}.
-\todo{Below, we present how we approached each research question.}
-
-\textbf{RQ1} What are the size and complexity characteristics of Scratch programs?
-quantitative analysis, provide and analyze descriptive statistics on the size, complexity metrics, statistical analysis .
-mentioned taht it was imported to db and done and statistical data from db queries?
-The outcomes are presented in Section...
-
-\textbf{RQ2} Which coding abstractions and programming concepts and features are commonly used when programming in the Scratch environment?
-Using the code in the dataset, we perform a detailed statistical analysis of code characteristics to answer RQ2 (Section ??).
-We manually assigned types to commands/blocks.
-
-\textbf{RQ3} Which code smells and bad programming practices are common in Scratch programs?  to examine if coding style conventions are used, we qualitatively analyze..
+To answer our three research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch project repository.
+In the following paragraphs we describe the dataset, the  process and the tools we used for analyzing it, and the methods we followed for detecting code smells.
 
@@
 \label{dataset}
-\subsubsection{Obtaining Scratch programs}
+
 We obtained the set of Scratch programs by using a scraping program.
@@
 Out of the \nPrograms~, we failed to parse and further analyze 2.367 programs due technical difficulties with the provided JSON files.
-Kragle, as well as all scraped programs and our analysis files are available\footnote{\url{https://github.com/ScratchLover42/ICER-Data-Code}}.
+Kragle, as well as all scraped programs and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}
 
-\subsubsection{Analyzing the Scratch programs}
-Once we obtained the Scratch programs, we parsed the JSON files according to the specification of the format\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}}.
-This resulted in a list of used blocks per program, with the sprites and the stage of the program.
+Once we obtained the Scratch programs, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per program, with the sprites and the stage of the program.
 We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks.
 For example, When Green Flag Clicked is a \emph{Hat block} from the \emph{Events category}.
+
+\subsection{Data analysis}
+\label{dataAnalysis}
+All scraped data, including the list of used blocks and parameters per project, was imported to a relational database.
+We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions.
+This data was then imported to spreadsheets for the calculation of statistics and the generation of graphical representations.
+We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects, lists of which we also provide as part of the dataset.\footnotemark[\ref{repo}]
+
+For RQ1 we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}.
+We used the McCabe cyclomatic complexity metric\cite{mcCabe76} for measuring the complexity of the projects.
+
+For RQ2, we used the data on the code blocks and their categories to perform a detailed statistical analysis of applied programming abstractions and concepts.
+Similarly to \cite{Maloney_2008}, we consider the use of certain blocks to indicate that a programming abstraction or concept is being used in a certain project.
+In Section \ref{RQ2} we present the results related to the utilization of procedures, variables, loops, conditional statements, user interactivity and synchronization.
+
+For RQ3, we focused on four types of code smells: duplicated code, dead code, large script and large sprite.
+For the duplicated code smell analysis, our first step was to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same program.
+For the identification of clones we did not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal.
+We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}.
+The next step in the analysis was to determine the minimum size of the scripts that are considered clones instead of incidentally similar.
+For this, we examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}.
+Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/11cloneslines}
+	\caption{Number of cloned scripts of different block sizes across and within sprites}
+	\label{fig:cloneslines}
+\end{figure}
+
+To examine the long method and the large class smells, we consider them in the context of Scratch as large script and large sprite smells respectively.
+For these two smells we use the number of blocks as the size metric.
+Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset.
+We used these numbers to split the dataset and retrieve the top 10\% \todo{felienne, can we justify why 10 percent?} largest scripts and sprites, and in this way we found the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
+The results we obtained using these thresholds are presented in Section \ref{RQ3}.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\textwidth]{fig/charts/13longmethod}
+	\caption{Size of sprites and scripts in number of blocks}
+	\label{fig:longmethod}
+\end{figure}
 
@@
 \subsection{Program Size and Complexity}
-
+\label{RQ1}
 The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%).
@@
 On the other end, 5\% of the projects (11,712) have more than 18 sprites and 4.8\% (11,214) consist of more than 500 lines of code.
-The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code\footnote{The Scratch identifiers of these projects can be found in \todo{link}}.
+The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code, whose Scratch identifiers are made available for further inspection.\footnotemark[\ref{repo}]
 
@@
 The complexity is higher, over 4, for 3.67\% of the scripts.
-The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246\footnote{The Scratch identifiers of these projects can be found in \todo{link}}.
-Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.
+The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246.\footnotemark[\ref{repo}] Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.
 
@@
 \subsection{Programming Abstractions and Concepts}
-
+\label{RQ2}
 The first method for abstraction that we investigate is the use of procedures.
@@
 Examining the origin of procedure calls, we observed that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, making them recursive calls.
-These recursive procedures are found in 1,052 projects\footnote{list of recursive procedures published \todo{link}}.
+These recursive procedures are found in 1,052 projects, whose identifiers are made available.\footnotemark[\ref{repo}]
 
@@
 \subsection{Code smells and bad programming practices}
-
+\label{RQ3}
 The duplicated code smell is the first smell that we examine.
-The first step for our analysis is to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same program.
-For the identification of clones we do not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal.
-We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}.
-The next step in the analysis is to determine the minimum size of the scripts that are considered clones instead of incidentally similar.
-For this, we examine the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}.
-Based on this distribution, we opt to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.45\textwidth]{fig/charts/11cloneslines}
-	\caption{Number of cloned scripts of different block sizes across and within sprites}
-	\label{fig:cloneslines}
-\end{figure}
-
+As explained in Section \ref{dataAnalysis}, we use 5 as the minimum number of blocks for the identified clones.
 In total, in the dataset we found 170,532 scripts cloned across sprites in 59,634 (25.54\% of the total non-empty) projects.
@@
 Figure \ref{fig:clonescopies} plots the number of copies of the identified clones.
-It is of interest that 79,378 (46.55\%) of the identified clones are copied three or more times, and even in 585 cases from 411 projects they are copied more than 50 times and up to 974 \footnote{\todo{link}}.
-Manual inspection of the mostly copied clones revealed that they were scripts used multiple times to cater for different parameter values, sometimes in multiple sprites.
+It is of interest that 79,378 (46.55\%) of the identified clones are copied three or more times, and even in 585 cases from 411 projects they are copied more than 50 times and up to 974.\footnotemark[\ref{repo}] Manual inspection of the mostly copied clones revealed that they were scripts used multiple times to cater for different parameter values, sometimes in multiple sprites.
 
@@
 Examining the size of these dead scripts, 72.34\% are composed of a single block.
-As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610\footnote{\todo{link}}.
+As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}]
 
@@
 
-Finally, we examine the long method and the large class smell, considering them in the context of Scratch as large script and large sprite smells respectively.
-For these two smells we use the number of blocks as the size metric.
-Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset.
-We used these numbers to split the dataset and retrieve the top 10\% \todo{felienne, can we justify why 10 percent?} largest scripts and sprites, and in this way we found the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
-Using these thresholds, the number of projects exhibiting the large script smell, i.e., containing at least one script with 18 or more blocks, is 69,521 (29.77\% of the total projects) and the number of projects with the large sprite smell is 31,954 (13.68\%).
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.45\textwidth]{fig/charts/13longmethod}
-	\caption{Size of sprites and scripts in number of blocks}
-	\label{fig:longmethod}
-\end{figure}
+Finally, we examine the large script and the large sprite smells.
+The thresholds we use for the identification of large scripts and large sprites are 18 blocks and 59 blocks respectively, as calculated in Section \ref{dataAnalysis}.
+The number of projects exhibiting the large script smell, i.e., containing at least one script with 18 or more blocks, is 69,521 (29.77\% of the total projects) and the number of projects with the large sprite smell is 31,954 (13.68\%).
 
@@
 
-In addition to the findings presented in this paper, we provide as contributions the dataset that we used for our study \todo{do we?}, as well as information on the edge cases that we found in the dataset in terms of size and number of procedures, variables, cyclomatic complexity, clones and dead code.
+In addition to the findings presented in this paper, we provide as contributions the dataset that we used for our study, as well as information on the edge cases that we found in the dataset in terms of size and number of procedures, variables, cyclomatic complexity, clones and dead code.\footnotemark[\ref{repo}]
 

### 2016-04-22 Paper/paper.tex
Removed indentation metrics

@@
 		Scripts per project&17.35&1&2&5&12&3038\\
-		Maximum indentation per project&4.56&1&3&3&5&488\\
 		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
@@
 	\centering
-	\includegraphics[width=0.45\textwidth]{fig/charts/1indent}
-	\caption{Histogram of the maximum code indentation level in the analyzed projects}
-	\label{fig:size-indent}
-\end{figure}
-\begin{figure}
-	\centering
 	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
@@
 We furthermore measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines.
-Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and \ref{fig:size-indent} plot the distribution of these size metrics.
+Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and plot the distribution of these size metrics.
 

### 2016-04-22 Paper/paper.tex
Added line to table 2
changed program to project

@@
 We present an analysis of these projects in three different dimensions.
-Firstly, we look at the types of blocks used and the size of the programs.
+Firstly, we look at the types of blocks used and the size of the projects.
 We then investigate complexity, used abstractions and programming concepts.
 Finally we look at coding style and detect \emph{code smells and clones}.
-We find that programming abstraction concepts like functions are not commonly used.
+We find that programming abstraction concepts like procedures are not commonly used.
 We further investigate the presence of code smells and bad programming practices, including code duplication, dead code, long method and large class smells.
@@
 
-Our study is based on data from the Scratch program repository.
-By scraping the list of recent programs \footnote{\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch programs and performed analysis on them.
+Our study is based on data from the Scratch project repository.
+By scraping the list of recent projects \footnote{\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed analysis on them.
 
@@
 \begin{itemize}
-	\item{A public data set of \nScriptPrograms~Scratch programs (Section \ref{dataset})}
+	\item{A public data set of \nScriptPrograms~Scratch projects (Section \ref{dataset})}
 	\item{An evaluation of the data set in terms of (Section \ref{sec:results})}
@@
 
-The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch programs.
-To answer our three research questions, we conducted an empirical quantitative evaluation of program data we collected from the Scratch project repository.
+The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch projects.
+To answer our three research questions, we conducted an empirical quantitative evaluation of project data we collected from the Scratch project repository.
 In the following paragraphs we describe the dataset, the  process and the tools we used for analyzing it, and the methods we followed for detecting code smells.
@@
 
-We obtained the set of Scratch programs by using a scraping program.
-Our scraping program called Kragle starts by reading \url{https://scratch.mit.edu/explore/projects/all/} and thus obtaining program ids of programs that were most recently shared.
-Subsequently, Kragle obtains the JSON code for each of the listed programs.
+We obtained the set of Scratch projects by using a scraping program.
+Our scraping program called Kragle starts by reading \url{https://scratch.mit.edu/explore/projects/all/} and thus obtaining project ids of projects that were most recently shared.
+Subsequently, Kragle obtains the JSON code for each of the listed projects.
 
@@
 
-We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 programs.
-Out of the \nPrograms~, we failed to parse and further analyze 2.367 programs due technical difficulties with the provided JSON files.
-Kragle, as well as all scraped programs and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}
+We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 projects.
+Out of the \nPrograms~, we failed to parse and further analyze 2.367 projects due technical difficulties with the provided JSON files.
+Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}
 
-Once we obtained the Scratch programs, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per program, with the sprites and the stage of the program.
+Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per project, with the sprites and the stage of the project.
 We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks.
@@
 For RQ3, we focused on four types of code smells: duplicated code, dead code, large script and large sprite.
-For the duplicated code smell analysis, our first step was to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same program.
+For the duplicated code smell analysis, our first step was to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same project.
 For the identification of clones we did not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal.
@@
 The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%).
-Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other programs they were entirety empty apart from the Scratch cat added by default.
+Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other projects they were entirety empty apart from the Scratch cat added by default.
 Since these projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.
@@
 		&\textbf{Number of}&\\
-		&\textbf{projects}&\textbf{Percentage}\\
+		&\textbf{projects}& \textbf{Percentage}\\
 		\hline
@@
 		Non-empty (used for statistics) & \textbf{233,491} & \\
+
+		\textit{Projects with:} & & \\
 		\hline
@@
 	\end{tabular}
-	\caption{Characteristics in the projects in the dataset}
+	\caption{Elements and characteristics of the projects in the dataset}
 	\label{tbl-characteristics}
@@
 
-In Table \ref{tbl-size} we summarize statistics for metrics related to size and complexity.
+In Table \ref{tbl-size} we summarize statistics for the analyzed metrics.
 We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project.
@@
 
-To understand the complexity of the Scratch programs in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code.
+To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76},  a quantitative measure of the number of linearly independent paths through a program's source code.
 This is calculated per script by counting the number of decision points in the script plus one.
@@
 \subsubsection{Dead Code}
-In our analysis, we find that almost one quarter of the Scratch programs contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality.
+In our analysis, we find that almost one quarter of the Scratch projects contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality.
 \todo{Fenia, this is correct, right, the 23\% is just these types and not also uncalled procs and unmatched broadcasts?}.
@@
 \subsubsection{Exact Clones between Sprites}
-With occurrences in 11\% of the Scratch programs in our dataset, the use of exactly identical clones between sprites is relatively common.
+With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common.
 In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them.
@@
 \subsubsection{Sharing of Scripts and Procedures}
-Investigating the use of clones between programs, we observe that there are 1700 scripts that are used in multiple programs, sometimes as often as in 1600 different programs.
-This seems to indicate that there are common patterns in Scratch programs, which means it might be very beneficial to Scratch programmers if they could not only share their programs, but also share some of their functionality, for others to use, like a library.
+Investigating the use of clones between projects, we observe that there are 1700 scripts that are used in multiple projects, sometimes as often as in 1600 different projects.
+This seems to indicate that there are common patterns in Scratch projects, which means it might be very beneficial to Scratch programmers if they could not only share their projects, but also share some of their functionality, for others to use, like a library.
 An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'.
@@
 \subsection{Threads to validity}
-A thread to validity of this study is the fact that we did not scrap a random sample, but the most recent 250,000 programs.
+A thread to validity of this study is the fact that we did not scrap a random sample, but the most recent 250,000 projects.
 It could be the case that programming habits of Scratch users are changing over time.
 However, we counterbalanced that by using a very large dataset.
-Furthermore, for some of the analyses we manually inspected programs \todo{Fenia, which ones?}
+Furthermore, for some of the analyses we manually inspected projects \todo{Fenia, which ones?}
 
-Furthermore we use the number of blocks in the Scratch programs as a measure for the length of a program, while this does not exactly correspond to the `length' of a program in lines, and there can be multiple Scratch blocks on one line.
+Furthermore we use the number of blocks in the Scratch projects as a measure for the length of a program, while this does not exactly correspond to the `length' of a program in lines, and there can be multiple Scratch blocks on one line.
 For example, in Figure \ref{fig:ui} the `if' and `touching' blocks are present on the same `line'.
@@
 We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality.
-We found that functions and conditional loops are not commonly used.
+We found that procedures and conditional loops are not commonly used.
 We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells.

### 2016-04-22 Paper/paper.tex
Removed mentions of bad programming practices

@@
 We then investigate complexity, used abstractions and programming concepts.
-Finally we look at coding style and detect \emph{code smells and clones}.
+Finally we detect \emph{code smells and clones}.
 We find that programming abstraction concepts like procedures are not commonly used.
-We further investigate the presence of code smells and bad programming practices, including code duplication, dead code, long method and large class smells.
+We further investigate the presence of code smells, namely code duplication, dead code, long method and large class smells.
 Our findings indicate that Scratch programs suffer from code smells and especially code duplication.
@@
 \item[RQ2] Which coding abstractions and programming concepts and features are commonly used when programming in the Scratch environment?
-\item[RQ3] How common are code smells and bad programming practices in Scratch programs?
+\item[RQ3] How common are code smells in Scratch programs?
 \end{itemize}
@@
 Examining the initialization of variables through the \texttt{set <variable> to <value>} blocks, we found that for 4.83\% of all variables this was missing.
-While failing to initialize a variable in Scratch will not result in a runtime error as in some other programming languages, it is considered bad style \todo{citation needed}.
+While failing to initialize a variable in Scratch will not result in a runtime error as in some other programming languages, correctly setting the initial state of the program is important \cite{boe_hairball:_2013}.
 
@@
 
-\subsection{Code smells and bad programming practices}
+\subsection{Code smells}
 \label{RQ3}

### 2016-04-22 Paper/paper.tex
Changed lines of code to number of blocks

@@
 		Scripts per project&17.35&1&2&5&12&3038\\
-		Lines of code (LOC) per project&154.55&1&12&29&76&34622\\
-		LOC in Stage per project&4.80&0&0&0&3&2613\\
-		LOC in Sprites per project&115.57&0&10&26&68&34613\\
-		LOC in Procedures per project&34.17&0&0&0&0&20552\\
+		Number of blocks per project&154.55&1&12&29&76&34622\\
+		Blocks in Stage per project&4.80&0&0&0&3&2613\\
+		Blocks in Sprites per project&115.57&0&10&26&68&34613\\
+		Blocks in Procedures per project&34.17&0&0&0&0&20552\\
 		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
-	\caption{Histogram of the lines of code in the analyzed projects}
+	\caption{Histogram of the number of blocks in the analyzed projects}
 	\label{fig:size-locs}
@@
 In Table \ref{tbl-size} we summarize statistics for the analyzed metrics.
-We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts per project.
-We furthermore measure the lines of code per project, considering each Scratch block to represent exactly one line of code, thus not counting the closing constructs of loops or conditional statements as separate lines.
+We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts and blocks per project.
 Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and plot the distribution of these size metrics.
 
-We find that the majority of Scratch projects are small; 75\% of the projects have up to 5 sprites, 12 scripts and 76 lines of code, while one fourth of the projects have up to 12 lines of code.
-On the other end, 5\% of the projects (11,712) have more than 18 sprites and 4.8\% (11,214) consist of more than 500 lines of code.
-The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 lines of code, whose Scratch identifiers are made available for further inspection.\footnotemark[\ref{repo}]
+We find that the majority of Scratch projects are small; 75\% of the projects have up to 5 sprites, 12 scripts and 76 blocks, while one fourth of the projects have up to 12 blocks.
+On the other end, 5\% of the projects (11,712) have more than 18 sprites and 4.8\% (11,214) consist of more than 500 blocks.
+The analysis also highlighted some surprisingly large projects: 135 with more than 300 sprites and even 30 projects with more than 20,000 blocks, whose Scratch identifiers are made available for further inspection.\footnotemark[\ref{repo}]
 
-The lines of code metric was further analyzed to understand code organization.
-As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC~lines of code---is written within sprites.
-An additional 3.1\% of the total lines are found in the stage class.
-More interestingly, the remaining 22.11\% are lines within defined procedures, which are found in only 7.7\% (17,979) of the projects.
-The projects that contain procedures use them a lot; almost half of their total lines of code (48.81\%) are lines within procedures.
+The number of blocks metric was further analyzed to understand code organization.
+As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC~blocks---is written within sprites.
+An additional 3.1\% of the total blocks are found in the stage class.
+More interestingly, the remaining 22.11\% are blocks within defined procedures, which are found in only 7.7\% (17,979) of the projects.
+The projects that contain procedures use them a lot; almost half of their total blocks (48.81\%) are within procedures.
 
@@
 	\begin{minipage}{0.46\textwidth}
-		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 lines of code and no decision points.
+		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 blocks and no decision points.
 Most code is written in sprites.
@@
 		\includegraphics[width=6cm]{fig/Unconnected_clones.png}
-		\caption{Cloned blocks that are not connected to an event. Program id: 12237615}
+		\caption{Cloned blocks that are not connected to an event. Project id: 12237615}
 		\label{fig:Unconnected_clones}

### 2016-04-22 Paper/paper.tex
Changes recursive loops to loops

@@
 		Conditional statements per project&10.02&0&0&0&3&5950\\
-		Recursive loop statements per project&7.65&0&1&2&5&2503\\
+		Loop statements per project&7.65&0&1&2&5&2503\\
 		User input blocks per project&4.77&0&0&1&4&1889\\
@@
 Regarding program control features, conditional statements (blocks \texttt{if <condition> then} and \texttt{if <condition> then else}) are used by 39.81\% of the projects.
-Recursive loops (blocks \texttt{repeat <times>}, \texttt{forever} and \texttt{repeat until <condition>}) are more common, used by 77.18\% of the projects.
-The most common of the three is the \texttt{forever} block, accounting for 51.86\% of all recursive loops and the least common one is the \texttt{repeat until <condition>} block, accounting for 11.57\% and used in 13.59\% of the total projects.
+Loops (blocks \texttt{repeat <times>}, \texttt{forever} and \texttt{repeat until <condition>}) are more common, used by 77.18\% of the projects.
+The most common of the three is the \texttt{forever} block, accounting for 51.86\% of all loops and the least common one is the \texttt{repeat until <condition>} block, accounting for 11.57\% and used in 13.59\% of the total projects.
 
@@
 RQ2: One third of the projects use variables, sometimes without initializing them.
-39.81\% of the projects contain conditional statements and 77.18\% recursive loops, but conditional recursion is rarely used.
+39.81\% of the projects contain conditional statements and 77.18\% contain loops, but conditional loops are rarely used.
 More than half of the projects are interactive.

### 2016-04-22 Paper/paper.tex
updated the dead code smell

@@
 		Exact clones within sprites & 2,043 & 0.87\%\\
-		Dead code & 56,890	& 24.36\%\\
+		Dead code & 65,760 & 28.16\%\\
 		Large scripts & 69,521 & 29.77\%\\
@@
 The use of procedures in projects was further investigated through the procedure calls, summarized in Figure \ref{fig:procedurecalls}.
-A significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects.
 Most procedures are called exactly once (62.32\% of them) or twice (14.30\%) and from exactly one script (85.92\% of them).
@@
 
-Users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}. 
-These blocks are used by 29.57\% of the projects.
+Users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}.These blocks are used by 29.57\% of the projects.
 \texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects.
-The events are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received.
-This lack of syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).
 
@@
 More than half of the projects are interactive.
-29.57\% of the projects use broadcast and receive blocks, but in 7.99\% of the projects they are not synchronized.
+29.57\% of the projects use broadcast and receive blocks.
 Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
@@
 The second smell that we examine is the dead code smell.
-Already in the previous section, Scratch projects were found to have two types of dead code: defined procedures that are not called (5.06\% of the defined procedures in 2,079 projects) and unmatched broadcast-receive messages (in 7.99\% of the projects).
-In this section we examine an additional case: scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality.
+We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts.
+Investigating the first type, we find that a significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects.
+This is also shown in Figure \ref{fig:procedurecalls} and it occurs in 2,079 projects.
+For the second type, we examined the broadcast-receive messages and found that they are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received.
+This lack of syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).
+
+The third and fourth cases are detected in scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality.
 A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects.
@@
 
-\begin{figure}
-	\begin{center}
-		\includegraphics[width=6cm]{fig/Unconnected_clones.png}
-		\caption{Cloned blocks that are not connected to an event. Project id: 12237615}
-		\label{fig:Unconnected_clones}
-	\end{center}
-\end{figure}
+The number of projects exhibiting the dead code smell, considering all four types of dead code, is 65,760 (28.16\% of the total projects).
 
@@
 RQ3: Code clones are found in 25.54\% of the projects, with 46.55\% of the clones copied three or more times, in the same or across sprites.
-24.36\% of the projects contain scripts that are never invoked, and thus exhibit the dead code smell.
+28.16\% of the projects contain code that is never invoked, and thus exhibit the dead code smell.
 In some cases these scripts are large.

### 2016-04-22 Paper/paper.tex
Added info on introduction

@@
 Several studies have shown that block-based languages are powerful as a tool for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003}.
-\todo{Fenia, do you have some work on Research on programming habits}
+Research works involving static analysis of Scratch programs \cite{Maloney_2008} \cite{moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively.
+Recent work has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells and bad programming practices \cite{boe_hairball:_2013}\cite{moreno_automatic_2014}.
+Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \todo{cite}.
 
-Recent work has focused on \emph{code smells} within Scratch programs, parts of a program that are not designed in the best possible way.
-\todo{add dr.
-scratch papers here} Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \todo{cite}.
-
-The goal of this paper is to obtain a deep understanding of how people program in Scratch and to analyze the characteristics of Scratch programs.
+The goal of this paper is to obtain a deep understanding of how people program in Scratch, to analyze the characteristics of Scratch programs, and to quantitatively evaluate the use of programming abstractions and concepts.
 Moreover, knowing that bad programming habits and code smells can be harmful, we also want to explore whether they are common.
@@
 By scraping the list of recent projects \footnote{\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed analysis on them.
+To the best of our knowledge, this is the first large-scale exploratory study on Scratch projects.
 

### 2016-04-22 Paper/paper.tex
Explanation for the use of term procedures

@@
 Scratch users can define their own blocks, which users can name themselves, called custom blocks.
-To align with other papers on this topic \todo{Fenia, a reference?} we will refer to them as `procedures' in the following.
+The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}.
+Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following.
 Procedures can have input parameters of type string, number, and boolean.
@@
 \subsubsection{Dead Code}
-In our analysis, we find that almost one quarter of the Scratch projects contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, like in the example in Figure \ref{fig:Unconnected_clones}, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality.
-\todo{Fenia, this is correct, right, the 23\% is just these types and not also uncalled procs and unmatched broadcasts?}.
+In our analysis, we find that more than one quarter of the Scratch projects contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, scripts that are comprised of only a \texttt{when <trigger>} block without any functionality, procedures that are not invoked, or unmatched broadcast-receive messages.
 

### 2016-04-22 Paper/paper.tex
added justification for the 10%

@@
 Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset.
-We used these numbers to split the dataset and retrieve the top 10\% \todo{felienne, can we justify why 10 percent?} largest scripts and sprites, and in this way we found the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
+We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artefacts like spreadsheets~\cite{hermans2015detecting}.
+Using that strategy, we set the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
 The results we obtained using these thresholds are presented in Section \ref{RQ3}.
@@
 
-Also related to our research are works on code smells, initiated by the work by Fowler~\cite{fowler_refactoring:_1999}.
+%Also related to our research are works on code smells, initiated by the work by Fowler~\cite{fowler_refactoring:_1999}.
 His book gives an overview of code smells and corresponding refactorings.
@@
 \label{sec:conclusion}
-In this paper we presented a large scale study on \nAnalyzedPrograms projects we scraped from the Scratch repository.
+In this paper we presented a large scale study on \nAnalyzedPrograms~projects we scraped from the Scratch repository.
 We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality.

### 2016-04-22 Paper/paper.tex
Fixed table 3

--- 2016-04-22 Paper/paper.tex
@@
 \newcommand{\nScriptPrograms}{233,491}
-\newcommand{\nProgramsWithClones}{67,177}
 \newcommand{\nLOC}{36,085,654}
 \newcommand{\nscripts}{4,049,356}
-\newcommand{\dup}{\emph{Duplication}~}
 
@@
 We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks.
-For example, When Green Flag Clicked is a \emph{Hat block} from the \emph{Events category}.
+For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
 
@@
 \label{RQ1}
-The dataset contains a relatively small number of projects without any code: \nemptyPrograms (5.77\%).
+The dataset contains a relatively small number of projects without any code: \nemptyPrograms~ (5.77\%).
 Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other projects they were entirety empty apart from the Scratch cat added by default.
@@
 	\centering
-	\begin{tabular}{lrrrrrr}
+	\begin{tabular}{llrrrrrr}
+		&&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
 		\hline
-		&\textbf{mean}&\textbf{min}&\textbf{Q1}&\textbf{median}&\textbf{Q3}&\textbf{max}\\
+		Size&Sprites with code per project&5.68&1&1&2&5&525\\
+		&Scripts per project&17.35&1&2&5&12&3,038\\
+		&Number of blocks per project&154.55&1&12&29&76&34,622\\
+		&Blocks in Stage per project&4.80&0&0&0&3&2,613\\
+		&Blocks in Sprites per project&115.57&0&10&26&68&34,613\\
+		&Blocks in Procedures per project&34.17&0&0&0&0&20,552\\
 		\hline
-		Sprites with code per project&5.68&1&1&2&5&525\\
-		Scripts per project&17.35&1&2&5&12&3038\\
-		Number of blocks per project&154.55&1&12&29&76&34622\\
-		Blocks in Stage per project&4.80&0&0&0&3&2613\\
-		Blocks in Sprites per project&115.57&0&10&26&68&34613\\
-		Blocks in Procedures per project&34.17&0&0&0&0&20552\\
-		McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
-		McCabe CC per procedure script&3.75&1&1&2&4&183\\
-		Procedures per project with procedures&11.50&1&1&2&6&847\\
-		Arguments per Procedure&0.95&0&0&0&1&53\\
-		Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
-		Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
-		Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
-		Calls per procedure&2.14&0&1&1&2&526\\
-		Scripts with calls per procedure&1.13&0&1&1&1&59\\
-		Variables per project&2.06&0&0&0&1&340\\
-		Scripts utilizing variable&4.97&1&1&3&5&1127\\
-		Lists per project&0.55&0&0&0&0&319\\
-		Conditional statements per project&10.02&0&0&0&3&5950\\
-		Loop statements per project&7.65&0&1&2&5&2503\\
-		User input blocks per project&4.77&0&0&1&4&1889\\
-		Broadcast-receive statements per project&8.57&0&0&0&2&2460\\	
+		Complexity&McCabe Cyclomatic Complexity (CC) per script&1.58&1&1&1&1&246\\
+		&McCabe CC per procedure script&3.75&1&1&2&4&183\\
+		\hline
+		Procedures&Procedures per project with procedures&11.50&1&1&2&6&847\\
+		&Arguments per Procedure&0.95&0&0&0&1&53\\
+		&Numerical arguments per procedure with arguments&1.73&0&1&1&2&22\\
+		&Text arguments per procedure with arguments&0.28&0&0&0&1&24\\
+		&Boolean arguments per procedure with arguments&0.13&0&0&0&0&14\\
+		&Calls per procedure&2.14&0&1&1&2&526\\
+		&Scripts with calls per procedure&1.13&0&1&1&1&59\\
+		\hline
+		Programming&Variables per project&2.06&0&0&0&1&340\\
+		concepts&Scripts utilizing variable&4.97&1&1&3&5&1,127\\
+		&Lists per project&0.55&0&0&0&0&319\\
+		&Conditional statements per project&10.02&0&0&0&3&5,950\\
+		&Loop statements per project&7.65&0&1&2&5&2,503\\
+		&User input blocks per project&4.77&0&0&1&4&1,889\\
+		&Broadcast-receive statements per project&8.57&0&0&0&2&2,460\\	
 		\hline
@@
 We further analyzed the utilization frequency of the different block shapes and categories, as defined in the Scratch documentation.
-Figures \ref{fig:categories} and \ref{fig:shapes} present the results in terms of number of blocks from the total \nLOC blocks in the dataset projects.
+Figures \ref{fig:categories} and \ref{fig:shapes} present the results in terms of number of blocks from the total \nLOC~blocks in the dataset projects.
 The most commonly used blocks are from the Control and Data categories.
@@
 	\end{tabular}
-	\caption{Frequency of use of user input blocks in the \nScriptPrograms projects of the dataset}
+	\caption{Frequency of use of user input blocks in the \nScriptPrograms~projects of the dataset}
 	\label{tbl-userinput}

### 2016-04-22 Paper/paper.tex
small change in abstract and moved fowler to intro

--- 2016-04-22 Paper/paper.tex
@@
 We then investigate complexity, used abstractions and programming concepts.
-Finally we detect \emph{code smells and clones}.
+Finally we detect \emph{code smells} like long scripts and duplicated code blocks.
 We find that programming abstraction concepts like procedures are not commonly used.
@@
 Research works involving static analysis of Scratch programs \cite{Maloney_2008} \cite{moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively.
-Recent work has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells and bad programming practices \cite{boe_hairball:_2013}\cite{moreno_automatic_2014}.
+Recent work has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013}\cite{moreno_automatic_2014}.
 Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \todo{cite}.

### 2016-04-22 Paper/paper.tex
fixed todos

--- 2016-04-22 Paper/paper.tex
@@
 Recent work has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells and bad programming practices \cite{boe_hairball:_2013}\cite{moreno_automatic_2014}.
-Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \todo{cite}.
+Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
 
@@
 Figure \ref{fig:clonescopies} plots the number of copies of the identified clones.
-It is of interest that 79,378 (46.55\%) of the identified clones are copied three or more times, and even in 585 cases from 411 projects they are copied more than 50 times and up to 974.\footnotemark[\ref{repo}] Manual inspection of the mostly copied clones revealed that they were scripts used multiple times to cater for different parameter values, sometimes in multiple sprites.
+It is of interest that 79,378 (46.55\%) of the identified clones are copied three or more times, and even in 585 cases from 411 projects they are copied more than 50 times and up to 974.\footnotemark[\ref{repo}]
 
@@
 
-We further inspected which of the identified clones were duplicated only within the same sprite: 63,682 (37.34\% of the total) clones, in 10.14\% of the projects.
-These are duplication smells that could be avoided by using procedures, which in Scratch can only be defined for specific sprites.
-The lack of support for project-wide procedures could be the cause for procedure clones.
-These were measured to 12,878 (7.55\% of the total) clones, in 2.12\% of the projects.
+We further inspected which of the identified clones were duplicated only within the same sprite: 63,682 (37.34\% of the total) clones, in 10.14\% of all projects.
+Procedure clones were measured to 12,878 (7.55\% of the total) clones, in 2.12\% of the projects.
 
@@
 \subsubsection{Popularity of Different Block Types}
-Our analysis shows that some categories of blocks are rarely used, like the `Pen' blocks, of which only \todo{x} occur within \todo{y} programs.
+Our analysis shows that some categories of blocks are rarely used, like the `Pen' blocks, of which only 194,885 occur within 19,090 (8.17\% of the total) programs.
 When further evolving the Scratch programming language, this category might be less important to users, and hence our current analysis supports the language designers in making future investments.
@@
 However, we counterbalanced that by using a very large dataset.
-Furthermore, for some of the analyses we manually inspected projects \todo{Fenia, which ones?}
 

### 2016-04-22 Paper/paper.tex

@@
 Research works involving static analysis of Scratch programs \cite{Maloney_2008} \cite{moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively.
-Recent work has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells and bad programming practices \cite{boe_hairball:_2013}\cite{moreno_automatic_2014}.
+Recent work has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013}\cite{moreno_automatic_2014}.
 Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.

### 2016-04-22 Paper/paper.tex
shortened background section

@@
 
-Scratch is a programming language developed to teach children programming by enabling them to create games and interactive animations.
+Scratch ~\cite{resnick_scratch:_2009} is a programming language developed to teach children programming by enabling them to create games and interactive animations.
 The public repository of Scratch programs contains over 12 million projects.
@@
 
-Several studies have shown that block-based languages are powerful as a tool for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003}.
-Research works involving static analysis of Scratch programs \cite{Maloney_2008} \cite{moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively.
-Recent work has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013}\cite{moreno_automatic_2014}.
+Several studies have shown that block-based languages are powerful as a tool for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}.
+Research works involving static analysis of Scratch programs \cite{Maloney_2008,moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively.
+Recent work has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
 Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
@@
 
-\section{Background and Motivation}
-\label{sec:background}
-Recently new block-based languages have gained widespread popularity, especially powered by Scratch~\cite{resnick_scratch:_2009} and Blockly\footnote{\url{https://developers.google.com/blockly/}}.
-Over 100 million students have tried Blockly via Code.org, and the Scratch repository currently hosts over 12 million projects.
-
-Since their introduction, studies have demonstrated the applicability of block-based languages as a tool for education.
-Scratch, for example, was evaluated with a two-hour introductory programming curriculum for 46 subjects aged 14~\cite{meerbaum-salant_learning_2010}.
-This study indicated that Scratch could be used to teach computer science concepts: analysis of the pre- and post-tests showed a significant improvement after the Scratch course, although some concepts like variables and concurrency remained hard for students.
-
-Moskal \emph{et al.}~\cite{b._moskal_evaluating_2005} compared computer science students who studied Alice before or during their first programming course to students that only took the introductory computer science course.
-Their results show that exposure to Alice significantly improved students' grades in the course, and their retention in computer science in general over a two year period.
-A follow-up study by Cooper \emph{et al.}~\cite{cooper_teaching_2003} obtained similar results, showing that a curriculum in Alice resulted in improved grades and higher retention in computer science.
-
-Most convincingly, Price and Barnes performed a controlled experiment in which students were randomly assigned to either a text-based or a block-based interface in  which they had to perform small programming tasks~\cite{price_comparing_2015}.
-Their experiment showed that students in the block-based interface were more focused and completed more of the activity's goals in less time.
-
-\subsection{Relevant Scratch Concepts}
+\section{Relevant Scratch Concepts}
 \label{sec:scratch}
@@
 Figure \ref{fig:ui} shows the Scratch user interface in the Chrome browser.
+The main concepts in the Scratch programming environment are:
 
-\subsubsection{Sprites}
-Scratch code is organized by `sprites': two-dimensional pictures each having their own associated code.
+\begin{description}
+	\item [Sprites] Scratch code is organized by `sprites': two-dimensional pictures each having their own associated code.
 Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes.
@@
 The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
+	\item [Scripts]Sprites can have multiple code blocks, called `scripts'.
+The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar.
+It is possible for a single sprite to have multiple scripts initiated by the same event.
+In that case, all scripts will be executed simultaneously.
+	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}).
+The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface.
+In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam.
+The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when right arrow key pressed}, each with associated blocks.
+	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
+All other sprites can then react by using the \texttt{when I receive} Hat block.
+In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano.
+	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks.
+The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}.
+Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following.
+Procedures can have input parameters of type string, number, and boolean.
+When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
+\end{description}
 
@@
 \end{figure}
-
-%\begin{figure}
-%	\begin{center}
-%		\includegraphics[width=2cm]{fig/bump.png}
-%		\caption{The script for the Piano sprite in Figure \ref{fig:ui}, causing a sound on receiving the `bump' signal.}
-%		\label{fig:bump}
-%	\end{center}
-%\end{figure}
-
-\subsubsection{Scripts}
-Sprites can have multiple code blocks, called `scripts'.
-The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar.
-It is possible for a single sprite to have multiple scripts initiated by the same event.
-In that case, all scripts will be executed simultaneously.
-
-\subsubsection{Events}
-Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}).
-The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface.
-In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam.
-The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when right arrow key pressed}, each with associated blocks.
-
-\subsubsection{Signals}
-Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
-All other sprites can then react by using the \texttt{when I receive} Hat block.
-In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano.
-
-\subsubsection{Custom Blocks}
-Scratch users can define their own blocks, which users can name themselves, called custom blocks.
-The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}.
-Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following.
-Procedures can have input parameters of type string, number, and boolean.
-When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
-
-%\begin{figure}
-%	\begin{center}
-%		\includegraphics[width=8cm]{fig/twoBlocks.png}
-%		\caption{The user interface for creating (left) and defining (right) user-defined blocks.}
-%		\label{fig:twoBlocks}
-%	\end{center}
-%\end{figure}
 

### 2016-04-22 Paper/paper.tex
new abstract

@@
 Finally we detect \emph{code smells} like long scripts and duplicated code blocks.
-We find that programming abstraction concepts like procedures are not commonly used.
-We further investigate the presence of code smells, namely code duplication, dead code, long method and large class smells.
-Our findings indicate that Scratch programs suffer from code smells and especially code duplication.
+Our results show that 1) most Scratch program are small, however Scratch programs consisting of over 100 sprites exist, 2) programming abstraction concepts like procedures are not commonly used and 3) Scratch programs suffer do commonly from code smells including long methods and unmatched broadcast signals.
 \end{abstract}

### 2016-04-22 Paper/paper.tex
made the table fit

@@
 		&\textbf{Number of}&\\
-		&\textbf{projects}& \textbf{Percentage}\\
+		&\textbf{projects}& \textbf{\%}\\
 		\hline
@@
 		Non-empty (used for statistics) & \textbf{233,491} & \\
+		\hline
+		\textit{Projects with:} & & \\
 
-		\textit{Projects with:} & & \\
-		\hline
 		Procedures & 17,979 & 7.70\%\\

### 2016-04-22 Paper/paper.tex
Removed double spaces
discussion update

@@
 \subsection{Implications}
-Our results have some interesting implications for designers of educational programming languages.
+We believe a large scale study of programs like ours can help language designers to tailor their language. In this section we highlight directions in which our study could support language design.
 
@@
 Our analysis shows that some categories of blocks are rarely used, like the `Pen' blocks, of which only 194,885 occur within 19,090 (8.17\% of the total) programs.
-When further evolving the Scratch programming language, this category might be less important to users, and hence our current analysis supports the language designers in making future investments.
+Hence, in future changes to the language, `Pen' blocks might be less important to users to support or maintain.
 
@@
 
-\subsubsection{Exact Clones between Sprites}
+\subsubsection{Exact Clones Between Sprites}
 With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common.
@@
 \subsection{Threads to validity}
-A thread to validity of this study is the fact that we did not scrap a random sample, but the most recent 250,000 projects.
+A thread to validity of this study is the fact that we did not scrape a random sample, but the most recent 250,000 projects.
 It could be the case that programming habits of Scratch users are changing over time.
-However, we counterbalanced that by using a very large dataset.
+However, we counterbalanced that by using a large dataset which comprises of around 2\% of all 14 million shared Scratch programs.\footnote{\url{https://scratch.mit.edu/statistics/}}.
 

### 2016-04-22 Paper/paper.tex
minor chanes

--- 2016-04-22 Paper/paper.tex
@@
 
-Scratch ~\cite{resnick_scratch:_2009} is a programming language developed to teach children programming by enabling them to create games and interactive animations.
-The public repository of Scratch programs contains over 12 million projects.
+Scratch \cite{resnick_scratch:_2009} is a programming language developed to teach children programming by enabling them to create games and interactive animations.
+The public repository of Scratch programs contains over 14 million projects.
 Scratch is a \emph{block-based} language: users manipulate blocks to program.
@@
 
-Several studies have shown that block-based languages are powerful as a tool for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}.
+Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}.
 Research works involving static analysis of Scratch programs \cite{Maloney_2008,moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively.
-Recent work has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
-Ar recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
+Recent works has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
+A recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
 
@@
 \begin{itemize}
-	\item{A public data set of \nScriptPrograms~Scratch projects (Section \ref{dataset})}
-	\item{An evaluation of the data set in terms of (Section \ref{sec:results})}
-	\item{A discussion of implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
+	\item{A public data set of \nScriptPrograms~non-empty Scratch projects (Section \ref{dataset})}
+	\item{An evaluation of the data set in terms of size, complexity, utilized programming concepts and features and identified smells (Section \ref{sec:results})}
+	\item{A discussion of the implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
 \end{itemize}
@@
 \begin{description}
-	\item [Sprites] Scratch code is organized by `sprites': two-dimensional pictures each having their own associated code.
+	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code.
 Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes.
@@
 The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
-	\item [Scripts]Sprites can have multiple code blocks, called `scripts'.
+	\item [Scripts]Sprites can have multiple code blocks, called scripts.
 The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar.
@@
 		\includegraphics[width=\columnwidth]{fig/ui.png}
-		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `Events' in the middle and the code associated with the sprite on the right.
-The upper right corner shows the actual location of the sprite.}
+		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `Events' in the middle and the code associated with the sprite on the right.}
 		\label{fig:ui}

### 2016-04-22 Paper/paper.tex
minor changes

@@
 Our study is based on data from the Scratch project repository.
-By scraping the list of recent projects \footnote{\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed analysis on them.
+By scraping the list of recent projects \footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed analysis on them.
 To the best of our knowledge, this is the first large-scale exploratory study on Scratch projects.
@@
 We obtained the set of Scratch projects by using a scraping program.
-Our scraping program called Kragle starts by reading \url{https://scratch.mit.edu/explore/projects/all/} and thus obtaining project ids of projects that were most recently shared.
+Our scraping program called Kragle starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtaining project ids of projects that were most recently shared.
 Subsequently, Kragle obtains the JSON code for each of the listed projects.
@@
 \label{dataAnalysis}
-All scraped data, including the list of used blocks and parameters per project, was imported to a relational database.
+All scraped project data, including the list of used blocks and parameters, was imported to a relational database.
 We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions.
@@
 
-Apart from whole scripts we also examined cases where scripts differed only in the first (Hat) block.This way we examine if Scratch programmers assign the same functionality to handle different types of events.
+Apart from whole scripts we also examined cases where scripts differed only in the first (Hat) block.
+This way we examine if Scratch programmers assign the same functionality to handle different types of events.
 Cloned functionality blocks are found to be rare: without considering the first block, only 2,243 additional clones were found in 920 projects.
@@
 For the second type, we examined the broadcast-receive messages and found that they are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received.
-This lack of syncronizations occurs in 18,669 projects (7.99\% of the total non-empty ones).
+This lack of syncronizations occurs in 18,669 (7.99\% of the total) projects.
 
@@
 Examining the size of these dead scripts, 72.34\% are composed of a single block.
-As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}]
+As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}] The number of projects exhibiting the dead code smell, considering all four types of dead code, is 65,760 (28.16\% of the total projects).
 
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/12deadcode}
-	\caption{Size of scripts that are identified as dead code}
+	\caption{Size of scripts identified as dead code}
 	\label{fig:deadcode}
 \end{figure}
-
-The number of projects exhibiting the dead code smell, considering all four types of dead code, is 65,760 (28.16\% of the total projects).
 
@@
 
-\subsubsection{Popularity of Different Block Types}
+\subsubsection{Popularity of different block types}
 Our analysis shows that some categories of blocks are rarely used, like the `Pen' blocks, of which only 194,885 occur within 19,090 (8.17\% of the total) programs.
@@
 
-\subsubsection{Dead Code}
+\subsubsection{Dead code}
 In our analysis, we find that more than one quarter of the Scratch projects contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, scripts that are comprised of only a \texttt{when <trigger>} block without any functionality, procedures that are not invoked, or unmatched broadcast-receive messages.
@@
 
-\subsubsection{Exact Clones Between Sprites}
+\subsubsection{Exact clones between sprites}
 With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common.
@@
 
-\subsubsection{Sharing of Scripts and Procedures}
+\subsubsection{Sharing of scripts and procedures}
 Investigating the use of clones between projects, we observe that there are 1700 scripts that are used in multiple projects, sometimes as often as in 1600 different projects.

### 2016-04-22 Paper/paper.tex
very small discussion update
Bass changes upto and including pag 3

@@
 Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}.
-Research works involving static analysis of Scratch programs \cite{Maloney_2008,moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively.
+Previous work involving static analysis of Scratch programs \cite{Maloney_2008,moreno_automatic_2014} has evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively.
 Recent works has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
@@
 Moreover, knowing that bad programming habits and code smells can be harmful, we also want to explore whether they are common.
-Specifically, the research questions that we are trying to answer are:
+To address this goal, we answer the following research questions:
 
@@
 Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes.
-The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a Piano.
+The Scratch program in Figure \ref{fig:ui} contains two one sprites, the cat, which is Scratch's default sprite and a Piano.
 The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
@@
 The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}.
-Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following.
+Because the term `procedures' is common in related work, we will refer to custom blocks as `procedures' in the following.
 Procedures can have input parameters of type string, number, and boolean.
@@
 We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 projects.
-Out of the \nPrograms~, we failed to parse and further analyze 2.367 projects due technical difficulties with the provided JSON files.
+Out of the \nPrograms~, we failed to parse and further analyze 2.367 projects due to technical difficulties with the provided JSON files.
 Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}
@@
 \label{sec:results}
-In the following sections we describe the results obtained for each of the research questions through the analysis of the \nAnalyzedPrograms~Scratch projects in our dataset.
+In the following sections, for each of the research questions, we describe the results obtained through the analysis of the  \nAnalyzedPrograms~Scratch projects in our dataset.
 
@@
 We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts and blocks per project.
-Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and plot the distribution of these size metrics.
+Figures \ref{fig:size-sprites}, \ref{fig:size-scripts} and \ref{fig:size-locs} plot the distribution of these size metrics.
 
@@
 \subsection{Implications}
-We believe a large scale study of programs like ours can help language designers to tailor their language. In this section we highlight directions in which our study could support language design.
+We believe a large scale study of programs like ours can help language designers to tailor their language.
+In this section we highlight directions in which our study could support language design.
+There are many other implications to be considered, which is why we have made our dataset public.
 

### 2016-04-22 Paper/paper.tex
Sohon changes upto and including pag 3

@@
 In this paper we explore the characteristics of Scratch programs.
-To that end we have scraped the Scratch public repository and retrieved 250.000 projects.
+To that end we have scraped the Scratch public repository and retrieved 250,000 projects.
 We present an analysis of these projects in three different dimensions.
@@
 
-We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 projects.
-Out of the \nPrograms~, we failed to parse and further analyze 2.367 projects due to technical difficulties with the provided JSON files.
+We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250,000 projects.
+Out of the \nPrograms~, we failed to parse and further analyze 2,367 projects due to technical difficulties with the provided JSON files.
 Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}

### 2016-04-22 Paper/paper.tex
minor changes

--- 2016-04-22 Paper/paper.tex
@@
 In this paper we explore the characteristics of Scratch programs.
-To that end we have scraped the Scratch public repository and retrieved 250.000 projects.
+To this end we have scraped the Scratch public repository and retrieved 250.000 projects.
 We present an analysis of these projects in three different dimensions.
-Firstly, we look at the types of blocks used and the size of the projects.
+Initially, we look at the types of blocks used and the size of the projects.
 We then investigate complexity, used abstractions and programming concepts.
-Finally we detect \emph{code smells} like long scripts and duplicated code blocks.
-Our results show that 1) most Scratch program are small, however Scratch programs consisting of over 100 sprites exist, 2) programming abstraction concepts like procedures are not commonly used and 3) Scratch programs suffer do commonly from code smells including long methods and unmatched broadcast signals.
+Finally we detect \emph{code smells} like large scripts, dead code and duplicated code blocks.
+Our results show that 1) most Scratch programs are small, however Scratch programs consisting of over 100 sprites exist, 2) programming abstraction concepts like procedures are not commonly used and 3) Scratch programs do suffer from code smells including large scripts and unmatched broadcast signals.
 \end{abstract}
@@
 Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}.
-Research works involving static analysis of Scratch programs \cite{Maloney_2008,moreno_automatic_2014} have evaluated the use of various programming concepts in 536 and 100 Scratch projects respectively.
-Recent works has focused on bad programming practices and habits within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
-A recent controlled experiment found that long scripts and duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
+Research works involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}.
+Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
+A recent controlled experiment found that long scripts and code duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
 
 The goal of this paper is to obtain a deep understanding of how people program in Scratch, to analyze the characteristics of Scratch programs, and to quantitatively evaluate the use of programming abstractions and concepts.
-Moreover, knowing that bad programming habits and code smells can be harmful, we also want to explore whether they are common.
+Moreover, knowing that bad programming habits and code smells can be harmful \cite{hermans_2016}, we also want to explore whether they are common.
 Specifically, the research questions that we are trying to answer are:
@@
 Our study is based on data from the Scratch project repository.
-By scraping the list of recent projects \footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed analysis on them.
-To the best of our knowledge, this is the first large-scale exploratory study on Scratch projects.
+By scraping the list of recent projects,\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} we have obtained \nPrograms~public Scratch projects and performed analysis on them.
+To the best of our knowledge, this is the first large-scale exploratory study of Scratch projects.
 
@@
 	\item{A public data set of \nScriptPrograms~non-empty Scratch projects (Section \ref{dataset})}
-	\item{An evaluation of the data set in terms of size, complexity, utilized programming concepts and features and identified smells (Section \ref{sec:results})}
+	\item{An evaluation of the data set in terms of size, complexity, utilized programming concepts and features, and identified smells (Section \ref{sec:results})}
 	\item{A discussion of the implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
@@
 Scratch is a block-based programming language aimed at children, developed by MIT.
-Scratch can be used to create games and interactive animations, and is available both as a stand-alone application and as a web application.
-Figure \ref{fig:ui} shows the Scratch user interface in the Chrome browser.
+Scratch can be used to create games and interactive animations, and is available both as a stand-alone and as a web application.
+Figure \ref{fig:ui} shows the Scratch user interface.
 The main concepts in the Scratch programming environment are:
@@
 Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes.
-The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a Piano.
+The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a piano.
 The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
@@
 
-We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250.000 projects.
-Out of the \nPrograms~, we failed to parse and further analyze 2.367 projects due technical difficulties with the provided JSON files.
+We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250,000 projects.
+Out of the \nPrograms~, we failed to parse and further analyze 2,367 projects due technical difficulties with the provided JSON files.
 Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}
@@
 \label{dataAnalysis}
-All scraped project data, including the list of used blocks and parameters, was imported to a relational database.
+All scraped project data, including the list of used blocks and parameters, was imported in a relational database.
 We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions.
-This data was then imported to spreadsheets for the calculation of statistics and the generation of graphical representations.
 We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects, lists of which we also provide as part of the dataset.\footnotemark[\ref{repo}]
@@
 For RQ1 we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}.
-We used the McCabe cyclomatic complexity metric\cite{mcCabe76} for measuring the complexity of the projects.
+We used the McCabe cyclomatic complexity metric\cite{mcCabe76} for measuring the complexity of the scripts.
 
@@
 Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset.
-We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artefacts like spreadsheets~\cite{hermans2015detecting}.
+We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artifacts like spreadsheets~\cite{hermans2015detecting}.
 Using that strategy, we set the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
-	\caption{Histogram of the number of sprites in the analyzed projects}
+	\caption{Number of sprites in the analyzed projects}
 	\label{fig:size-sprites}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
-	\caption{Histogram of the number of scripts in the analyzed projects}
+	\caption{Number of scripts in the analyzed projects}
 	\label{fig:size-scripts}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
-	\caption{Histogram of the number of blocks in the analyzed projects}
+	\caption{Number of blocks in the analyzed projects}
 	\label{fig:size-locs}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
-	\caption{Histogram of the McCabe cyclomatic complexity of the \nscripts~analyzed scripts}
+	\caption{McCabe cyclomatic complexity of the \nscripts~analyzed scripts}
 	\label{fig:cyclomatic}
@@
 
-To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76}, a quantitative measure of the number of linearly independent paths through a program's source code.
+To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76}, a quantitative measure of the number of independent paths through a program's source code.
 This is calculated per script by counting the number of decision points in the script plus one.
@@
 \label{RQ2}
-The first method for abstraction that we investigate is the use of procedures.
+The first method for abstraction that we investigate are procedures.
 In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects.
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/6proceduresperproject}
-	\caption{Histogram of the number of procedures for the 17,979 projects that include at least one}
+	\caption{Number of procedures for the 17,979 projects that include at least one}
 	\label{fig:proceduresperproject}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/5arguments}
-	\caption{Histogram of the number of arguments for the procedures in the dataset}
+	\caption{Number of arguments for the procedures in the dataset}
 	\label{fig:procedurearguments}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/7procedurecalls}
-	\caption{Histogram of the number of calls of each procedure in the dataset}
+	\caption{Number of calls of each procedure in the dataset}
 	\label{fig:procedurecalls}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/8variableslists}
-	\caption{Histogram of the number of variables and lists used in the projects}
+	\caption{Number of variables and lists used in the projects}
 	\label{fig:variables}
@@
 Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
-Recursive procedure calls exist in 1052 (0.45\% of the total) projects.
+Recursive procedure calls exist in 1,052 (0.45\% of the total) projects.
 	\end{minipage}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/11clonesprojects}
-	\caption{Histogram of the number of cloned scripts}
+	\caption{Cloned scripts in the dataset projects}
 	\label{fig:clonesprojects}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/11clonescopies}
-	\caption{Histogram of the number of copies of the identified clones}
+	\caption{Number of copies of the identified clones}
 	\label{fig:clonescopies}
@@
 For the second type, we examined the broadcast-receive messages and found that they are not always synchronized: 3,33\% of the \texttt{when I receive} blocks were found to wait for a message that is never being broadcasted, while 4,4\% of the \texttt{broadcast} blocks broadcast a message that is not being received.
-This lack of syncronizations occurs in 18,669 (7.99\% of the total) projects.
+This lack of syncronization occurs in 18,669 (7.99\% of the total) projects.
 
@@
 In some cases these scripts are large.
-The large script smell is found in 29.77\% of the projects and the large sprite in 13.68\%.
+The large script smell is found in 29.77\% of the projects and the large sprite one in 13.68\%.
 	\end{minipage}}
@@
 
-In a sense, the dead scripts are harmless, as they are not executed, however, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead.
-In contrast with other visual educational languages, Scratch does on indicate scripts are dead.
+In a sense, the dead scripts are harmless, as they are not executed.
+However, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead.
+In contrast to other visual educational languages, Scratch does on indicate scripts that are dead.
 LEGO Mindstorms, for example, does give the user feedback by making unconnected blocks gray.
@@
 We envision that would be like the `backpack' meant to move scripts across sprites.
-In order to help novice programmers to keep their code clean, the programming interface could actively encourage users to move unconnected blocks to that workspace when they exit the environment.
+In order to help novice programmers keep their code clean, the programming interface could actively encourage users to move unconnected blocks to that workspace when they exit the environment.
 
@@
 In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them.
-So in many cases there is no other way to share the functionality than by making a copy.
+So in many cases there is no way to share the functionality other than by making a copy.
 We are not aware of the underlying rationale of the Scratch team that lead to this decision, however it seems that a large part of the Scratch users would use the functionality to call procedures between sprites.
@@
 \subsubsection{Sharing of scripts and procedures}
-Investigating the use of clones between projects, we observe that there are 1700 scripts that are used in multiple projects, sometimes as often as in 1600 different projects.
+Investigating the use of clones between projects, we observe that there are 1,700 scripts that are used in multiple projects, sometimes as often as in 1,600 different projects.
 This seems to indicate that there are common patterns in Scratch projects, which means it might be very beneficial to Scratch programmers if they could not only share their projects, but also share some of their functionality, for others to use, like a library.
 An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'.
-This might empower new Scratch users to get started quicker.
+This might empower new Scratch users to get started faster.
 
-\subsection{Threads to validity}
-A thread to validity of this study is the fact that we did not scrape a random sample, but the most recent 250,000 projects.
-It could be the case that programming habits of Scratch users are changing over time.
-However, we counterbalanced that by using a large dataset which comprises of around 2\% of all 14 million shared Scratch programs.\footnote{\url{https://scratch.mit.edu/statistics/}}.
+\subsection{Threats to validity}
+A threat to the validity of this study is the fact that we did not scrape a random sample, but the most recent 250,000 projects.
+It could be the case that the programming habits of Scratch users are changing over time.
+However, we counterbalanced that by using a large dataset which comprises of around 2\% of all 14 million shared Scratch projects.\footnote{\url{https://scratch.mit.edu/statistics/}}.
 
@@
 \label{sec:conclusion}
-In this paper we presented a large scale study on \nAnalyzedPrograms~projects we scraped from the Scratch repository.
+In this paper we presented a large-scale study on 247,798 projects we scraped from the Scratch repository.
 We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality.
@@
 We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells.
-Our findings indicate that Scratch programs suffer from code smells and especially dead code and code duplication.
+Our findings indicate that Scratch programs suffer from code smells and especially from dead code and code duplication.
 

### 2016-04-22 Paper/paper.tex
Changes based on Felienne's comments

@@
 Our study is based on data from the Scratch project repository.
-By scraping the list of recent projects,\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} we have obtained \nPrograms~public Scratch projects and performed analysis on them.
+By scraping the list of recent projects,\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} we have obtained \nPrograms~public Scratch projects and performed source code analysis on them.
 To the best of our knowledge, this is the first large-scale exploratory study of Scratch projects.
@@
 	\item{A public data set of \nScriptPrograms~non-empty Scratch projects (Section \ref{dataset})}
-	\item{An evaluation of the data set in terms of size, complexity, utilized programming concepts and features, and identified smells (Section \ref{sec:results})}
+	\item{An evaluation of the data set in terms of size, complexity, programming concepts and smells (Section \ref{sec:results})}
 	\item{A discussion of the implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
@@
 In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam.
-The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when right arrow key pressed}, each with associated blocks.
+The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when <right arrow> key pressed}, each with associated blocks.
 	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
@@
 We obtained the set of Scratch projects by using a scraping program.
-Our scraping program called Kragle starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtaining project ids of projects that were most recently shared.
-Subsequently, Kragle obtains the JSON code for each of the listed projects.
+Our scraping program called Kragle starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtains project ids of projects that were most recently shared.
+Subsequently, Kragle retrieves the JSON code for each of the listed projects.
 
@@
 We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250,000 projects.
-Out of the \nPrograms~, we failed to parse and further analyze 2,367 projects due technical difficulties with the provided JSON files.
+Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due technical difficulties with the provided JSON files.
 Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}
@@
 Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per project, with the sprites and the stage of the project.
-We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks.
+We cross referenced also all blocks with the Scratch wiki to determine the shapes and the category of all blocks.
 For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
@@
 \label{dataAnalysis}
-All scraped project data, including the list of used blocks and parameters, was imported in a relational database.
+All scraped project data, including the list of used blocks and parameters, were imported in a relational database.
 We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions.
@@
 For RQ1 we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}.
-We used the McCabe cyclomatic complexity metric\cite{mcCabe76} for measuring the complexity of the scripts.
+For measuring the complexity of the scripts we used the McCabe cyclomatic complexity metric \cite{mcCabe76}, a quantitative measure of the number of independent paths through a program's source code.
+This is calculated per script by counting the number of decision points in the script plus one.
+In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks.
 
-For RQ2, we used the data on the code blocks and their categories to perform a detailed statistical analysis of applied programming abstractions and concepts.
+For RQ2, we used the data on the code blocks and their categories to perform statistical analysis of applied programming abstractions and concepts.
 Similarly to \cite{Maloney_2008}, we consider the use of certain blocks to indicate that a programming abstraction or concept is being used in a certain project.
@@
 The next step in the analysis was to determine the minimum size of the scripts that are considered clones instead of incidentally similar.
-For this, we examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}.
+We examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}.
 Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.
@@
 We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artifacts like spreadsheets~\cite{hermans2015detecting}.
-Using that strategy, we set the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
+Using that strategy, we set the thresholds for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
 The results we obtained using these thresholds are presented in Section \ref{RQ3}.
@@
 
-In Table \ref{tbl-size} we summarize statistics for the analyzed metrics.
+In Table \ref{tbl-size} we summarize the statistics for the analyzed metrics.
 We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts and blocks per project.
@@
 
-To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76}, a quantitative measure of the number of independent paths through a program's source code.
-This is calculated per script by counting the number of decision points in the script plus one.
-In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks.
-The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}.
+To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity.
+The results of this metric per script are plotted in Figure \ref{fig:cyclomatic}.
 The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point.
@@
 The distribution of variable and list utilization is shown in Figure \ref{fig:variables}.
-Exceptional cases exist: the analysis highlighted 842 projects with more than 100 variables and up to 340.
+Exceptional cases exist: the analysis highlighted 842 projects with more than 100 variables and with a maximum of 340.
 Examining the initialization of variables through the \texttt{set <variable> to <value>} blocks, we found that for 4.83\% of all variables this was missing.
@@
 
-The third and fourth cases are detected in scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality.
+The third and fourth cases are incomplete scripts.
+They are either never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or are comprised of only a \texttt{when <trigger>} block without any functionality.
 A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects.
@@
 Examining the size of these dead scripts, 72.34\% are composed of a single block.
-As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}] The number of projects exhibiting the dead code smell, considering all four types of dead code, is 65,760 (28.16\% of the total projects).
+As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}]
+
+The number of projects exhibiting the dead code smell, considering all four types of dead code combined, is 65,760 (28.16\% of the total projects).
 
@@
 Finally, we examine the large script and the large sprite smells.
-The thresholds we use for the identification of large scripts and large sprites are 18 blocks and 59 blocks respectively, as calculated in Section \ref{dataAnalysis}.
+The thresholds we use for the identification of large scripts and large sprites are 18 blocks and 59 blocks respectively, as explained in Section \ref{dataAnalysis}.
 The number of projects exhibiting the large script smell, i.e., containing at least one script with 18 or more blocks, is 69,521 (29.77\% of the total projects) and the number of projects with the large sprite smell is 31,954 (13.68\%).
@@
 However, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead.
-In contrast to other visual educational languages, Scratch does on indicate scripts that are dead.
-LEGO Mindstorms, for example, does give the user feedback by making unconnected blocks gray.
+In contrast to other visual educational languages, Scratch does not indicate scripts that are dead.
+LEGO Mindstorms, for example, does give the user such feedback by making unconnected blocks gray.
 
@@
 
-A study on the internalization of programming concepts with Scratch with 46 students was presented in \cite{meerbaum-salant_learning_2010}.
+For example, a study on the internalization of programming concepts with Scratch with 46 students was presented in \cite{meerbaum-salant_learning_2010}.
 Concepts like loops, conditional loops, message passing, initialization, variables and concurency were examined, and it was found that students had problems with the last three.
@@
 In this paper we presented a large-scale study on 247,798 projects we scraped from the Scratch repository.
-We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality.
-We found that procedures and conditional loops are not commonly used.
-We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells.
+We analyze these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality.
+We find that procedures and conditional loops are not commonly used.
+We further investigate the presence of code smells, including code duplication, dead code, long method and large class smells.
 Our findings indicate that Scratch programs suffer from code smells and especially from dead code and code duplication.

### 2016-04-22 Paper/paper.tex

@@
 Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}.
-Research works involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}.
+Previous work involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}.
 Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
@@
 Moreover, knowing that bad programming habits and code smells can be harmful \cite{hermans_2016}, we also want to explore whether they are common.
-Specifically, the research questions that we are trying to answer are:
+To address this goal, we answer the following research questions:
 
@@
 Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes.
-The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a piano.
+The Scratch program in Figure \ref{fig:ui} contains two one sprites, the cat, which is Scratch's default sprite and a piano.
 The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.

### 2016-04-22 Paper/paper.tex
Changes to address Bas's comments

@@
 
-\keywords{Scratch, programming practices, code smells, static analysis}
+\keywords{Scratch, block-based languages, programming practices, code smells, static analysis}
 
@@
 Several studies have shown that block-based languages are powerful as tools for teaching programming~\cite{meerbaum-salant_learning_2010, b._moskal_evaluating_2005,cooper_teaching_2003,price_comparing_2015}.
-Previous work involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}.
+Previous works involving static analysis of Scratch programs have evaluated the application of various programming concepts in Scratch projects \cite{Maloney_2008,moreno_automatic_2014}.
 Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
@@
 The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}.
-Because the term `procedures' is common in related work, we will refer to custom blocks as `procedures' in the following.
+Because the term `procedures' is common in related work, we will refer to custom blocks as `procedures' in the remainder of this paper.
 Procedures can have input parameters of type string, number, and boolean.
@@
 
-We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250,000 projects.
-Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due technical difficulties with the provided JSON files.
+We ran Kragle on March 2nd 2016 for 24 hours and, in that time, it obtained a little over 250,000 projects.
+Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due to technical difficulties with the provided JSON files.
 Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}
@@
 We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions.
-We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects, lists of which we also provide as part of the dataset.\footnotemark[\ref{repo}]
+We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects.
+Data for these cases are provided as part of the dataset.\footnotemark[\ref{repo}]
 
@@
 The number of blocks metric was further analyzed to understand code organization.
-As shown in Table \ref{tbl-size}, the majority of Scratch code---74.78\% out of \nLOC~blocks---is written within sprites.
+The majority of Scratch code---74.78\% out of \nLOC~blocks---is written within sprites.
 An additional 3.1\% of the total blocks are found in the stage class.
@@
 The complexity is higher, over 4, for 3.67\% of the scripts.
-The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246.\footnotemark[\ref{repo}] Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.
+The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246.\footnotemark[\ref{repo}] Cyclomatic complexity was greater (mean value of 3.32) in defined procedures, with 56.46\% of the procedures having at least one decision point.
 
@@
 Most code is written in sprites.
-A small number of projects (7.7\%) use procedures, but they use them a lot and for more complex code.
 There exist surprisingly large and complex projects.
@@
 	\begin{minipage}{0.46\textwidth}
-RQ2: One third of the projects use variables, sometimes without initializing them.
-39.81\% of the projects contain conditional statements and 77.18\% contain loops, but conditional loops are rarely used.
+RQ2: A small number of projects (8\%) use procedures, but they use them a lot and for more complex code.
+Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
+Recursive procedure calls exist in 1,052 (0.5\% of the total) projects.
+One third of the projects use variables, sometimes without initializing them.
+40\% of the projects contain conditional statements and 77\% contain loops, but conditional loops are rarely used.
 More than half of the projects are interactive.
-29.57\% of the projects use broadcast and receive blocks.
-Most procedures are called once or twice, from a single script which, in more than half of the cases, is another procedure.
-Recursive procedure calls exist in 1,052 (0.45\% of the total) projects.
+30\% of the projects use broadcast and receive blocks.
 	\end{minipage}
@@
 	\begin{minipage}{0.46\textwidth}
-RQ3: Code clones are found in 25.54\% of the projects, with 46.55\% of the clones copied three or more times, in the same or across sprites.
-28.16\% of the projects contain code that is never invoked, and thus exhibit the dead code smell.
+RQ3: Code clones are found in 26\% of the projects, with almost half of the clones copied three or more times, in the same or across sprites.
+28\% of the projects contain code that is never invoked, and thus exhibit the dead code smell.
 In some cases these scripts are large.
-The large script smell is found in 29.77\% of the projects and the large sprite one in 13.68\%.
+The large script smell is found in 30\% of the projects and the large sprite one in 14\%.
 	\end{minipage}}

### 2016-04-22 Paper/paper.tex
Changes based on Sohon's comments
Smaller figure margins
Added comment on temporary githib link

@@
 In this paper we explore the characteristics of Scratch programs.
-To this end we have scraped the Scratch public repository and retrieved 250.000 projects.
+To this end we have scraped the Scratch public repository and retrieved 250,000 projects.
 We present an analysis of these projects in three different dimensions.
@@
 		\includegraphics[width=\columnwidth]{fig/ui.png}
+		\vspace{-1em}
 		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `Events' in the middle and the code associated with the sprite on the right.}
@@
 Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due to technical difficulties with the provided JSON files.
-Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}
+Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code} (temporary link to protect the anonymity of the authors)}
 
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/11cloneslines}
+	\vspace{-1em}
 	\caption{Number of cloned scripts of different block sizes across and within sprites}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/13longmethod}
+	\vspace{-1em}
 	\caption{Size of sprites and scripts in number of blocks}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/1sprites}
+	\vspace{-1em}
 	\caption{Number of sprites in the analyzed projects}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/1scripts}
+	\vspace{-1em}
 	\caption{Number of scripts in the analyzed projects}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/1locs}
+	\vspace{-1em}
 	\caption{Number of blocks in the analyzed projects}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/3cyclomatic}
+	\vspace{-1em}
 	\caption{McCabe cyclomatic complexity of the \nscripts~analyzed scripts}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/16Categories}
+	\vspace{-1em}
 	\caption{Number of blocks from each category in the analyzed projects}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/16Shapes}
+	\vspace{-1em}
 	\caption{Number of blocks of each shape in the analyzed projects}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/6proceduresperproject}
+	\vspace{-1em}
 	\caption{Number of procedures for the 17,979 projects that include at least one}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/5arguments}
+	\vspace{-1em}
 	\caption{Number of arguments for the procedures in the dataset}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/7procedurecalls}
+	\vspace{-1em}
 	\caption{Number of calls of each procedure in the dataset}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/8variableslists}
+	\vspace{-1em}
 	\caption{Number of variables and lists used in the projects}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/11clonesprojects}
+	\vspace{-1em}
 	\caption{Cloned scripts in the dataset projects}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/11clonescopies}
+	\vspace{-1em}
 	\caption{Number of copies of the identified clones}
@@
 	\includegraphics[width=0.45\textwidth]{fig/charts/12deadcode}
+	\vspace{-1em}
 	\caption{Size of scripts identified as dead code}

### 2016-06-13 Paper/paper.tex
Fixed minor error
De-anonymized

@@
 
+\def\sharedaffiliation{%
+\end{tabular}
+\begin{tabular}{c}}
+
 \begin{document}
@@
 
-\numberofauthors{1}
+\numberofauthors{2}
 \author{
-% 1st. author
-\alignauthor
-(anonymized submission)
-%Ben Trovato\titlenote{Dr.~Trovato insisted his name be first.}\\
-%       \affaddr{Institute for Clarity in Documentation}\\
-%       \affaddr{1932 Wallamaloo Lane}\\
-%       \affaddr{Wallamaloo, New Zealand}\\
-%       \email{trovato@corporation.com}
-% 2nd. author
-%\alignauthor
-%G.K.M. Tobin\titlenote{The secretary disavows
-%any knowledge of this author's actions.}\\
-%       \affaddr{Institute for Clarity in Documentation}\\
-%       \affaddr{P.O. Box 1212}\\
-%       \affaddr{Dublin, Ohio 43017-6221}\\
-%       \email{webmaster@marysville-ohio.com}
+	\alignauthor Efthimia Aivaloglou\\      
+	\email{e.aivaloglou@tudelft.nl}
+	%
+	\alignauthor Felienne Hermans\\     
+	\email{f.f.j.hermans@tudelft.nl}
+	%
+	\sharedaffiliation
+	\affaddr{Software Engineering Research Group}\\
+	\affaddr{Delft University of Technology }   \\
+	\affaddr{Mekelweg 4, 2628 CD, Delft, the Netherlands }
 }
@@
 
-\terms{terms}
+%\terms{terms}
 
@@
 Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
-A recent controlled experiment found that long scripts and code duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
+In a recent controlled experiment we found that long scripts and code duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
 
@@
 By scraping the list of recent projects,\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} we have obtained \nPrograms~public Scratch projects and performed source code analysis on them.
-To the best of our knowledge, this is the first large-scale exploratory study of Scratch projects.
+To the best of our knowledge, this is the first large-scale exploratory study of Scratch programs.
 
@@
 Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes.
-The Scratch program in Figure \ref{fig:ui} contains two one sprites, the cat, which is Scratch's default sprite and a piano.
+The Scratch program in Figure \ref{fig:ui} contains two sprites, the cat, which is Scratch's default sprite and a piano.
 The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
@@
 Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due to technical difficulties with the provided JSON files.
-Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code} (temporary link to protect the anonymity of the authors)}
+Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code} \todo{Felienne, what is the new address?} }
 

### 2016-06-13 Paper/paper.tex
Additions to related work

@@
 
+Large-scale analyses of Scratch projects have been made using the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab, which contains data for Scratch projects created until 2012, when the web-based programming environment was introduced.
+In \cite{yang_2015}, this dataset was used for exploring the learning patterns of users in terms of block use over their first 50 projects.
+Dasgupta \emph{et al.} investigated how project remixing relates to the adoption of computational thinking concepts \cite{Dasgupta_2016}.
+In \cite{fields_2014}, the use of programming concepts was examined in relation to the level of participation (commenting, remixing, etc), the gender, and the account age of 5,000 Scratch programmers.
+
 Most related to our study for the second research question of programming abstractions and concepts is the work by Maloney \emph{et al.} \cite{Maloney_2008}, who analyzed 536 Scratch projects for blocks that relate to programming concepts including loops, conditional statements, variables, user interaction, synchronization, and random numbers.

### 2016-06-14 Paper/paper.tex
Removed deep understanding, clarified that only product data is used

@@
 
-The goal of this paper is to obtain a deep understanding of how people program in Scratch, to analyze the characteristics of Scratch programs, and to quantitatively evaluate the use of programming abstractions and concepts.
+The goal of this paper is to obtain an understanding of how people program in Scratch by analyzing their programming artifacts, i.e., the saved and shared Scratch projects, and to quantitatively evaluate the use of programming abstractions and concepts.
 Moreover, knowing that bad programming habits and code smells can be harmful \cite{hermans_2016}, we also want to explore whether they are common.

### 2016-06-14 Paper/paper.tex
Corrected mistake on percentage of decision points

@@
 	\begin{minipage}{0.46\textwidth}
-		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts, 76 blocks and no decision points.
+		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts and 76 blocks.
+The majority (78\%) of scripts contain no decision points.
 Most code is written in sprites.

### 2016-06-14 Paper/paper.tex
added section Insights for Computing Education

@@
 \subsubsection{Exact clones between sprites}
+\label{implicationsClones}
 With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common.
@@
 This might empower new Scratch users to get started faster.
+
+\subsection{Insights for computing education}
+Our findings confirm that Scratch is mainly used for its intended purpose as a first-exposure programming environment for creating simple programs and interactive animations: the majority of programs are small and more than half are interactive and contain no conditional statements.
+The analysis also indicates that computational concepts like conditionals and variables are being applied ---conditional statements are found in 40\% of the projects and variables in 32\%.
+The same does not hold for loops: even though 77\% of the projects contain loops, only 14\% contains conditional loops.
+We attribute the increased use of the \texttt{forever} loop to the Scratch language design and we are skeptical over whether it indicates an understanding of loop concepts.
+Only 8\% of the projects contain procedures, and those use them a lot and for more complex code, which is an indication of use by more experienced programmers.
+This very essential programming concept is therefore not sufficiently applied, which can be attributed both to limitations imposed by the Scratch programming environment, like the local scope of procedures, and to the difficulty for internalizing certain computational thinking concepts before a certain age \cite{seiter_2013}.
+
+The high occurrence of cloned scripts (in 26\% of the projects) could be the result of the limited use of procedures.
+Other code smells are also frequently found: 28\% of the projects include dead code and 30\% have large scripts.
+Knowing from prior research that long scripts and code duplication decreases ability to understand and modify Scratch programs \cite{hermans_2016}, and that Scratch programmers tend to exhibit certain bad programming habits \cite{Meerbaum_habits_2011}, we believe educating novice programmers on code quality concepts is an issue that should be further researched.
+
+It must be noted that the scope of this study includes the programming artifacts alone.
+Our findings are limited by the lack of (1) process data and (2) data on the age and other characteristics related to the programmers.
+However, the project data that is available in our dataset can facilitate further studies on computing education, and this is the reason that we are publishing it.
 

### 2016-06-14 Paper/paper.tex
Updated link to dataset

@@
 Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due to technical difficulties with the provided JSON files.
-Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code} \todo{Felienne, what is the new address?} }
+Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/TUDelftScratchLab/ScratchDataset}}
 

### 2016-06-15 Paper/paper.tex
Added reference

@@
 We then investigate complexity, used abstractions and programming concepts.
-Finally we detect \emph{code smells} like large scripts, dead code and duplicated code blocks.
+Finally we detect \emph{code smells} such as large scripts, dead code and duplicated code blocks.
 Our results show that 1) most Scratch programs are small, however Scratch programs consisting of over 100 sprites exist, 2) programming abstraction concepts like procedures are not commonly used and 3) Scratch programs do suffer from code smells including large scripts and unmatched broadcast signals.
@@
 
-Block-based languages have existed since the eighties, but have recently found adoption as tools for programming education.
+Block-based languages have existed since the eighties \cite{e._glinert_towards_1986}, but have recently found adoption as tools for programming education.
 In addition to Scratch, also Alice~\cite{conway_alice:_1994}, Blockly\footnote{\url{https://developers.google.com/blockly/}} and App Inventor~\cite{wolber_app_2011} are block-languages aimed at novice programmers.

### 2016-06-15 Paper/paper.tex
Intergrated Felienne's comments

@@
 Block-based languages have existed since the eighties \cite{e._glinert_towards_1986}, but have recently found adoption as tools for programming education.
-In addition to Scratch, also Alice~\cite{conway_alice:_1994}, Blockly\footnote{\url{https://developers.google.com/blockly/}} and App Inventor~\cite{wolber_app_2011} are block-languages aimed at novice programmers.
+In addition to Scratch, also Alice~\cite{conway_alice:_1994}, Blockly\footnote{\url{https://developers.google.com/blockly/}} and App Inventor~\cite{wolber_app_2011} are block-based and aimed at novice programmers.
 
@@
 Recent works have focused on bad programming practices within Scratch programs \cite{Meerbaum_habits_2011}, and automated quality assessment tools have been proposed for identifying code smells~\cite{fowler_refactoring:_1999}~and bad programming practices \cite{boe_hairball:_2013,moreno_automatic_2014}.
-In a recent controlled experiment we found that long scripts and code duplication decreases a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
+In a recent controlled experiment we found that long scripts and code duplication decrease a novice programmer's ability to understand and modify Scratch programs \cite{hermans_2016}.
 
-The goal of this paper is to obtain an understanding of how people program in Scratch by analyzing their programming artifacts, i.e., the saved and shared Scratch projects, and to quantitatively evaluate the use of programming abstractions and concepts.
+The goal of this paper is to obtain an understanding of how people program in Scratch by analyzing their programming artifacts, i.e., the saved and shared Scratch projects.
+With this we aim to quantitatively evaluate the use of programming abstractions and concepts.
 Moreover, knowing that bad programming habits and code smells can be harmful \cite{hermans_2016}, we also want to explore whether they are common.
@@
 Our study is based on data from the Scratch project repository.
-By scraping the list of recent projects,\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} we have obtained \nPrograms~public Scratch projects and performed source code analysis on them.
+By scraping the list of recent projects\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}}, we have obtained \nPrograms~public Scratch projects and performed source code analysis on them.
 To the best of our knowledge, this is the first large-scale exploratory study of Scratch programs.
@@
 \label{sec:scratch}
-This paper is by no means an introduction into Scratch programming, we refer the reader to \cite{brennan_creative_2014} for an extensive overview.
+This paper is by no means an introduction into Scratch programming.
+We refer the reader to \cite{brennan_creative_2014} for an extensive overview.
 To make this paper self-contained, however, we explain a number of relevant concepts in this section.
@@
 The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when <right arrow> key pressed}, each with associated blocks.
-	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
+	\item [Signals] Events within Scratch can be user-generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
 All other sprites can then react by using the \texttt{when I receive} Hat block.
 In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano.
-	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks.
+	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called \textit{custom blocks}.
 The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}.
 Because the term `procedures' is common in related work, we will refer to custom blocks as `procedures' in the remainder of this paper.
-Procedures can have input parameters of type string, number, and boolean.
-When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
+Procedure parameters can be textual, numeric or boolean.
+When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their procedure.
 \end{description}
@@
 
-The main focus of this study is to understand how people program in Scratch by analyzing the characteristics of Scratch projects.
+The main focus of this study is to understand how people program in Scratch, by analyzing the characteristics of Scratch projects.
 To answer our three research questions, we conducted an empirical quantitative evaluation of project data we collected from the Scratch project repository.
@@
 
-We obtained the set of Scratch projects by using a scraping program.
-Our scraping program called Kragle starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtains project ids of projects that were most recently shared.
+In order to collect the set of Scratch projects in the dataset we built a web scraping program.
+The scraping program, called Kragle, starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtains the project identifiers of projects that were most recently shared.
 Subsequently, Kragle retrieves the JSON code for each of the listed projects.
@@
 
-We ran Kragle on March 2nd 2016 for 24 hours and, in that time, it obtained a little over 250,000 projects.
-Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due to technical difficulties with the provided JSON files.
+We ran Kragle on March 2nd 2016 for 24 hours and, during that time, it obtained a little over 250,000 projects.
+Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due to technical difficulties with the JSON files.
 Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/TUDelftScratchLab/ScratchDataset}}
 
-Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per project, with the sprites and the stage of the project.
-We cross referenced also all blocks with the Scratch wiki to determine the shapes and the category of all blocks.
+Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}}.
+This resulted in a list of used blocks per project, with the sprites and the stage of the project.
+We also cross referenced all blocks with the Scratch wiki to determine the shapes and the category of all blocks.
 For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.
@@
 
-For RQ1 we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}.
+To answer RQ1, we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}.
 For measuring the complexity of the scripts we used the McCabe cyclomatic complexity metric \cite{mcCabe76}, a quantitative measure of the number of independent paths through a program's source code.
-This is calculated per script by counting the number of decision points in the script plus one.
-In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks.
+This is commonly calculated per script by counting the number of decision points in the script plus one.
+The decision points that Scratch supports are the \texttt{if} and the \texttt{if else} blocks.
 
 For RQ2, we used the data on the code blocks and their categories to perform statistical analysis of applied programming abstractions and concepts.
-Similarly to \cite{Maloney_2008}, we consider the use of certain blocks to indicate that a programming abstraction or concept is being used in a certain project.
+Similarly to \cite{Dasgupta_2016} and \cite{Maloney_2008}, we consider the use of certain blocks to indicate that a programming abstraction or concept is being used in a certain project.
 In Section \ref{RQ2} we present the results related to the utilization of procedures, variables, loops, conditional statements, user interactivity and synchronization.
@@
 For RQ3, we focused on four types of code smells: duplicated code, dead code, large script and large sprite.
-For the duplicated code smell analysis, our first step was to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same project.
+In the duplicated code smell analysis, our first step was to define what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same project.
 For the identification of clones we did not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal.
-We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}.
+We also investigated the case of clones with the same parameter values, and we refer to them as \textit{exact clones}.
 The next step in the analysis was to determine the minimum size of the scripts that are considered clones instead of incidentally similar.
 We examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}.
-Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.
+Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is a minimum size of five blocks per script.
 
@@
 
-To examine the long method and the large class smells, we consider them in the context of Scratch as large script and large sprite smells respectively.
+In the context of Scratch, we consider the long method and the large class smells as large script and large sprite smells respectively.
 For these two smells we use the number of blocks as the size metric.
-Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset.
-We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artifacts like spreadsheets~\cite{hermans2015detecting}.
-Using that strategy, we set the thresholds for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively.
+Figure \ref{fig:longmethod} presents the number of blocks in scripts and sprites of our dataset.
+We used these statistics to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artifacts like spreadsheets~\cite{hermans2015detecting}.
+Using that strategy, we set the thresholds for the calculation of the large script and large sprite: 18 blocks and 59 blocks respectively.
 The results we obtained using these thresholds are presented in Section \ref{RQ3}.
@@
 The dataset contains a relatively small number of projects without any code: \nemptyPrograms~ (5.77\%).
-Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while in other projects they were entirety empty apart from the Scratch cat added by default.
-Since these projects are empty in terms of code we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.
+Through random manual sampling we found that in some cases these projects contains only sprites and costumes, but no code, while other projects were entirely empty apart from the Scratch cat added by default.
+Since these projects are empty in terms of code, we excluded them from further analysis, leaving the final number of analyzed non-empty projects to \nScriptPrograms.
 
@@
 	\end{tabular}
-	\caption{Summary statistics from the dataset of \nScriptPrograms~non-empty Scratch projects}
+	\caption{Summary statistics of the dataset of \nScriptPrograms~non-empty Scratch projects}
 	\label{tbl-size}
@@
 
-The number of blocks metric was further analyzed to understand code organization.
+The number of blocks metric was further analyzed to understand code organization in more depth.
 The majority of Scratch code---74.78\% out of \nLOC~blocks---is written within sprites.
@@
 	\begin{minipage}{0.46\textwidth}
-		RQ1: The majority of Scratch projects are small and simple; 75\% of the projects have up to 5 sprites, 12 scripts and 76 blocks.
+		RQ1: The majority of Scratch projects are small and simple: 75\% of the projects have up to 5 sprites, 12 scripts and 76 blocks.
 The majority (78\%) of scripts contain no decision points.
@@
 In the dataset we found 206,799 procedures in 17,979 (7.7\%) projects.
-As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, but with 53.59\% of these projects having up to 2.
+As summarized in Table \ref{tbl-size}, the projects that contain procedures have an average of 11.5 procedures, with 53.59\% of these projects having up to 2.
 Figure \ref{fig:proceduresperproject} shows the distribution of procedures in projects.
@@
 	\vspace{-1em}
-	\caption{Number of procedures for the 17,979 projects that include at least one}
+	\caption{Number of procedures for the 17,979 projects that include at least one procedure}
 	\label{fig:proceduresperproject}
@@
 
-The use of procedures in projects was further investigated through the procedure calls, summarized in Figure \ref{fig:procedurecalls}.
+The use of procedures in projects was further investigated through the use of procedure calls, summarized in Figure \ref{fig:procedurecalls}.
 Most procedures are called exactly once (62.32\% of them) or twice (14.30\%) and from exactly one script (85.92\% of them).
-Examining the origin of procedure calls, we observed that most of the calls (56.09\%) originate from other procedures, and even 1.06\% originate from the same procedure, making them recursive calls.
+Examining the origin of procedure calls, we observed that most of the calls (56.09\%) originate from other procedures, and 1.06\% even originate from the same procedure, making them recursive calls.
 These recursive procedures are found in 1,052 projects, whose identifiers are made available.\footnotemark[\ref{repo}]
@@
 As shown in Table \ref{tbl-characteristics}, almost one-third of the projects use variables and a small number (4.01\%) use lists.
-The number of variables that is being used is also limited, with only 7.48\% of the projects having 5 or more variables.
+The number of variables that is being used is also limited, with only 7.48\% of the projects having five or more variables.
 The distribution of variable and list utilization is shown in Figure \ref{fig:variables}.
@@
 
-Regarding program control features, conditional statements (blocks \texttt{if <condition> then} and \texttt{if <condition> then else}) are used by 39.81\% of the projects.
-Loops (blocks \texttt{repeat <times>}, \texttt{forever} and \texttt{repeat until <condition>}) are more common, used by 77.18\% of the projects.
-The most common of the three is the \texttt{forever} block, accounting for 51.86\% of all loops and the least common one is the \texttt{repeat until <condition>} block, accounting for 11.57\% and used in 13.59\% of the total projects.
+Regarding program control features, conditionals (\texttt{if <condition> then} and \texttt{if <condition> then else} blocks) are used by 39.81\% of the projects.
+Loops (blocks \texttt{repeat <tim\-es>}, \texttt{forever} and \texttt{repeat until <condition>}) are more common, used by 77.18\% of the projects.
+The most common of the three is the \texttt{forever} block, accounting for 51.86\% of all loops and the least common one is the \texttt{repeat until <condition>} block, accounting for 11.57\% of all loops and used in 13.59\% of the total projects.
 
@@
 Table \ref{tbl-userinput} lists the frequency of use of user input controls.
-We do not include the \texttt{when Green flag pressed} block here, as this is just used to start a Scratch program and hence cannot really be considered input into the program.
+We do not include the \texttt{when Green flag pressed} block here, as this is just used to start a Scratch program and hence cannot be considered input into the program.
 The most commonly used user input block is the \texttt{when key pressed}, found in 71,096 (30.45\% of the total) projects.
@@
 
-Users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}.These blocks are used by 29.57\% of the projects.
+As detailed in Section \ref{sec:scratch}, users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}.These blocks are used in 29.57\% of the projects.
 \texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects.
@@
 We further inspected which of the identified clones were duplicated only within the same sprite: 63,682 (37.34\% of the total) clones, in 10.14\% of all projects.
-Procedure clones were measured to 12,878 (7.55\% of the total) clones, in 2.12\% of the projects.
+2.12\% of the projects contain cloned procedure definitions, which were measured to 12,878 (7.55\% of the total) clones.
 
@@
 
-Apart from whole scripts we also examined cases where scripts differed only in the first (Hat) block.
+Apart from whole scripts, we also examined cases where scripts differed only in the first (Hat) block.
 This way we examine if Scratch programmers assign the same functionality to handle different types of events.
@@
 The second smell that we examine is the dead code smell.
-We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts.
+We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts, i.e., scripts that contain an event block alone.
 Investigating the first type, we find that a significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects.
@@
 The third and fourth cases are incomplete scripts.
-They are either never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or are comprised of only a \texttt{when <trigger>} block without any functionality.
+Those are either never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or are comprised of only a \texttt{when <trigger>} block without any functionality.
 A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects.
-The majority of these scripts (86.6\%) are scripts missing the starting block.
+The majority of these scripts (86.6\%) are scripts missing a Hat block.
 Examining the size of these dead scripts, 72.34\% are composed of a single block.
@@
 \subsection{Implications}
-We believe a large scale study of programs like ours can help language designers to tailor their language.
+We believe a large scale study of programs like the current can help language designers to tailor their language.
 In this section we highlight directions in which our study could support language design.
@@
 In contrast to other visual educational languages, Scratch does not indicate scripts that are dead.
-LEGO Mindstorms, for example, does give the user such feedback by making unconnected blocks gray.
+LEGO Mindstorms, for example, does give the user feedback on dead blocks by making them gray.
 
 Looking at the number of unconnected blocks, we hypothesize that Scratch programmers have a need for a separate workspace to store unconnected blocks temporarily.
-We envision that would be like the `backpack' meant to move scripts across sprites.
+We envision that would be like the `backpack', a Scratch feature meant to move scripts across sprites.
 In order to help novice programmers keep their code clean, the programming interface could actively encourage users to move unconnected blocks to that workspace when they exit the environment.
@@
 With occurrences in 11\% of the Scratch projects in our dataset, the use of exactly identical clones between sprites is relatively common.
-In a sense, the Scratch users are not to blame here, as Scratch does not support procedure calls between sprites, only within them.
+In a sense, the Scratch users are not to blame for that, as Scratch does not support procedure calls between sprites, only within them.
 So in many cases there is no way to share the functionality other than by making a copy.
@@
 We attribute the increased use of the \texttt{forever} loop to the Scratch language design and we are skeptical over whether it indicates an understanding of loop concepts.
+
 Only 8\% of the projects contain procedures, and those use them a lot and for more complex code, which is an indication of use by more experienced programmers.
@@
 
-The high occurrence of cloned scripts (in 26\% of the projects) could be the result of the limited use of procedures.
+The high occurrence of cloned scripts (in 26\% of the proj\-ects) could be the result of the limited use of procedures.
 Other code smells are also frequently found: 28\% of the projects include dead code and 30\% have large scripts.
@@
 It could be the case that the programming habits of Scratch users are changing over time.
-However, we counterbalanced that by using a large dataset which comprises of around 2\% of all 14 million shared Scratch projects.\footnote{\url{https://scratch.mit.edu/statistics/}}.
+However, we counterbalanced that by using a large dataset which comprises around 2\% of all 14 million shared Scratch projects\footnote{\url{https://scratch.mit.edu/statistics/}}.
 
@@
 They connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops, a finding that is verified by our study.
-In \cite{wilson_evaluation_2012} 29 Scratch projects created from 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
+In \cite{wilson_evaluation_2012}, 29 Scratch projects created by 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
 
-Large-scale analyses of Scratch projects have been made using the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab, which contains data for Scratch projects created until 2012, when the web-based programming environment was introduced.
+Large-scale analyses of Scratch projects have been performed using the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab, which contains data for Scratch projects created until 2012, when the web-based programming environment was introduced.
 In \cite{yang_2015}, this dataset was used for exploring the learning patterns of users in terms of block use over their first 50 projects.
@@
 In their work \cite{moreno_automatic_2014}, Moreno and Robles extended Hairball to detect two bad programming habits in Scratch: not changing the default object names and duplicating scripts, and apply them for evaluating 100 projects from the Scratch repository.
-The results on script duplication are substantially different than ours---we find projects with script clones to appear half as frequently.
+The results on script duplication are substantially different from ours---we find projects with script clones to appear half as frequently.
 The Dr.

### 2016-06-15 Paper/paper.tex
Shortened

@@
 \label{sec:scratch}
-This paper is by no means an introduction into Scratch programming.
-We refer the reader to \cite{brennan_creative_2014} for an extensive overview.
-To make this paper self-contained, however, we explain a number of relevant concepts in this section.
-
 Scratch is a block-based programming language aimed at children, developed by MIT.
 Scratch can be used to create games and interactive animations, and is available both as a stand-alone and as a web application.
-Figure \ref{fig:ui} shows the Scratch user interface.
-The main concepts in the Scratch programming environment are:
+The main concepts in the Scratch programming environment are (we refer the reader to \cite{brennan_creative_2014} for an extensive overview):
 
@@
 Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes.
-The Scratch program in Figure \ref{fig:ui} contains two sprites, the cat, which is Scratch's default sprite and a piano.
-The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
 	\item [Scripts]Sprites can have multiple code blocks, called scripts.
-The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar.
 It is possible for a single sprite to have multiple scripts initiated by the same event.
@@
 In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam.
-The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when <right arrow> key pressed}, each with associated blocks.
-	\item [Signals] Events within Scratch can be user-generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}.
+	\item [Signals] Events within Scratch can be user-generated too: users can broadcast a message, for example when two sprites touch each other.
 All other sprites can then react by using the \texttt{when I receive} Hat block.
-In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano.
 	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called \textit{custom blocks}.
@@
 \end{description}
-
-\begin{figure}
-	\begin{center}
-		\includegraphics[width=\columnwidth]{fig/ui.png}
-		\vspace{-1em}
-		\caption{The Scratch user interface consisting of the `cat' sprite on the left, the toolbox with available blocks in the category `Events' in the middle and the code associated with the sprite on the right.}
-		\label{fig:ui}
-	\end{center}
-\end{figure}
 
@@
 	\vspace{-1em}
-	\caption{Number of arguments for the procedures in the dataset}
+	\caption{Number of arguments of the procedures in the dataset}
 	\label{fig:procedurearguments}
@@
 	\vspace{-1em}
-	\caption{Number of variables and lists used in the projects}
+	\caption{Number of used variables and lists}
 	\label{fig:variables}
@@
 The most common of the three is the \texttt{forever} block, accounting for 51.86\% of all loops and the least common one is the \texttt{repeat until <condition>} block, accounting for 11.57\% of all loops and used in 13.59\% of the total projects.
-
-Investigating user interactivity functionality, we found that 56.24\% of the projects in the dataset contain user input blocks---an average of 8.48 blocks per such project.
-Table \ref{tbl-userinput} lists the frequency of use of user input controls.
-We do not include the \texttt{when Green flag pressed} block here, as this is just used to start a Scratch program and hence cannot be considered input into the program.
-The most commonly used user input block is the \texttt{when key pressed}, found in 71,096 (30.45\% of the total) projects.
-The most frequently used parameter for the \texttt{key} attribute is the space key, followed by the arrows and then the letters and numbers.
 
@@
 
-As detailed in Section \ref{sec:scratch}, users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}.These blocks are used in 29.57\% of the projects.
-\texttt{broadcast and wait} is rarely used, in only 3.87\% of the projects.
+Investigating user interactivity functionality, we found that 56.24\% of the projects in the dataset contain user input blocks---an average of 8.48 blocks per such project.
+Table \ref{tbl-userinput} lists the frequency of use of user input controls.
+We do not include the \texttt{when Green flag pressed} block here, as this is just used to start a Scratch program and hence cannot be considered input into the program.
+The most commonly used user input block is the \texttt{when key pressed}, found in 71,096 (30.45\% of the total) projects.
+The most frequently used parameter for the \texttt{key} attribute is the space key, followed by the arrows and then the letters and numbers.
+
+As detailed in Section \ref{sec:scratch}, users can define their own events, using the blocks \texttt{broadcast}, \texttt{broadcast and wait} and \texttt{when I receive}.
+These blocks are used in 29.57\% of the projects.
+\texttt{broadcast and wait} is rarely used, in 3.87\% of the projects.
 
@@
 Figure \ref{fig:clonesprojects} plots the distribution of clones across projects.
-The majority of projects contain up to two cloned scripts; 7.24\% of the projects contain three or more.
+The majority of projects contain up to two cloned scripts.
 Figure \ref{fig:clonescopies} plots the number of copies of the identified clones.
@@
 
-We further inspected which of the identified clones were duplicated only within the same sprite: 63,682 (37.34\% of the total) clones, in 10.14\% of all projects.
+We further inspected which of the identified clones were duplicated within the same sprite: 63,682 (37.34\% of the total) clones, in 10.14\% of all projects.
 2.12\% of the projects contain cloned procedure definitions, which were measured to 12,878 (7.55\% of the total) clones.
-
 Exact clones were found in 11.81\% of the total projects.
 Their total number was 66,750 (39.14\% of the total) clones.
-Exact clones in the same sprite are cases of clearly redundant code.
-These were rare, found in only 0.87\% of the projects.
+Exact clones in the same sprite were rare, found in 0.87\% of the projects.
 
-Apart from whole scripts, we also examined cases where scripts differed only in the first (Hat) block.
+Apart from whole scripts, we also examined cases where scripts differ only in the first (Hat) block.
 This way we examine if Scratch programmers assign the same functionality to handle different types of events.
@@
 The second smell that we examine is the dead code smell.
-We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts, i.e., scripts that contain an event block alone.
+We identify four types of dead code: (1) procedures that are not invoked, (2) unmatched broadcast-receive messages, (3) code that is not invoked and (4) empty event scripts.
 Investigating the first type, we find that a significant number of the defined procedures (13,036 or 5.06\%) are not called in the projects.
@@
 Examining the size of these dead scripts, 72.34\% are composed of a single block.
-As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}]
+However, some dead scripts are considerably large; 2,358 dead scripts in 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}]
 
 The number of projects exhibiting the dead code smell, considering all four types of dead code combined, is 65,760 (28.16\% of the total projects).
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.45\textwidth]{fig/charts/12deadcode}
-	\vspace{-1em}
-	\caption{Size of scripts identified as dead code}
-	\label{fig:deadcode}
-\end{figure}
 
@@
 \subsubsection{Dead code}
-In our analysis, we find that more than one quarter of the Scratch projects contain dead `scripts': scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, scripts that are comprised of only a \texttt{when <trigger>} block without any functionality, procedures that are not invoked, or unmatched broadcast-receive messages.
-
+In our analysis, we find that more than one quarter of the Scratch projects contain dead scripts.
 In a sense, the dead scripts are harmless, as they are not executed.
@@
 This seems to indicate that there are common patterns in Scratch projects, which means it might be very beneficial to Scratch programmers if they could not only share their projects, but also share some of their functionality, for others to use, like a library.
-An example of such a library could be: functions for platforming games, including the movement of a player, collision detection and the implementation of `lives'.
+An example of such a library could be: functions for platforming games, including the movement of sprites, collision detection and the implementation of `lives'.
 This might empower new Scratch users to get started faster.
@@
 Only 8\% of the projects contain procedures, and those use them a lot and for more complex code, which is an indication of use by more experienced programmers.
-This very essential programming concept is therefore not sufficiently applied, which can be attributed both to limitations imposed by the Scratch programming environment, like the local scope of procedures, and to the difficulty for internalizing certain computational thinking concepts before a certain age \cite{seiter_2013}.
+This very essential programming concept is therefore not sufficiently applied, which can be attributed both to limitations imposed by the Scratch environment, like the local scope of procedures, and to the difficulty for internalizing certain computational thinking concepts before a certain age \cite{seiter_2013}.
 
@@
 Furthermore we use the number of blocks in the Scratch projects as a measure for the length of a program, while this does not exactly correspond to the `length' of a program in lines, and there can be multiple Scratch blocks on one line.
-For example, in Figure \ref{fig:ui} the `if' and `touching' blocks are present on the same `line'.
 We however believe that the number of blocks is a good proxy for size, and we plan a future experiment in which we will compare `lines of Scratch code' to `number of blocks'.
@@
 They connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops, a finding that is verified by our study.
-In \cite{wilson_evaluation_2012}, 29 Scratch projects created by 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
+In \cite{wilson_evaluation_2012}, 29 projects created by 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.
 

### 2016-06-29 Paper/paper.tex
Changed style to sig-alternate-05-2015 for camera ready

@@
-\documentclass{sig-alternate}
+\documentclass{sig-alternate-05-2015}
 
@@
 % --- End of Author Metadata ---
+
+% Copyright
+\CopyrightYear{2016} 
+\setcopyright{othergov}
+\conferenceinfo{ICER '16,}{September 08-12, 2016, Melbourne, VIC, Australia}
+\isbn{978-1-4503-4449-4/16/09}\acmPrice{\$15.00}
+\doi{http://dx.doi.org/10.1145/2960310.2960325}
+
+\clubpenalty=10000 
+\widowpenalty = 10000
 
@@
 
-% A category with the (minimum) three required fields
-%\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
-
-%\terms{terms}
-
-\keywords{Scratch, block-based languages, programming practices, code smells, static analysis}
+\keywords{Scratch; block-based languages; programming practices; code smells; static analysis}
 
@@
 Large-scale analyses of Scratch projects have been performed using the dataset made available by the Lifelong Kindergarten Group at the MIT Media Lab, which contains data for Scratch projects created until 2012, when the web-based programming environment was introduced.
-In \cite{yang_2015}, this dataset was used for exploring the learning patterns of users in terms of block use over their first 50 projects.
+In \cite{yang_2015}, this dataset was used for exploring the learning patterns of programmers in terms of block use over their first 50 projects.
 Dasgupta \emph{et al.} investigated how project remixing relates to the adoption of computational thinking concepts \cite{Dasgupta_2016}.
@@
 
-In addition to the findings presented in this paper, we provide as contributions the dataset that we used for our study, as well as information on the edge cases that we found in the dataset in terms of size and number of procedures, variables, cyclomatic complexity, clones and dead code.\footnotemark[\ref{repo}]
+In addition to the findings presented in this paper, we provide as contributions the dataset that we used for our study, as well as the project identifiers and information on the edge cases that we found in the dataset in terms of size and number of procedures, variables, cyclomatic complexity, clones and dead code.\footnotemark[\ref{repo}]
 

### 2016-06-29 Paper/paper.tex
Included the bibliography in the tex file

@@
 
-\bibliographystyle{abbrv}
-\bibliography{zotero}
+\begin{thebibliography}{10}
+
+	\bibitem{alves_deriving_2010}
+	T.~L. Alves, C.~Ypma, and J.~Visser.
+	\newblock Deriving metric thresholds from benchmark data.
+	\newblock In {\em 26th {IEEE} {International} {Conference} on {Software}
+		{Maintenance} ({ICSM} 2010)}, pages 1--10. IEEE Computer Society, 2010.
+
+	\bibitem{boe_hairball:_2013}
+	B.~Boe, C.~Hill, M.~Len, G.~Dreschler, P.~Conrad, and D.~Franklin.
+	\newblock Hairball: {Lint}-inspired {Static} {Analysis} of {Scratch}
+	{Projects}.
+	\newblock In {\em Proceeding of the 44th {ACM} {Technical} {Symposium} on
+		{Computer} {Science} {Education}}, {SIGCSE} '13, pages 215--220, New York,
+	NY, USA, 2013. ACM.
+
+	\bibitem{brennan_creative_2014}
+	K.~Brennan, C.~Balch, and M.~Chung.
+	\newblock {\em {Creative} {Computing}}.
+	\newblock Harvard Graduate School of Education, 2014.
+
+	\bibitem{conway_alice:_1994}
+	M.~Conway, R.~Pausch, R.~Gossweiler, and T.~Burnette.
+	\newblock Alice: {A} {Rapid} {Prototyping} {System} for {Building} {Virtual}
+	{Environments}.
+	\newblock In {\em Conference {Companion} on {Human} {Factors} in {Computing}
+		{Systems}}, {CHI} '94, pages 295--296, New York, NY, USA, 1994. ACM.
+
+	\bibitem{cooper_teaching_2003}
+	S.~Cooper, W.~Dann, and R.~Pausch.
+	\newblock Teaching {Objects}-first in {Introductory} {Computer} {Science}.
+	\newblock In {\em Proceedings of the 34th {SIGCSE} {Technical} {Symposium} on
+		{Computer} {Science} {Education}}, {SIGCSE} '03, pages 191--195, New York,
+	NY, USA, 2003. ACM.
+
+	\bibitem{Dasgupta_2016}
+	S.~Dasgupta, W.~Hale, A.~Monroy-Hern\'{a}ndez, and B.~M. Hill.
+	\newblock Remixing as a pathway to computational thinking.
+	\newblock In {\em Proceedings of the 19th ACM Conference on Computer-Supported
+		Cooperative Work \& Social Computing}, CSCW '16, pages 1438--1449, New York,
+	NY, USA, 2016. ACM.
+
+	\bibitem{fields_2014}
+	D.~A. Fields, M.~Giang, and Y.~Kafai.
+	\newblock Programming in the wild: Trends in youth computational participation
+	in the online scratch community.
+	\newblock In {\em Proceedings of the 9th Workshop in Primary and Secondary
+		Computing Education}, WiPSCE '14, pages 2--11, New York, NY, USA, 2014. ACM.
+
+	\bibitem{fowler_refactoring:_1999}
+	M.~Fowler.
+	\newblock {\em Refactoring: improving the design of existing code}.
+	\newblock Addison-Wesley Longman Publishing Co., Inc., Boston, MA, USA, 1999.
+
+	\bibitem{e._glinert_towards_1986}
+	E.~Glinert.
+	\newblock {Towards} "{Second} {Generation}" {Interactive}, {Graphical}
+	{Programming} {Environments}.
+	\newblock In {\em Proceedings of the {IEEE} {Workshop} on {Visual}
+		{Languages}}, pages 61---70, 1986.
+
+	\bibitem{hermans_2016}
+	F.~Hermans and E.~Aivaloglou.
+	\newblock Do code smells hamper novice programming?
+	\newblock In {\em Proceedings of the International Conference on Program
+		Comprehension}, 2016.
+	\newblock to appear.
+
+	\bibitem{hermans2015detecting}
+	F.~Hermans, M.~Pinzger, and A.~van Deursen.
+	\newblock Detecting and refactoring code smells in spreadsheet formulas.
+	\newblock {\em Empirical Software Engineering}, 20(2):549--575, 2015.
+
+	\bibitem{Maloney_2008}
+	J.~H. Maloney, K.~Peppler, Y.~Kafai, M.~Resnick, and N.~Rusk.
+	\newblock Programming by choice: Urban youth learning programming with scratch.
+	\newblock In {\em Proceedings of the 39th SIGCSE Technical Symposium on
+		Computer Science Education}, SIGCSE '08, pages 367--371, New York, NY, USA,
+	2008. ACM.
+
+	\bibitem{mcCabe76}
+	T.~J. McCabe.
+	\newblock A complexity measure.
+	\newblock {\em {IEEE} Trans. Software Eng.}, 2(4):308--320, 1976.
+
+	\bibitem{Meerbaum_habits_2011}
+	O.~Meerbaum-Salant, M.~Armoni, and M.~Ben-Ari.
+	\newblock Habits of programming in scratch.
+	\newblock In {\em Proceedings of the 16th Annual Joint Conference on Innovation
+		and Technology in Computer Science Education}, ITiCSE '11, pages 168--172,
+	New York, NY, USA, 2011. ACM.
+
+	\bibitem{meerbaum-salant_learning_2010}
+	O.~Meerbaum-Salant, M.~Armoni, and M.~M. Ben-Ari.
+	\newblock Learning {Computer} {Science} {Concepts} with {Scratch}.
+	\newblock In {\em Proceedings of the {Sixth} {International} {Workshop} on
+		{Computing} {Education} {Research}}, {ICER} '10, pages 69--76, New York, NY,
+	USA, 2010. ACM.
+
+	\bibitem{moreno_automatic_2014}
+	J.~Moreno and G.~Robles.
+	\newblock Automatic detection of bad programming habits in scratch: {A}
+	preliminary study.
+	\newblock In {\em 2014 {IEEE} {Frontiers} in {Education} {Conference} ({FIE})},
+	pages 1--4, Oct. 2014.
+
+	\bibitem{moreno-leon_dr._2015}
+	J.~Moreno-Le{\'o}n, G.~Robles, and M.~Rom{\'a}n-Gonz{\'a}lez.
+	\newblock Dr. {Scratch}: {Automatic} {Analysis} of {Scratch} {Projects} to
+	{Assess} and {Foster} {Computational} {Thinking}.
+	\newblock {\em RED : Revista de Educaci{\'o}n a Distancia}, (46):1--23, Jan.
+	2015.
+
+	\bibitem{b._moskal_evaluating_2005}
+	B.~Moskal, S.~Cooper, and D.~Lurie.
+	\newblock Evaluating the {Effectiveness} of a {New} {Instructional} {Approach}.
+	\newblock In {\em Proceedings of the {SIGCSE} technical symposium on {Computer}
+		science education}, 2005.
+
+	\bibitem{price_comparing_2015}
+	T.~W. Price and T.~Barnes.
+	\newblock Comparing {Textual} and {Block} {Interfaces} in a {Novice}
+	{Programming} {Environment}.
+	\newblock In {\em Proceedings of the {Eleventh} {Annual} {International}
+		{Conference} on {International} {Computing} {Education} {Research}}, {ICER}
+	'15, pages 91--99, New York, NY, USA, 2015. ACM.
+
+	\bibitem{resnick_scratch:_2009}
+	M.~Resnick, J.~Maloney, A.~Monroy-Hernández, N.~Rusk, E.~Eastmond, K.~Brennan,
+	A.~Millner, E.~Rosenbaum, J.~Silver, B.~Silverman, and Y.~Kafai.
+	\newblock Scratch: {Programming} for {All}.
+	\newblock {\em Commun. ACM}, 52(11):60--67, Nov. 2009.
+
+	\bibitem{seiter_2013}
+	L.~Seiter and B.~Foreman.
+	\newblock Modeling the learning progressions of computational thinking of
+	primary grade students.
+	\newblock In {\em Proceedings of the Ninth Annual International ACM Conference
+		on International Computing Education Research}, ICER '13, pages 59--66, New
+	York, NY, USA, 2013. ACM.
+
+	\bibitem{wilson_evaluation_2012}
+	A.~Wilson, T.~Hainey, and T.~Connolly.
+	\newblock Evaluation of computer games developed by primary school children to
+	gauge understanding of programming concepts.
+	\newblock In {\em European Conference on Games Based Learning}, page 549.
+	Academic Conferences International Limited, 2012.
+
+	\bibitem{wolber_app_2011}
+	D.~Wolber, H.~Abelson, E.~Spertus, and L.~Looney.
+	\newblock {\em App {Inventor}: {Create} {Your} {Own} {Android} {Apps}}.
+	\newblock O'Reilly Media, Sebastopol, Calif, 1 edition edition, May 2011.
+
+	\bibitem{yang_2015}
+	S.~Yang, C.~Domeniconi, M.~Revelle, M.~Sweeney, B.~U. Gelman, C.~Beckley, and
+	A.~Johri.
+	\newblock Uncovering trajectories of informal learning in large online
+	communities of creators.
+	\newblock In {\em Proceedings of the Second (2015) ACM Conference on Learning @
+		Scale}, L@S '15, pages 131--140, New York, NY, USA, 2015. ACM.
+
+\end{thebibliography}
+
 \end{document}
